# Comparing quantitative data between individuals {#SummariseQuantData}


## Introduction

Relational and interventional RQs often require comparing groups.
This chapter considers how to compare *qualitative* variables in different groups.
Graphs are very useful this purpose, and a table of the numerical summaries usually are also produced.


## Graphs

When a *quantitative* variable is compared across different groups (i.e., comparing between individuals), the individual variables can be plotted.
However, to *compare* the quantitative variable in different groups, options for plotting include: 

* [*Back-to-back stemplots*](#BackToBackStem):
  Best for small amounts of data when the qualitative variable has *two levels*; 
* [*2-D dot charts*](#TwoDDot): 
  Best choice for small to moderate amounts of data;
* [*Boxplots*](#Boxplot):
  Best choice, except for small amounts of data.        

These situations have *one qualitative variable* and *one quantitative variable*.


### Back-to-back stemplot {#BackToBackStem}

Back-to-back stemplots are two stemplots (Sect.\ \@ref(StemAndLeafPlots)) sharing the same stems; one group has the leaves emerging left-to-right from the stem, and the second group has the leaves emerging right-to-left from the stem.
Back-to-back stemplots can only be used when *two* groups are being compared.
Again, one advantage of using stemplots over other plots is that the original data are retained.



::: {.example #KrillData name="Back-to-back stemplots"}
A study recorded the number chest-beats by gorillas [@wright2021chest], for gorillas under 20 years old (young) and 20 years and over (older)
`r if (knitr::is_latex_output()) {
   '(Table\\ \\@ref(tab:KrillDataTable)).'
} else {
   '(Table\\ \\@ref(tab:KrillDataTableHTML)).'
}`
A back-to-back stemplot allows the two groups to be compared visually 
`r if (knitr::is_latex_output()) {
   '(Fig.\\ \\@ref(fig:KrillEggsDoubleStem)).'
} else {
   '(Fig.\\ \\@ref(fig:KrillEggsDoubleStemHTML)).'
}`
The leaves for younger gorillas go from right-to-left, and the leaves for older gorillas go left-to-right, sharing the same stems.
The younger groups has a faster chest-beating rate.
One older gorilla has a much faster rate that the other older gorillas. 
:::


```{r}
data(Gorillas)
CBRyounger <- sort( Gorillas$ChestBeatRate[Gorillas$Age20 == "Younger"] )
CBRolder <- sort( Gorillas$ChestBeatRate[Gorillas$Age20 == "Older"] )

Gorilla2 <- cbind( CBRyounger[1:7],
                   CBRyounger[8:14],
                   CBRolder[1:7],
                   c(CBRolder[8:11], NA, NA, NA ) )
```

`r if (knitr::is_html_output()) '<!--'`
\begin{figure}
\begin{minipage}{0.4\textwidth}
\captionof{table}{The chest-beating rate for gorillas\label{tab:KrillDataTable}}
\fontsize{10}{12}\selectfont
```{r KrillDataTable}

if( knitr::is_latex_output() ) {
  kable( surroundMaths(Gorilla2,
                       decDigits = 1),
         format = "latex",
         longtable = FALSE,
         table.env = "@empty",
         booktabs = TRUE,
         escape = FALSE,
         linesep = c("", "", "", "\\addlinespace"),
         align = c("c","c")) %>%
         #col.names = c("Treatment group", "Control group"),
    add_header_above(header = c("Younger " = 2, 
                                "Older" = 2), 
                     bold = TRUE, 
                     align = "c")
}
```
\end{minipage}
\hspace{0.05\textwidth}
\begin{minipage}{0.60\textwidth}%
\centering
```{r KrillEggsDoubleStem, fig.width=6.5, fig.height=3, out.width='100%'}

  stemgrid.x <- 1
  stemgrid.y <- 6
  stemcounts <- array(1, 
                      dim = stemgrid.y)

  par( mar = c(0.25, 0.25, 4, 0.25) + 0.1)

  plot( expand.grid(1:stemgrid.x, 0:stemgrid.y), 
      type = "n",
      xlab = "",
      ylab = "",
      main = "Chest-beating rate of gorillas\n(1|6 means 1.6 beats per 10 hours)",
      xlim = c(-11, 8),
      ylim = c(-0.2, 5.2),
      axes = FALSE)
  
### Treatment data in cols 1  to 11 (x = -11 to x = -1)
### Stems in col 12  (x=0)
### Control data in cols 13 to 19 (x= 1 to x = 7)
  
  # Divide the stems and leaves
  lines(c(-0.5, -0.5),
        c(0.5, 6.5),
         lwd = 2,
         col = "grey")
  lines(c(0.5, 0.5),
        c(0.5, 6.5),
         lwd = 2,
         col = "grey")
  text(-5, 0, 
       "Younger", 
       font = 2)
  text( 4, 0, 
        "Older", 
	font = 2)
  
  # Plot stems
  points( c(0, 0, 0, 0, 0), 
          1:5,
          pch = as.character(4:0))
  
  
  ### Plot leaves
  for (j in (1:2) ){

    if (j == 1) { # Right-hand side
      CBR <- CBRolder
      step <- 1
      stemcounts <- array(1, dim = stemgrid.y)
    }
    if (j == 2) { # Left-hand side
      CBR <- CBRyounger
      step <- -1
      stemcounts <- array(-1, dim = stemgrid.y)
    }
    
    for (i in 1:length(CBR)){
    
      if ( CBR[i] < 1) {
        points( stemcounts[1], 5, 
	        pch = substr( sprintf("%1.1f", CBR[i]), 3, 3)) 
        stemcounts[1] <- stemcounts[1] + step
      }

    if ( (CBR[i] >= 1) & (CBR[i] < 2) ) {
      points( stemcounts[2], 4, 
              pch = substr( sprintf("%1.1f", CBR[i] - 1), 3, 3)) 
      stemcounts[2] <- stemcounts[2] + step
    }

    if ( (CBR[i] >= 2) & (CBR[i] < 3) ) {
      points( stemcounts[3], 3, 
              pch = substr( sprintf("%1.1f", CBR[i] - 2), 3, 3)) 
      stemcounts[3] <- stemcounts[3] + step
    }
    if ( ( CBR[i] >= 3 ) & (CBR[i] < 4) ) {
      points( stemcounts[4], 2, 
              pch = substr( sprintf("%1.1f", CBR[i] - 3), 3, 3)) 
      stemcounts[4] <- stemcounts[4] + step
    }
    if ( ( CBR[i] >= 4 ) & (CBR[i] < 5) ) {
      points( stemcounts[5], 1, 
              pch = substr( sprintf("%1.1f", CBR[i] - 4), 3, 3)) 
      stemcounts[5] <- stemcounts[5] + step
    }
  }
}
```
\caption{The number of eggs from krill, for treatment and control groups}\label{fig:KrillEggsDoubleStem}
\end{minipage}
\end{figure}
`r if (knitr::is_html_output()) '-->'`






`r if (knitr::is_latex_output()) '<!--'`
```{r KrillDataTableHTML}
if (knitr::is_html_output()) {
  kable( Gorilla2,
         format = "html",
         longtable = FALSE,
         booktabs = TRUE,
         escape = FALSE,
         caption = "The chest-beating rate for gorillas, for younger and older gorillas",
         linesep = c("", "", "\\addlinespace"),
         align = c("c","c")) %>%
    add_header_above(header = c("Younger " = 2, 
                                "Older" = 2), 
                     bold = TRUE, 
                     align = "c")
}
```




```{r KrillEggsDoubleStemHTML, fig.width=6.0, out.width='100%', fig.cap="The number of eggs from krill, for treatment and control groups", fig.cap="center"}

data(Gorillas)
CBRyounger <- sort( Gorillas$ChestBeatRate[Gorillas$Age20 == "Younger"] )
CBRolder <- sort( Gorillas$ChestBeatRate[Gorillas$Age20 == "Older"] )

Gorilla2 <- cbind( CBRyounger[1:7],
                   CBRyounger[8:14],
                   CBRolder[1:7],
                   c(CBRolder[8:11], NA, NA, NA ) )

stemgrid.x <- 1
stemgrid.y <- 6
stemcounts <- array(1, dim=stemgrid.y)

par( mar = c(0.5, 0.5, 4, 0.5) + 0.1)

plot( expand.grid(1:stemgrid.x, 0:stemgrid.y), 
      type = "n",
      xlab = "",
      ylab = "",
      main = "Stemplot of egg counts of krill\n(1|6 means 16 eggs)",
      xlim = c(-11, 7),
      ylim = c(-0.2, 6.2),
      axes = FALSE)

### Treatment data in cols 1  to 11 (x = -11 to x = -1)
### Stems in col 12  (x=0)
### Control data in cols 13 to 19 (x= 1 to x = 7)

# Divide the stems and leaves
lines(c(-0.5, -0.5),
      c(0.5, 5.5),
      lwd = 2,
      col = "grey")
lines(c(0.5, 0.5),
      c(0.5, 6.5),
      lwd = 2,
      col = "grey")
text(-5, 0, 
     "Younger", 
     font = 2)
text( 4, 0, 
      "Older", 
      font = 2)

# Plot stems
points( c(0, 0, 0, 0, 0), 
        1:5,
        pch = as.character(4:0))
# Plot leaves
for (j in (1:2) ){
  
  if (j == 1) { # Right-hand side
    CBR <- CBRolder
    step <- 1
    stemcounts <- array(1, dim=stemgrid.y)
  }
  if (j == 2) { # Left-hand side
    CBR <- CBRyounger
      step <- -1
      stemcounts <- array(-1, dim = stemgrid.y)
    }
    
    for (i in 1:length(CBR)){
    
      if ( CBR[i] < 10) {
        points( stemcounts[1], 6, 
	        pch = as.character(CBR[i]))
        stemcounts[1] <- stemcounts[1] + step
      }

    if ( (CBR[i] >= 10) & (CBR[i] < 20) ) {
      points( stemcounts[2], 5, 
              pch = as.character(CBR[i] - 10) )
      stemcounts[2] <- stemcounts[2] + step
    }

    if ( (CBR[i] >= 20) & (CBR[i] < 30) ) {
      points( stemcounts[3], 4, 
              pch = as.character(CBR[i] - 20))
      stemcounts[3] <- stemcounts[3] + step
    }
    if ( ( CBR[i] >= 30 ) & (CBR[i] < 40) ) {
      points( stemcounts[4], 3, 
              pch = as.character(CBR[i] - 30))
      stemcounts[4] <- stemcounts[4] + step
    }
    if ( ( CBR[i] >= 40 ) & (CBR[i] < 50) ) {
      points( stemcounts[5], 2, 
              pch = as.character(CBR[i] - 40))
      stemcounts[5] <- stemcounts[5] + step
    }
    if ( ( CBR[i] >= 50 ) & (CBR[i] < 60) ) {
      points( stemcounts[6], 1, 
              pch = as.character(CBR[i] - 50))
      stemcounts[6] <- stemcounts[6] + step
    }
  }
}
```
`r if (knitr::is_latex_output()) '-->'`



### 2-D dot charts {#TwoDDot}

A 2-dimensional (2-D) dot chart places a dot for each observation, separated for each level of the qualitative variable (also see Sect.\ \@ref(DotChartsOneQual)).
For the same krill data in Example\ \@ref(exm:KrillData), a dot chart is shown in Fig.\ \@ref(fig:TwoDDotchart).

Many observations are the same, so some points would be *over-plotted* if points were not *stacked* (left panel), or some randomness added (a 'jitter') in the vertical direction, before plotting (right panel).


```{r TwoDDotchart, fig.height=6, out.width='100%',fig.cap="Two variations of a 2-D dot chart for the chest-beating data to avoid overplotting: stacking (left) and jittering (right)", fig.align="center", fig.width=8.75, fig.height=2.5}
old.par <- par()
par(mar = c(5, 6, 4, 2) + 0.1)
par(mfrow = c(1,2))

stripchart( list(Younger = CBRyounger, 
                 Older = CBRolder), 
            main = "A 2-D dot chart; the chest-beating\nrate of gorillas (points stacked)",
            method = "stack", 
            las = 1,
            pch = 19,
            xlim = c(0, 5),
            cex = 0.7,
            ylim = c(0.6, 2.4),
            xlab = "Chest-beating rate")

stripchart( list(Treatment = CBRyounger,
                 Control = CBRolder), 
            main = "A 2-D dot chart; the chest-beating\nrate of gorillas (points jittered)",
            method = "jitter",
            jitter = 0.18, 
            las = 1,
            pch = 19,
            xlim = c(0, 5),
            cex = 0.7,
            ylim = c(0.6, 2.4),
            xlab = "Chest-beating rate")

#par( old.par )
```


### Boxplots {#Boxplot}

A boxplot is a picture of the quantiles (Sect.\ \@ref(VariationIQR)).
The distribution for each group is summarised by five numbers: the minimum value, $Q_1$, the median, $Q_3$ and the maximum value.
Outliers, identified using the IQR rule (Sect. \@ref(OutliersIQRrule)), are also shown.
Distributions can be compared by comparing the values of $Q_1$, the medians, and the values of $Q_3$.
Since each distribution is summarised by just five numbers, a lot of detail of the distribution is lost.
For this reason, boxplots are best for *comparing* distributions. 
Software uses approximations to compute these values (and different software may use different rules, and hence may produce slightly different values for quartiles).

An example of a boxplot, for the chest-beating data in Example\ \@ref(exm:KrillData), is shown in Fig.\ \@ref(fig:BoxplotGorillas).
No outliers are identified for younger gorillas; one large outlier is identified for the older gorillas.
The boxplot shows a distinct difference between the chest-beating rates of older and younger gorillas.


```{r BoxplotGorillas, fig.cap="A boxplot for the chest-beating data", fig.align="center",  fig.height=3.5, , fig.width=4.5, out.width='50%'}
CBRyounger.quantiles <- quantile(CBRyounger)
CBR.C2 <- CBRolder[ -which.max(CBRolder)]
KrillC2.quantiles <- quantile(CBR.C2)

par( mar = c(4.5, 3, 4, 2) + 0.1 )
Gorillas$Age20 <- ordered(Gorillas$Age20,
                          levels = c("Younger", 
                                     "Older"))
boxplot( ChestBeatRate ~ Age20,
         data = Gorillas,
         col = plot.colour,
         las = 1,
         pch = 19,
         ylim = c(0, 4.5),
         xlab = "Age group",
         ylab = "Chest-beating rate\n (per 110 h)",
         main = "Chest-beating rate\nor gorillas")

```


The boxplots are explained in Fig.\ \@ref(fig:BoxplotGorillasExplain).
First focus on just the boxplot for the *younger* gorillas (i.e., the left box).
Boxplots have five horizontal lines; from the top to the bottom of the plot:

1. *Top line*:
   The *fastest* chest-beating rate is $`r CBRyounger.quantiles[5]`$ per $10$\ h.
2. *Second line from top*:
   $75$% of observations are smaller than about $`r CBRyounger.quantiles[4]`$, represented by the line at the top of the central box. 
   This is the *third quartile* ($Q_3$).
3. *Middle line*:
   $50$% of observations are smaller than about $`r CBRyounger.quantiles[3]`$, represented by the line inside the central box. 
   This is an 'average' value, the *second quartile* ($Q_2$).
4. *Second line from bottom*:
   $25$% of observations are smaller than about $`r CBRyounger.quantiles[2]`$, represented by the line at the bottom of the central box. 
   This is the *first quartile* ($Q_1$).
5. *Bottom line*:
   The *slowest* chest-beating rate is $`r CBRyounger.quantiles[1]`$ per $10$\ h. 




```{r BoxplotGorillasExplain, fig.cap="Explaining the boxplots for the chest-beating data", fig.align="center",  fig.height=4.25, , fig.width=10.5, out.width='100%'}

source("R/showExplainBoxplots.R")
```


The box for the *older* gorillas is slightly different
`r if (knitr::is_latex_output()) {
   '(Fig.\\ \\@ref(fig:BoxplotKrill), left panel):'
} else {
   '(Fig.\\ \\@ref(BoxplotKrillHTML)):'
}`
one observation is identified with a point, *above* the top line.
Computer software has identified this observation as a potentially *outlier* (in this case, unusually *large*), and has plotted this point separately, using the IQR rule (Sect.\ \@ref(OutliersIQRrule)).
Mild and extreme outliers are plotted in this way (Sect.\ \@ref(OutliersIQRrule)) then then the values of the quantiles ($Q_1$, $Q_2$ and $Q_3$) are computed as usual.

```{r}
outC <- boxplot( list(Older = CBRolder),
                 plot = FALSE)
CBR.quantiles <- outC$stats
```

The values of $Q_1$, the median and $Q_3$ are all substantially larger for the younger gorillas, suggesting that younger gorillas have, in general, faster chest-beating rates.




::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
Boxplots summarise data with only five numbers, so a lot of detail of the distribution is lost.
For this reason, boxplots are excellent for *comparing* distributions, but histograms are better for displaying a single quantitative distribution.
:::


::: {.example #BoxplotsHorizontal name="Boxplots"}
Boxplots can be plotted horizontally too, which leaves space for long labels.
In Fig.\ \@ref(fig:CementBuildBoxplot) (left panel; based on @data:Silva2016:rootcanal), the three dental cements are very different regarding their push-out forces.
:::


```{r CementBuildBoxplot, fig.cap="Comparing three push-out values for three dental cements", fig.align="center", fig.height=2.75, fig.width=8.5, fig.show='hold', out.width='100%'}
MTA    <- c(0.05, 0.1, 0.2, 0.25, 0.45)
Endo   <- c(0.1, 0.2, 0.4,   1.0, 2.05)
AHplus <- c(0.5, 1.1, 2.25, 2.9, 4.4)

CementStats <- list( n = c(5, 5, 5),
                     stats = matrix( cbind(MTA, 
                                           Endo, 
                                           AHplus), 
                                     ncol = 3, 
                                     byrow = FALSE),
                     conf = matrix( c(-1.5, 7, -2, 10, 3, 3), 
		                 ncol = 2, 
				             byrow = FALSE),  ### NOT USEFUL, only for notches
                  out = NULL,
                  group = NULL,
                  names = c("MTA Fillapex", 
		                        "Endo Seal", 
			                      "AH Plus"))

par( mfrow = c(1, 2),
     mar = c(4, 6, 3, 2) + 0.1)
plot( c(0, 5), 
      c(0.4, 3.6),
      type = "n",
      axes = FALSE,
      main = "Boxplot of push-out force",
      ylab = "",
      xlab = "Push-out force (MPa)")

axis(side = 1,
     las = 1)
abline(h = 0, 
       col = "grey", 
       lwd = 2)
bxp(CementStats,
    las = 1,
    horizontal = TRUE,
    add = TRUE,
    boxfill = plot.colour)

######

DB  <-  c( 7,   3, 1.5, 1, -1.5)
DBB <- c(10, 5.5, 3.5, 2, -2)

EngStats <- list( n = c(5, 5),
                  stats = matrix( cbind(DB, DBB), 
                                  ncol = 2, 
                                  byrow = FALSE),
                  conf = matrix( c(-1.5, 7, -2, 10), 
                                 ncol = 2, 
                                 byrow = FALSE),
                  out = c(8, 20, 12.5, 11.5, -7),
                  group = c(1, 2, 2, 2, 2),
                  names = c("DB", 
                            "DBB"))
plot( c(0.4, 2.6), 
      c(-10, 20),
      type = "n",
      axes = FALSE,
      main = "Boxplot of project costs",
      ylab = "Cost increase (%)",
      xlab = "Project delivery method")

axis(side = 2,
     las = 1)
abline(h = 0, 
       col = "grey", 
       lwd = 2)
bxp(EngStats,
    las = 1,
    add = TRUE,
    boxfill = plot.colour)
```


::: {.example #Boxplots2 name="Boxplots"}
A study of different engineering project delivery methods [@data:Hale2009:ProjectDelivery] produced the boxplot in Fig.\ \@ref(fig:CementBuildBoxplot) (right panel): the increase in the costs of projects seem to differ between the two methods.
The quantitative variable is the percentage increase in costs, and the qualitative variable is the method used.

The Design/Build (DB) method produces a smaller project cost growth on average (the centre line of the boxplot), but the Design/Bid/Build (DBB) method produces more variation in project cost growth.
Outliers are present for both methods, as indicated by the dots.
:::



```{r}
htmltools::tags$video(src = "./videos/Boxplots.mp4", 
                      width = "550", 
                      controls = "controls", 
                      loop = "loop", 
                      style = "padding:5px; border: 2px solid gray;")
```


## Summary tables {#CompareQuantTables}

The numerical summary information can be collated in a table, or a *tabular summary*.
The data should be summarised the quantitative variable for each group, and (when appropriate) the changes should also be summarised.

::: {.example #GorillaSummarytable name="Numerical summary table"}
For the gorill data, the summary of the data can be tabulated as in Table\ \@ref(tab:GorillaTable).
Notice that no standard deviation or sample size is provided; these make no sense.

Based on the table, do you think a difference exists between the mean chest-beating rate for younger and older gorillas?
Remember, the data in Table\ \@ref(tab:GorillaTable) is for a *sample*, but the RQ is asking about the *population*.
:::

```{r GorillaTable}
GorillaSummaryTable <- array( dim = c(3, 3))

rownames(GorillaSummaryTable) <- c("Younger",
                        "Older",
                        "Difference") 
colnames(GorillaSummaryTable) <- c("Mean",
                        "Std. dev.",
                        "Sample size") 

GorillaSummaryTable[, 1] <- c( round( mean(CBRyounger), 2),
                               round( mean(CBRolder), 2),
                               round( mean(CBRyounger) - mean(CBRolder), 2) )
GorillaSummaryTable[, 2] <- c( round( sd(CBRyounger), 2),
                                  round( sd(CBRolder), 2),
                                  NA )
GorillaSummaryTable[, 3] <- c( length(CBRyounger),
                                  length(CBRolder), 
                                  NA )

knitr::kable( surroundMaths(GorillaSummaryTable,
                            decDigits = c(1, 2, 0)),
              format = "latex",
              longtable = FALSE,
              booktabs = TRUE,
              escape = FALSE,
              digits = 1,
              align = c("r", "r", "r", "r"),
              caption = "A numerical summary of the gorillas data") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(row = 0, 
           bold = TRUE) %>%
  row_spec(row = 3,
           italic = TRUE) %>%
  kable_styling(font_size = 10)
```



## Exercises

Selected answers are available in Sect

::: {.exercise #GraphsAIS}
A study of athletes at the *Australian Institute of Sport* (AIS) measured numerous physical and blood measurements from high performance athletes [@data:Telford1991:sexsportsize].
The graph in Fig.\ \@ref(fig:AISfemalesport) compares the heights of females in two similar sports: basketball and netball.
(Netball was derived from basketball.)
How would you interpret the graph, regarding the heights of the athletes in the two sports?
:::

```{r}
data(AISsub)

subAIS <- subset(AISsub, 
                 Sex=="F" & (Sport=="BBall" | Sport=="Netball"))
subAIS$Sport <- factor(subAIS$Sport)
```

```{r AISfemalesport, fig.cap="The heights of female basketball and netball players attending the AIS", fig.align="center", fig.width=5, fig.height=3.5}
boxplot(Ht ~ Sport, 
        data = subAIS,
        las = 1,
        ylim = c(165, 200),
        col = plot.colour,
        xlab = "Sport",
        ylab = "Height (in cm)",
        main = "Boxplot of heights:\nbasketballers and netballers",
        names = c("Basketball", 
                  "Netball")
)
```




::: {.exercise #GreenBuilding}
In a study of the temperature in offices, @data:Paul2008:Comfort compared the temperature in three offices (during working hours) at Charles Sturt University (Australia); the data are summarised in Table\ \@ref(tab:OfficeTemps).
Using this information, draw the boxplot comparing the three offices.
What do we learn from this graph? 
:::


```{r OfficeTemps}
OfficeTemps <- array( dim = c(6, 3) )
colnames(OfficeTemps) <- c( "Office A", 
                            "Office B", 
                            "Office C")
rownames(OfficeTemps) <- c("Mean", 
                           "Minimum", 
                           "$Q_1$", 
                           "Median", 
                           "$Q_3$", 
                           "Maximum")

OfficeTemps[1, ] <- c(24.1, 25.3, 25.7)
OfficeTemps[2, ] <- c(16.4, 15.9, 20.1)
OfficeTemps[3, ] <- c(22.8, 23.8, 24.6)
OfficeTemps[4, ] <- c(24.4, 25.5, 26.1)
OfficeTemps[5, ] <- c(25.5, 26.9, 27.2)
OfficeTemps[6, ] <- c(27.4, 31.0, 30.3)

if( knitr::is_latex_output() ) {
    kable(surroundMaths(OfficeTemps,
                        decDigits = 1),
        format = "latex",
        longtable = FALSE,
        booktabs = TRUE,
        escape = FALSE, # For latex to work
        linesep = c( "", "", "", "", "", ""), # Add a bit of space between all rows. 
        caption = "A summary of the temperature (in degrees C) in three offices at CSU during working hours
according to current smoking status",
        align = c("c", "c"))   %>%
   kable_styling(full_width = FALSE, 
                 font_size = 10) %>%
  row_spec(0, bold = TRUE) %>% # Columns headings (row 0) in bold
  column_spec(0, bold = TRUE)     # Row headings in bold
}

if( knitr::is_html_output() ) {
  kable(OfficeTemps,
        format = "html",
        align = c("c", "c"),
        longtable = FALSE,
        caption = "A summary of the temperature (in degrees C) in three offices at CSU during working hours according to current smoking status",
        booktabs = TRUE) 
}
```


