# Summarising qualitative data {#SummariseQualData}
\index{Qualitative data!summarizing|(}


<!-- Introductions; easier to separate by format -->
```{r, child = if (knitr::is_html_output()) {'./introductions/13-SummaryQual-HTML.Rmd'} else {'./introductions/13-SummaryQual-LaTeX.Rmd'}}
```


## Introduction {#SummaryQual-Intro}

Except for very small amounts of data, understanding data is difficult without a summary.
As with quantitative data, qualitative data can be summarised by displaying the *distribution* of the data: the values taken by the variable, and how often they appear (Def.\ \@ref(def:Distribution)).

The distribution can be displayed using a frequency table (Sect.\ \@ref(QualitativeTables)) or a graph (Sect.\ \@ref(QualitativeGraphs)).
The distribution of qualitative data can be summarised numerically (Sect.\ \@ref(QualitativeNumbers)) by computing proportions (or percentages) or odds, modes; ordinal qualitative data can be also summarised numerically using medians and (under rare and very special circumstances) using means.


## Frequency tables for qualitative data {#QualitativeTables}
\index{Qualitative data!frequency table}

Qualitative data are typically collated in a *frequency table*.
The rows (or the columns) should list the *levels* of the variable (Sect.\ \@ref(def:QualitativeData)), and these should be *exhaustive* (cover all levels) and *exclusive* (observations belong to only one level).

For *nominal* data, the levels of the variables can be displayed alphabetically, by size, by personal preference, or other way: use the order most likely to be useful to readers.
For *ordinal* data, the natural order of the levels should almost always be used.


::: {.example #AVstudy name="Opinions of AV vehicles"}
Researchers [@pyrialakou2020perceptions] surveyed $400$ residents of Phoenix (Arizona) about their opinions of autonomous vehicles (AVs).
Demographic information was collected (Table\ \@ref(tab:AVtable1)), as well as respondents opinions of sharing roads with AVs (Table\ \@ref(tab:AVtable2)).

The gender of the respondent is *nominal* (two levels), while the age group is *ordinal* (six levels); the levels are in the rows.
The three questions about safety (Table\ \@ref(tab:AVtable2)) all yield *ordinal* responses (five levels, in columns). 
:::

```{r}
AVtable1 <- array( dim = c(8, 2) )

rownames(AVtable1) <- c("Female",
                        "Male",
                        "$18$ to $24$",
                        "$25$ to $34$",
                        "$35$ to $44$",
                        "$45$ to $54$",
                        "$55$ to $64$",
                        "$65$ and over")
colnames(AVtable1) <- c("Number",
                        "Percentage")
AVtable1[1, ] <- c(204, 51)
AVtable1[2, ] <- c(196, 49)
AVtable1[3, ] <- c(52, 13)
AVtable1[4, ] <- c(76, 19)
AVtable1[5, ] <- c(76, 19)
AVtable1[6, ] <- c(72, 18)
AVtable1[7, ] <- c(56, 14)
AVtable1[8, ] <- c(68, 17)

###########

AVtable2 <- array( dim = c(10, 4) )

colnames(AVtable2) <- c("",
                        "Driving near an AV",
                        "Cycling near an AV",
                        "Walking near an AV")
#rownames(AVtable2) <- c("Unsafe",
#                        "Somewhat unsafe",
#                        "Neutral",
#                        "Somewhat safe",
#                        "Safe")

rownames(AVtable2) <- c("Unsafe", "",
                   "Somewhat unsafe", "",
                   "Neutral", "",
                   "Somewhat safe", "",
                   "Safe", "")

AVtable2[ c(1, 3, 5, 7, 9), 2] <- c(58, 79, 96, 97, 70)
AVtable2[ c(1, 3, 5, 7, 9), 3] <- c(77, 104, 87, 76, 56)
AVtable2[ c(1, 3, 5, 7, 9), 4] <- c(63, 86, 103, 82, 66)

AVtable2[ c(2, 4, 6, 8, 10), 2] <- round( AVtable2[ c(1, 3, 5, 7, 9), 2] / 400 * 100, 0)
AVtable2[ c(2, 4, 6, 8, 10), 3] <- round( AVtable2[ c(1, 3, 5, 7, 9), 3] / 400 * 100, 0)
AVtable2[ c(2, 4, 6, 8, 10), 4] <- round( AVtable2[ c(1, 3, 5, 7, 9), 4] / 400 * 100, 0)

AVtable2[, 1] <- c("$n$", "Percent",
                        "$n$", "Percent",
                        "$n$", "Percent",
                        "$n$", "Percent",
                        "$n$", "Percent")

AVdemographics <- AVtable1[, 1]

AVquestions <- as.numeric( AVtable2[ c(1, 3, 5, 7, 9), 2:4] )
dim(AVquestions) <- c(5, 3)
rownames(AVquestions) <- rownames(AVtable2)[ c(1, 3, 5, 7, 9)]
colnames(AVquestions) <- colnames(AVtable2)[2:4]
```

```{r AVtable1}
if (knitr::is_html_output()) {
   knitr::kable(pad(AVtable1,
                   surroundMaths = TRUE,
                   targetLength = c(3, 2),
                   digits = 0),
               format = "html",
               longtable = FALSE,
               escape = FALSE,
               align = c("c", "c"),
               booktabs = TRUE,
               caption = "Demographic information for the AV data for $400$ respondents") %>%
    row_spec(0, bold = TRUE) %>%
    pack_rows( "Gender",
               start_row = 1,
               end_row = 2) %>%
    pack_rows( "Age group",
               start_row = 3,
               end_row = 8)
} else {
  knitr::kable(pad(AVtable1,
                   surroundMaths = TRUE,
                   targetLength = c(3, 2),
                   digits = 0),
               format = "latex",
               longtable = FALSE,
               escape = FALSE,
               align = c("c", "c"),
               booktabs = TRUE,
               caption = "Demographic information for the AV data for $400$ respondents") %>%
    kable_styling(font_size = 8) %>%
    row_spec(0, bold = TRUE) %>%
    pack_rows( "Gender",
               start_row = 1,
               end_row = 2) %>%
    pack_rows( "Age group",
               start_row = 3,
               end_row = 8) 
}
```


```{r AVtable2}
if (knitr::is_html_output()) {
   knitr::kable(pad(AVtable2,
                   surroundMaths = TRUE,
                   targetLength = c(0, 2, 3, 3),
                   digits = 0),
               format = "html",
               longtable = FALSE,
               escape = FALSE,
               align = c("r", "c", "c", "c"),
               linesep = c("", 
                           "\\addlinespace",
                           "", 
                           "\\addlinespace",
                           "", 
                           "\\addlinespace",
                           "", 
                           "\\addlinespace",
                           "", 
                           "\\addlinespace"),  # Otherwise adds a space after five lines... 
               booktabs = TRUE,
               caption = "Responses to three scenarios for the AV data for $400$ respondent") %>%
    row_spec(0, bold = TRUE) %>%
    column_spec(1, bold = TRUE) 
} else {
  
  knitr::kable(pad(AVtable2,
                   surroundMaths = TRUE,
                   targetLength = c(0, 2, 3, 3),
                   digits = 0),
               format = "latex",
               longtable = FALSE,
               escape = FALSE,
               align = c("r", "c", "c", "c"),
               col.names = c("",
                             "near an AV",
                             "near an AV",
                             "near an AV"),
               linesep = c("", 
                           "\\addlinespace",
                           "", 
                           "\\addlinespace",
                           "", 
                           "\\addlinespace",
                           "", 
                           "\\addlinespace",
                           "", 
                           "\\addlinespace"),  # Otherwise adds a space after five lines... 
               booktabs = TRUE,
               caption = "Responses to three scenarios for the AV data for $400$ respondent") %>%
    kable_styling(font_size = 8) %>%
    row_spec(0, bold = TRUE) %>%
    column_spec(1, bold = TRUE) %>%
    add_header_above( c(" " = 1,
                        " " = 1,
                        "Driving" = 1,
                        "Cycling" = 1,
                        "Walking" = 1),
                      line = FALSE,
                      bold = TRUE)

  
}
```          

## Graphs {#QualitativeGraphs}
\index{Qualitative data!graphs}\index{Graphs!qualitative data}


Three options for graphing qualitative data are:

* *Dot chart* (Sect.\ \@ref(DotChartsOneQual)):
  Usually a good choice.
* *Bar chart* (Sect,\ \@ref(BarCharts)):
  Usually a good choice.
* *Pie chart* (Sect.\ \@ref(PieCharts)):
  Only useful in special circumstances, and can be harder to interpret.

For *nominal* data, the levels of the variables can be displayed alphabetically, by size, by personal preference, or other way: use the order most likely to be useful to readers.
For *ordinal* data, the natural order of the levels should almost always be used.

Sometimes these graphs are also used for quantitative data with a small number of possible options.
Sometimes, graphs used for quantitative data (Sect.\ \@ref(QuantitativeGraphs)) may be useful for discrete quantitative data if many values are possible.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The purpose of a graph is to display the information in the clearest, simplest possible way, to facilitate understanding the message(s) in the data.
:::


### Dot charts (qualitative data) {#DotChartsOneQual}
\index{Dot charts!one qualitative variable}\index{Graphs!dot chart}

Dot charts indicate the counts (or the corresponding percentages) in each level, using dots on a line *starting at zero*.
The levels can be on the horizontal or vertical axis, and the counts or percentages in the other direction.
Placing the level names on the vertical axis often makes for easier reading, and space for long labels. 


::: {.example #DotPlotsQual name="Dot plots"}
For the AV study in Example\ \@ref(exm:AVstudy), a dot chart of the age group of respondents is shown in Fig.\ \@ref(fig:AVDotBarPie) (top left panel).
:::

```{r, AVDotBarPie, fig.align='center', fig.cap="The age group of respondents in the AV study. All the graphs present the same data.", out.width='100%', fig.width=9, fig.height=5.25, fig.show='hold'}

par(mfrow = c(2, 2))
par( mar = c(6, 4, 4, 2) + 0.1 )
cols <- viridis(9)[3:8]

AVdemographics2 <- AVdemographics
AVdemographics2.names <- names(AVdemographics2)

AVdemographics2.names <- gsub(pattern = "\\$",
                              replacement = "",
                              x = AVdemographics2.names)
names(AVdemographics2) <- AVdemographics2.names


dotchart(AVdemographics2[3:8],
         pch = 19, 
         labels = rep("                               ", 6), # Otherwise, these are coloured and hard to read; reinstate after
         xlim = c(0, 80),
         xlab = "Numbers of respondents",
         main = "Age group of respondents\nin the AV study",
                col = cols)
text(labels = AVdemographics2.names[3],
     x = -5,
     y = 1,
     adj = 1)
text(labels = AVdemographics2.names[4],
     x = -5,
     y = 2,
     adj = 1)
text(labels = AVdemographics2.names[5],
     x = -5,
     y = 3,
     adj = 1)
text(labels = AVdemographics2.names[6],
     x = -5,
     y = 4,
     adj = 1)
text(labels = AVdemographics2.names[7],
     x = -5,
     y = 5,
     adj = 1)
text(labels = AVdemographics2.names[8],
     x = -5,
     y = 6,
     adj = 1)
###

barplot(AVdemographics2[3:8],
        las = 2,
        ylim = c(0, 80),
        ylab = "Numbers of respondents",
        main = "Age group of respondents\nin the AV study",
                col = cols)

###
par( mar = c(0.4, 0.4, 3.4, 0.4) + 0.1 )

pie(AVdemographics2[3:8],
        main = "Age group of respondents\nin the AV study",
                col = cols)

###

par( mar = c(0.1, 0.1, 3.4, 0.1) + 0.1 )
plotrix::pie3D( AVdemographics2[3:8],
                main = "Age group of respondents\nin the AV study",
                theta = pi/4,
                labels = names(AVdemographics2[3:8]),
                labelcex = 0.9,
                col = cols)
```

For dot charts:

* place the qualitative variable on the horizontal or vertical axis (and label with the levels of the variable).
* use counts or percentages on the other axis.
* for nominal data, *think about the most helpful order* for the levels.


### Bar charts {#BarCharts}
\index{Bar chart}\index{Graphs!bar chart}

Bar charts indicate the counts in each category using bars *starting from zero*.
As with dot charts, the levels can be on the horizontal or vertical axis, but placing the level names on the vertical axis often makes for easier reading, and room for long labels. 


::: {.example #BarchartQual name="Dot plots"}
For the AV study in Example\ \@ref(exm:AVstudy), a bar chart of the age group of respondents is shown in Fig. \@ref(fig:AVDotBarPie) (top right panel).
:::

For bar charts:

* place the qualitative variable on the horizontal or vertical axis (and label with the levels of the variable).
* use counts or percentages on the other axis.
* for nominal data, levels can be ordered any way: *Think about the most helpful order*.
* bars have gaps between bars, as the bars represent distinct categories.


Bar charts have gaps between all of the bars.
In contrast, the bars in histograms are butted together (except when an interval has a count of zero), as the variable-axis represent a continuous numerical scale.



### Pie charts {#PieCharts}
\index{Pie chart}\index{Graphs!pie chart}

In pie charts, a circle is divided into segments proportional to the number in each level of the qualitative variable.


::: {.example #PieChartsQual name="Dot plots"}
For the AV study in Example\ \@ref(exm:AVstudy), a pie chart of the age group of respondents is shown in Fig. \@ref(fig:AVDotBarPie) (bottom left panel).
:::


Pie charts may present challenges:

* Pie charts only work when graphing parts of a whole.
* Pie charts only work when *all* options are present ('exhaustive').
* Pie charts are difficult to use with levels having zero counts, or small counts.
* Pie charts are difficult to read with many categories present.
* Pie charts are hard to read: Humans compare *lengths* (as in bar and dot charts) better than *angles* (as in pie charts) [@data:Friel:Graphs].


::: {.thinkBox .think data-latex="{iconmonstr-light-bulb-2-240.png}"}
Would a pie chart be *appropriate* for graphing the percentage of people who use Firefox, Chrome, and Safari as web browsers?\label{thinkBox:WhenPiechart}

`r if (knitr::is_latex_output()) '<!--'`
`r webexercises::hide()`
A pie chart is *not suitable*.
The three browsers are not *mutually exclusive* (people can use more than one of these browsers) nor *exhaustive* (some people may use browsers not listed). 
For example, the percentages *could* be that $65$% use Firefox, $84$% use Chrome, and $20$% use Safari. 
These add to more than $100$%.
`r webexercises::unhide()`
`r if (knitr::is_latex_output()) '-->'`
:::



### Comparing pie, bar and dot charts {#CompareBarPie}
\index{Bar chart!compared to other graphs}\index{Pie chart*!compared to other graphs}\index{Dot chart!compared to other graphs}

Consider the pie chart in Fig.\ \@ref(fig:AVDotBarPie) (bottom left panel).
Determining *which* age groups have the most respondents is hard.
The equivalent bar chart (or dot chart) makes the comparison easy: clearly the youngest age group has the smallest representation, while the\ $25$ to\ $34$ and the\ $35$ to\ $44$ age groups have the most respondents.
The *tilted* pie chart makes this comparison even harder (Fig.\ \@ref(fig:AVDotBarPie), bottom right panel).

Recall that the *purpose of a graph is to display information in the clearest, simplest possible way, to help the reader understand the message(s) in the data*.
A pie chart often makes the message hard to see [@siegrist1996use].\index{Pie chart!warnings}



<iframe src="https://learningapps.org/watch?v=pf4om4k5t22" style="border:0px;width:100%;height:500px" allowfullscreen="true" webkitallowfullscreen="true" mozallowfullscreen="true"></iframe>



## Numerical summaries {#QualitativeNumbers}


### Parameters and statistics {#Statistics2}
\index{Parameter}\index{Statistic}

In quantitative *research* (Sect.\ \@ref(TypesOfResearch)), both qualitative and quantitative *data* are summarised and analysed numerically.
In the following sections, methods for numerically summarising quantitative variables are described.
As described in Sect.\ \@ref(Statistics), numerical quantities are computed from one of countless possible samples, even though the whole population is of interest.

A statistic (a sample value) is a numerical value estimating the unknown parameter (population value).
Since countless possible samples are possible (Sect.\ \@ref(IdeaOfSampling)), countless possible values for the statistic---all of which are estimates of the value of the parameter---are possible.
The value of the statistic that is observed depends on which one of the countless possible samples is (randomly) selected.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
The RQ identifies the population, but in practice only one of the many possible samples is studied.
*Statistics* are estimates of *parameters*, and the value of the *statistic* is not the same for every possible *sample*.
We only observe one value of the statistic from our single sample.
:::


### Proportions and percentages {#QualitativeProportionsPercentages}
\index{Proportions}\index{Percentages}

Qualitative data can be summarised using *proportions* or *percentages*.
These can be given instead of, or with, the counts (as in Tables\ \@ref(tab:AVtable1) and\ \@ref(tab:AVtable2)).


::: {.definition #Proportion name="Proportion"}
A *proportion* is a fraction out of a total, and is a number between\ $0$ and\ $1$.
:::


::: {.definition #Percentage name="Percentages"}
A *percentage* is a proportion, multiplied by $100$.
In this context, percentages are numbers between $0$% and $100$%.
:::


*Population* proportions are almost always unknown. 
Instead, the *population* proportion (the parameter), denoted $p$, is estimated by a *sample* proportion (a statistic), denoted by $\hat{p}$. 


::: {.pronounceBox .pronounce data-latex="{iconmonstr-microphone-7-240.png}"}
The symbol $\hat{p}$ is pronounced 'pee-hat', and refers to a *sample* proportion.
:::


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
As always, only one of the many possible samples is studied.
*Statistics* are estimates of *parameters*, and the value of the *statistic* is not the same for every possible *sample*.
:::




::: {.example #AVProportionsPercentages name="Proportions and percentages"}
Consider the AV data in Table\ \@ref(tab:AVtable1), summarising results from a sample of $n = 400$ respondents.
The *sample proportion* of respondents aged $25$ to $34$ is $76\div 400$, or $0.19$.
The *sample percentage* of respondents aged $25$ to $34$ is $76\div 400 \times 100$, or $19$%, as given in the table.
:::



### Odds {#QualOdds}
\index{Odds}

For the AV data in Table\ \@ref(tab:AVtable1), the number of females is slightly larger than the number of males.
More specifically, the *ratio* of females to males is $204\div 196 = 1.04$; that is, there are $1.04$ *times* as many females as males.
This value of $1.04$ is the *odds* that a respondent is female.
An alternative interpretation is that there $1.04\times 100 = 104$ females for every $100$ males.

While proportions and percentages are computed as the number of interest divided by the *total number*, the *odds* are computed as the number of interest divided by *the remaining number*.


::: {.definition #Odds name="Odds"}
The *odds* are the number (or proportion, or percentage) of times that an event *happens*, divided by the number (or proportion, or percentage) of times that the event does *not happen*:  
\[
         \text{Odds} = \frac{\text{Number of times event happens}}{\text{Number of times event doesn't happen}}
\]
or (equivalently)  
\[
         \text{Odds} 
          = 
            \frac{\text{Proportion of times event happens}}
                 {\text{Proportion of times event doesn't happen}}.
\]
The *odds* are how many *times* an event *happens* compared to the event *not happening*.
:::


::: {.example #AVOddsMale name="Odds males"}
The AV data in Table\ \@ref(tab:AVtable1) includes $204$ females and $196$ males.
The *odds* that a respondent is female is $1.04$, as found above.
The odds is greater than one, as the number of females is larger than the number of males.

The *odds* that a respondent is male is $196/204 = 0.96$; that is, there are $0.96$ *times* as many males as females.
The odds is less than one, as the number of males is smaller than the number of females.
Alternatively, there are $96$ males for every $100$ females.
:::

::: {.example #AVOdds name="Odds"}
Consider the AV data in Table\ \@ref(tab:AVtable1), summarising results from a sample of $n = 400$ respondents.

The *percentage* of respondents that are female is $204\div400\times 100 = 51$%.
The *odds* that a respondent is female is $204\div(400 - 204) = 1.04$.

Similarly, the percentage of respondents aged $18$\ to $24$ is $52/400\times 400 = 13$%.
The *odds* that a respondent is aged $18$\ to $24$ is $52/(400 - 52) = 0.15$; that is, the odds that a respondent is aged $18$\ to\ $24$ is $0.15$.
This means that the number of respondents aged $18$\ to $24$ is $0.15$ times (i.e., less) than the number of respondents aged over $24$.

The *odds* that a respondent is aged $18$\ to $54$ is $(52 + 76 + 76 + 72)/(56 + 68) = 2.23$; that is, the odds that a respondent is aged $18$\ to\ $54$ is $2.23$.
This means that the number of respondents aged $18$\ to $54$ is $2.23$ times (i.e., greater) than the number of respondents aged over  $54$
:::


Take care interpreting odds:
  
* odds are *greater* than $1$: the event is *more* likely to happen than to not happen.
* odds are *equal to* $1$: the event is just as likely to happen as it is to not happen.
* odds are *less* than $1$: the event is *less* likely to happen than to not happen.


::: {.thinkBox .think data-latex="{iconmonstr-light-bulb-2-240.png}"}
What are the odds of rolling a 
`r if (knitr::is_latex_output()) {
   '\\Largedice{1}'
} else {
   '<span class="larger-die">&#9856;</span>'
}`
on a die?\label{thinkBox:Roll2}
`r if( knitr::is_html_output()) {
longmcq( c(
   "About 0.1666", 
   "6",
   "0.5",
   "1/6",
   answer = "0.2") )}`
:::




*Population* odds are almost always unknown. 
Instead, the *population* odds (the parameter) is estimated by a *sample* odds (a statistic). 


### Modes {#Modes}
\index{Mode}

Because qualitative data has *levels* (Def.\ \@ref(def:Levels)), all qualitative data (nominal; ordinal) can be numerically summarised by *counting* the number of observations in each level (or computing the percentage of observations in each level).
The *mode* is the level (or levels) with the *most* observations.


::: {.definition #Mode name="Mode"}
A *mode* is the level (or levels) of a qualitative variable with the most observations.
:::


::: {.example #OrdinalModes name="Modes" }
Consider again the data in Tables\ \@ref(tab:AVtable1) and\ \@ref(tab:AVtable2).
'Gender' is *nominal* qualitative; age group is *ordinal* qualitative.
The responses to the question are *ordinal*.
The mode could be used to summarise each variable:

* The *mode* for gender is 'Female' (with $204$ respondents).
* The *mode* age groups are $25$ to\ $34$ and $35$ to\ $44$ (both with $76$ respondents).
* The *modal* response to the question about *driving* near AVs is 'Somewhat safe' ($97$ respondents).
* The *modal* response to the question about *cycling* near AVs is 'Somewhat unsafe' ($104$ respondents).
* The *modal* response to the question about *walking* near AVs is 'Neutral' ($103$ respondents).
:::


*Population* modes are almost always unknown. 
Instead, the *population* mode (the parameter) is estimated by a *sample* mode (a statistic). 


### Medians (and means) for ordinal data {#QualitativeMedianMeans}
\index{Median!qualitative ordinal data}

*Ordinal* data can be summarised in ways that nominal data cannot be, since *ordinal* data have levels with a natural order. 
Ordinal qualitative data, but *not* nominal data, can be summarised using *medians*, by finding the location of the response that is in the middle, when the levels from all individuals are placed in order.


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
Medians can be used to summarise *quantitative data* and *ordinal* data, but *never* for nominal data.
:::


::: {.example #OrdinalMedians name="Medians"}
Consider again the data in Tables\ \@ref(tab:AVtable1) and\ \@ref(tab:AVtable2).
'Gender' is *nominal* qualitative, so medians are not appropriate.
However, the other variables are *ordinal*, so medians could be used to summarise each variable.
Since $n = 400$, the median response will be halfway between the location of the $200$th and $201$st response when ordered:

* the *median* age group is $35$ to\ $44$ (observations $200$ and $201$ fall into this level).
* the *median* response to the driving question is 'Neutral' (observations $200$ and $201$ fall in this level).
* the *median* response to the cycling question is 'Neutral' (observations $200$ and $201$ fall in this level).
* the *median* response to the walking question is 'Neutral' (observations $200$ and $201$ fall in this level).
:::


Means are generally not suitable for numerically summarising qualitative data.
However, *ordinal* data *may be* [numerically summarised like quantitative data](#NumericalQuant) in *rare and very special circumstances*: only when

* the levels are considered 'equally spaced'; *and*
* assigning a number to each level is appropriate (perhaps using a mid-point for numerical groups).


::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
Means can sometimes be used to summarise *quantitative ordinal* data, in rare and very special circumstances.
Means are *never* appropriate for nominal data.
:::

*Population* medians and means are almost always unknown. 
Instead, the *population* median and mean (both parameters) are estimated by a *sample* median or mean (both statistics). 


::: {.example #OrdinalMeans name="Means of ordinal data"}
Consider again the data in Tables\ \@ref(tab:AVtable1) and\ \@ref(tab:AVtable2).
Means are *not* suitable for summarising Gender, since it is *nominal*.
Means are also *not* suitable for summarising Age, since the final age group is open-ended (so computing a midpoint is not possible).

However, the mean response to the statements *may* be sensible *only if the levels are considered equally spaced* (i.e., the distances between successive points on the five-point scale are considered equal).
Then, assigning $1$ to 'Unsafe', $3$ to 'Neutral', up to $5$ for 'Safe':

* the *mean* response to the driving question is $3.11$ (slightly safer than 'Neutral').
* the *mean* response to the cycling question is $2.83$ (slightly unsafer than 'Neutral').
* the *mean* response to the walking question is $3.01$ (essentially 'Neutral').
:::


## Example: water access {#WaterAccessQual}

A study of three rural communities in Cameroon [@lopez2022farmers] recorded data about access to water (see Sect. \@ref(WaterAccessQuant)).
Numerous qualitative variables are recorded; some are displayed in Fig.\ \@ref(fig:WaterAcessQual), and a summary table in Table\ \@ref(tab:WaterAccessQual).
Notice that the levels of the two ordinal variables are displayed in the natural order.

The distance to the nearest water source is usually less than $1$\ km, and the wait at the source often over $1$\ mins.
The most common water source is a bore ($68.6$%).



```{r WaterAccessQual}
data(WaterAccess)

WaterAccessTable2A <- array( dim = c(6, 3) )

 
findOdds <- function(x){
  tablex <- table(x)
  oddsTab <- array( dim = length(tablex))

  for (i in 1:length(tablex)){
    oddsTab[i] <- tablex[i] / sum(tablex[-i])
  }
  oddsTab
}

qualSummary <- function(x){
  out <- cbind( pad2( t(t(table(x))),
                     targetLength = 2,
                     surroundMaths = TRUE),
                pad2( t( t( table(x) ) / sum(table(x)) * 100),
                      digits = 1,
                     targetLength = 4,
                     surroundMaths = TRUE),
                pad2( t( t(findOdds(x)) ),
                      digits = 2,
                     targetLength = 4,
                     surroundMaths = TRUE ) 
  )

  out
} 

qualSummary <- function(x){
  tab <- table(x)
  
  out <- array( dim = length(tab)) 
  
#  for (i in 1:length(tab)){
    
    # out <- cbind( pad( t(t(tab[i])),
    #                    targetLength = 2,
    #                    surroundMaths = TRUE),
    #               pad( format( round(t( t( tab[i] ) / sum(tab) * 100), 1), nsmall = 1),
    #                    targetLength = 4,
    #                    surroundMaths = TRUE),
    #               pad( format( round( t( t(findOdds(x)[i]) ), 2), nsmall = 2),
    #                    targetLength = 4,
    #                    surroundMaths = TRUE ) 
    # )
    out <- cbind( pad( t(t(tab)),
                       targetLength = 2,
                       digits = 0,
                       surroundMaths = TRUE),
                  pad( t( t( tab ) / sum(tab) * 100), 
                       targetLength = 4,
                       digits = 1,
                       surroundMaths = TRUE),
                  pad( t( t(findOdds(x)) ),
                        digits = 2,
                       targetLength = 4,
                       surroundMaths = TRUE ) 
    )
#  }
  
  out
} 


WaterAccessTable2A <- rbind(qualSummary(WaterAccess$SourceDistance), 
                            qualSummary(WaterAccess$SourceQueueTime) )

rownames(WaterAccessTable2A) <- c("Under $100$\\ m",
                                  "$100$\\ m to $1000$\\ m",
                                  "Over $1000$\\ m",
                                  "Under $5$\\ min",
                                  "$5$ to $15$\\ min",
                                  "Over $15$\\ min") 
 

colnames(WaterAccessTable2A) <- c("Num.",
                                  "\\%",
                                  "Odds")


###

WaterAccessTable2B <- array( dim = c(4, 3) )

rownames(WaterAccessTable2B) <- c( levels(WaterAccess$WaterSource) )
 

WaterAccessTable2B <- qualSummary(WaterAccess$WaterSource)

colnames(WaterAccessTable2B) <- c("Num.",
                                  "\\%",
                                  "Odds")


if( knitr::is_latex_output() ) {
  T1 <- kable( WaterAccessTable2A,
        format = "latex",
        longtable = FALSE,
        escape = FALSE,
        col.names = colnames(WaterAccessTable2A),
        booktabs = TRUE,
        align = c("r")) %>%
    row_spec(0, bold = TRUE) %>%
    pack_rows("Distance to water source",
              start_row = 1,
              end_row = 3,
              bold = TRUE) %>%
    pack_rows("Wait time at water source",
              start_row = 4,
              end_row = 6,
              bold = TRUE)
  
  T2 <- kable( WaterAccessTable2B,
        format = "latex",
        longtable = FALSE,
        escape = FALSE,
        col.names = colnames(WaterAccessTable2B),
        booktabs = TRUE,
        align = c("c")) %>%
    row_spec(0, bold = TRUE) %>%
    pack_rows("Water source",
              start_row = 1,
              end_row = 4,
              bold = TRUE)
  
    out <- knitr::kables(list(T1, T2),
                       format = "latex",
                       label = "WaterAccessQual",
                       caption = "Summarising some qualitative data in the water-access study") %>% 
    kable_styling(font_size = 8)
  out2 <- prepareSideBySideTable(out, 
                                 gap = "\\enskip") 
  out2
}
if( knitr::is_html_output() ) {
 kable( rbind(WaterAccessTable2A, 
              WaterAccessTable2B),
               format = "html",
               longtable = FALSE,
               col.names = colnames(WaterAccessTable2A),
               booktabs = TRUE,
               align = c("r"),
               caption = "Summarising some qualitative data in the water-access study") %>%
   kable_styling(full_width = FALSE) %>%
    pack_rows("Distance to water soource",
              start_row = 1,
              end_row = 3,
              bold = TRUE) %>%
    pack_rows("Wait time at water source",
              start_row = 4,
              end_row = 6,
              bold = TRUE) %>%
    pack_rows("Water source",
              start_row = 7,
              end_row = 10,
              bold = TRUE)
}
```

```{r WaterAcessQual, fig.align="center", fig.cap="The age of the women, the number of people in the household, and the number of children in the household for the water-access study. (Some data are missing.)", out.width='100%', fig.width=6.5, fig.height=1.7}

par(mfrow = c(1, 3),
    mar = c(3, 2, 4, 2) )

dotchart( as.numeric(table(WaterAccess$SourceDistance)),
          labels = names(table(WaterAccess$SourceDistance)),
     xlab = "Number",
     ylab = "",
     main = "Distance to\nwater source",
     xlim = c(0, 60),
     las = 1,
     pch = 19)

dotchart( as.numeric(table(WaterAccess$SourceQueueTime)),
          labels = names(table(WaterAccess$SourceQueueTime)),
     xlab = "Number",
     ylab = "",
     main = "Queue wait at\nwater source",
     xlim = c(0, 55),
     las = 1,
     pch = 19)

dotchart( as.numeric(table(WaterAccess$WaterSource)),
          labels = names(table(WaterAccess$WaterSource)),
     xlab = "Number",
     ylab = "",
     main = "Water sources",
     xlim = c(0, 100),
     las = 1,
     pch = 19)
```



\index{Qualitative data!summarizing|)}

## Chapter summary {#SummaryQual-Summary}

A qualitative variable can compared numerically summarized using the *mode*, *percentages* or *odds*.
Ordinal qualitative variables maybe numerically summarised using a *median* or, under rare and very special circumstances, a *mean*.



## Quick review questions {#SummaryQual-QuickReview}

Are the following statements *true* or *false*?

::: {.webex-check .webex-box}
1. Nominal data can be summarised using a median. \tightlist
`r if( knitr::is_html_output() ) {torf(answer = FALSE)}`
1. Ordinal data can be summarised using a mode.
`r if( knitr::is_html_output() ) {torf(answer = TRUE)}`
1. Odds are the ratio of how often an event occurs, to how often it does *not* occur.
`r if( knitr::is_html_output() ) {torf(answer = TRUE)}`
1. Proportions and percentages are the same.
`r if( knitr::is_html_output() ) {torf(answer = FALSE)}`
:::




## Exercises {#SummariseQualDataExercises}

Selected answers are available in Sect.\ \@ref(SummariseQualDataAnswer).

::: {.exercise #SpiderMonkeys}
A study of spider monkeys [@data:Chapman1990:SpiderMonkeys] examined the types of social groups present
`r if (knitr::is_latex_output()) {
   '(Table\\ \\@ref(tab:SpiderMonkeysLATEX)).'
} else {
   '(Table\\ \\@ref(tab:SpiderMonkeysHTML)).'
}`
Construct a suitable plot, and explain what the data reveal about the social groups of spider monkeys.
:::


```{r SpiderMonkeysData}
namesSM <- c("Solitary", 
             "All males", 
             "Female + no young", 
             "Mixed young", 
             "Mixed + no young", 
             "One female + offspring", 
             "Many females + offspring",
             NA)

SpiderMonkeyTable <- array( dim = c(8, 2) )
SpiderMonkeyTable[, 2] <- c(8, 3, 2, 15, 1, 23, 48, NA)
SpiderMonkeyTable[, 1] <- namesSM

rownames(SpiderMonkeyTable) <- namesSM
```

```{r}
if( knitr::is_latex_output() ) {
   T1 <- kable( pad( SpiderMonkeyTable[1:4,],
                     surroundMaths = TRUE,
                     targetLength = 2,
                     digits = 0), 
       format = "latex",
       longtable = FALSE,
       booktabs = TRUE,
       escape = FALSE,
       digits = 1,
       align = c("r", "c"),
       row.names = FALSE,
       col.names = c("Social group",
                     "Number")) %>%
  row_spec(row = 0, 
           bold = TRUE) 

   
  T2 <- kable( pad( SpiderMonkeyTable[5:8,],
                    surroundMaths = TRUE,
                     targetLength = 2,
                     digits = 0), 
       format = "latex",
       longtable = FALSE,
       booktabs = TRUE,
       escape = FALSE,
       digits = 1,
       align = c("r", "c"),
       row.names = FALSE,
       col.names = c("Social group",
                     "Number")) %>%
  row_spec(row = 0, 
           bold = TRUE)   
  
  
  out <- knitr::kables(list(T1, T2),
                       format = "latex",
                       label = "SpiderMonkeysLATEX",
                       caption = "Social groups for soldier monkeys") %>% 
    kable_styling(font_size = 8)
  out2 <- prepareSideBySideTable(out, 
                                 gap = "\\qquad") 
  out2
}
```


```{r SpiderMonkeysHTML}
if( knitr::is_html_output() ) {

kable( pad( SpiderMonkeyTable,
                     surroundMaths = TRUE,
                     targetLength = 2,
                     digits = 0), 
       format = "html",
       longtable = FALSE,
       booktabs = TRUE,
       escape = FALSE,
       digits = 1,
       align = c("r", "c"),
       row.names = FALSE,
       col.names = c("Social group",
                     "Number"),
       caption = "Social groups in spider monkeys") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(row = 0, 
           bold = TRUE) 
}
```


```{r LCgraphsHTML, fig.cap="Graphs of deaths from lung cancer in Fredericia (Denmark) from 1968 and 1971", fig.align="center", fig.width=5, fig.height=7.5}
if (knitr::is_html_output()) {
  
  data(DanishLC)
  
  par( mfrow = c(2, 1), 
       mar = c(5.5, 5, 4.1, 1))
  
  pie( DanishLC$Cases[DanishLC$City == "Fredericia"],
       labels = c("40 to 54", 
                  "55 to 59", 
                  "60 to 64", 
                  "65 to 69", 
                  "70 to 74", 
                  "Over 74"),
       col = viridis(6),
       mar = c(0.5, 0.5, 4.1, 0.5),
       main = "Pie chart of lung cancer deaths\nby age group")
  
  barplot(DanishLC$Cases[DanishLC$City == "Fredericia"], 
          las = 2, 
          names.arg = c("40 to 54", 
                        "55 to 59", 
                        "60 to 64", 
                        "65 to 69", 
                        "70 to 74", 
                        "Over 74"),
          col = viridis(6),
          ylim = c(0, 12),
          sub = "Age group", 
          ylab = "Number of deaths", 
          main = "Bar chart of lung cancer deaths\nby age group")
}
```

```{r LCgraphsPDF, fig.cap="Graphs of deaths from lung cancer in Fredericia (Denmark) from 1968 and 1971", fig.align="center", fig.width=10, fig.height=3.5, out.width='95%'}
if (knitr::is_latex_output()) {
  data(DanishLC)
  
  par( mar = c(0.5, 0.5, 0.5, 0.5) )
  
  nf <- layout( matrix(c(1, 2), 
                       ncol = 1),
                widths = c(3, 1) )
  
  par( mfrow = c(1, 2), 
       mar = c(5.5, 5, 4.1, 1))
  
  pie( DanishLC$Cases[DanishLC$City == "Fredericia"],
       labels = c("40 to 54", 
                  "55 to 59", 
                  "60 to 64", 
                  "65 to 69", 
                  "70 to 74", 
                  "Over 74"),
       col = viridis(6),
       main = "Pie chart of lung cancer deaths\nby age group")
  
  barplot(DanishLC$Cases[DanishLC$City == "Fredericia"], 
          las = 2, 
          names.arg = c("40 to 54", 
                        "55 to 59", 
                        "60 to 64", 
                        "65 to 69", 
                        "70 to 74", 
                        "Over 74"),
          col = viridis(6),
          ylim = c(0, 12),
          sub = "Age group", 
          ylab = "Number of deaths", 
          main = "Bar chart of lung cancer deaths\nby age group")
}
```


::: {.exercise #GraphsCars}
A study [@henderson1981building] recorded the number of cylinders in many models of cars: eleven cars had four cylinders, seven had six cylinders, and fourteen had eight cylinders.
The *number* of cylinders is quantitative discrete, but with so few different values, this data could be plotted with some of the graphs used for qualitative data.
For these data:

:::::: {.cols data-latex=""}
::: {.col data-latex="{0.45\textwidth}"}

1. Produce a dot chart.
2. Produce a histogram.

:::

::: {.col data-latex="{0.05\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::

::: {.col data-latex="{0.45\textwidth}"}

3. Produce a bar chart.
4. Produce a pie chart.

:::
::::::

\null\smallskip
What graph do you think is best?
Why?
:::





::: {.exercise #GraphSurveyData}
A 
`r if (knitr::is_latex_output()) {
   'survey of voice assistants'
} else {
   '[survey of voice assistants](https://www.nielsen.com/us/en/insights/news/2018/smart-speaking-my-language-despite-their-vast-capabilities-smart-speakers-all-about-the-music.print.html)'
}`
(such as Amazon Echo; Google Home; etc.) conducted by
`r if (knitr::is_latex_output()) {
   'Nielsen'
} else {
   '[Nielsen](https://www.nielsen.com/au/en.html)'
}`
asked respondents to indicate how they used their voice assistant.
The options were:

:::::: {.cols data-latex=""}
::: {.col data-latex="{0.46\textwidth}"}

* Listening to music;
* Listen to news;
* Chat with voice assistant for fun;
* Use alarms, timer.

:::

::: {.col data-latex="{0.03\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::

::: {.col data-latex="{0.46\textwidth}"}

* Search for real-time information (e.g., traffic; weather);
* Search for factual information (e.g., trivia; history);

:::
::::::

What would be the best graph for displaying respondents answers?
Would a pie chart be suitable? Explain your answer.
:::


::: {.exercise #OrdinalMedians}
In a study of the taste of bread with varying salt and fibre content [@gkebski2019impact], researchers recorded information from the $300$ subjects, including gender, place of residence, and the subjects' responses to the statement 'Rolls with lower salt content taste worse than regular ones', on a five-point ordinal scale from 'Strongly Agree' to 'Strongly Disagree'.

1. Identify the variables, then classify them as nominal or ordinal.
1. For which variables is a mode an appropriate summary (if any)?
1. For which variables is a median an appropriate summary (if any)?
1. For which variables is a mean an appropriate summary (if any)?
1. Compute the above where appropriate.
:::


```{r Bread}
BreadTable  <- array( dim = c(11, 3))

colnames(BreadTable) <- c("",
                          "Number",
                          "Percentage")

BreadTable[, 1] <- c("Female",
                     "Male",
                     "Rural",
                     "City up to $20\\, 000$ residents",
                     "City $20\\, 000$ to $100\\, 000$ residents",
                     "City $> 100\\, 000$ residents",
                     "Strongly agree",
                     "Agree",
                     "Neutral",
                     "Disagree","Strongly disagree"
                     )
BreadTable[, 2] <- c(150,
                     150,
                     49,
                     38,
                     83,
                     130,
                     30,
                     84,
                     78,
                     66,
                     42)
BreadTable[, 3] <- c(50,
                     50,
                     16,
                     13,
                     28,
                     43,
                     10,
                     28,
                     26,
                     22,
                     14)

if( knitr::is_latex_output() ) {
  knitr::kable(pad(BreadTable,
                   surroundMaths = TRUE,
                   targetLength = c(0, 0, 0),
                   digits = 0),
        format = "latex",
        longtable = FALSE,
        booktabs = TRUE,
        escape = FALSE,
        align = c("r", "c", "c"),
        linesep = c("", "\\addlinespace", "", "", "", "\\addlinespace", "", "", "", ""), 
        caption = "The bread-tasting data") %>%
   kable_styling(font_size = 8) %>%
   pack_rows("Gender", 1, 2) %>%
   pack_rows("Place of residence", 3, 6) %>%
   pack_rows("Response to statement", 7, 11) %>%
   row_spec(0, bold = TRUE)
}

if( knitr::is_html_output() ) {
  BreadTable[6, 1] <- "City more than $100\\, 000$ residents" 
  
  knitr::kable(pad(BreadTable,
                   surroundMaths = TRUE,
                   targetLength = c(0, 0, 0),
                   digits = 0),
        format = "html",
        longtable = FALSE,
        booktabs = TRUE,
        align = c("r", "c", "c"),
        caption = "The bread-tasting data")%>%
   #kable_styling(font_size = 10) %>%
   pack_rows("Gender", 1, 2) %>%
   pack_rows("Place of residence", 3, 6) %>%
   pack_rows("Response to statement", 7, 11) %>%
   row_spec(0, bold = TRUE)
}
```


::: {.exercise #ReclaimedWater}
A study asked $231$ farmers what they considered to be the advantages and disadvantages of using reclaimed water [@lopez2022farmers].
The responses are shown in Table\ \@ref(tab:ReclaimedWater) (not all farmers responded).

1. Produce two bar charts to display the data.
1. Produce two dot charts to display the data.
1. Produce two pie charts to display the data
1. Determine the mode for the advantages and disadvantages.
1. Compute the percentages for the advantages and disadvantages.
1. Compute the odds of a farmer stating 'high price' as a disadvantage, among all farmers in the study.
1. Compute the odds of a farmer stating 'high price' as a disadvantage, among all farmers who listed a disadvantage.
:::

```{r}
Advantages <- c(15, 27, 16)
namesAdvantages <- c("Water reutilization",
                     "Availabiity",
                     "Sustainability")
Disadvantages <- c(40, 12, 21)
namesDisadvantages <- c("High price",
                        "Growing conductivity",
                        "Lack of proper filtering")

AdTable <- DisadTable <- array( dim = c(3, 2))

AdTable[, 1] <- namesAdvantages
AdTable[, 2] <- Advantages
DisadTable[, 1] <- namesDisadvantages
DisadTable[, 2] <- Disadvantages

if( knitr::is_latex_output() ) {
  
  T1 <-  knitr::kable(pad(AdTable,
                          surroundMaths = TRUE,
                          targetLength = c(0, 2),
                          digits = 0),
                      format = "latex",
                      valign = 't',
                      align = c("r", "c"),
                      linesep = "",
                      col.names = c("Advantage", "No. farmers"),
                      row.names = FALSE,
                      escape = FALSE,
                      booktabs = TRUE) %>%
    row_spec(0, bold = TRUE)
  
  
  
  T2 <-  knitr::kable(pad(DisadTable,
                          surroundMaths = TRUE,
                          targetLength = c(0, 2),
                          digits = 0),
                      format = "latex",
                      valign = 't',
                      align = c("r", "c"),
                      linesep = "",
                      col.names = c("sDisadvantage", "No. farmers"),
                      row.names = FALSE,
                      escape = FALSE,
                      booktabs = TRUE) %>%
    row_spec(0, bold = TRUE)
  
  
  
  out <- knitr::kables( list(T1, T2),
                        format = "latex",
                        label = "ReclaimedWater",
                        caption = "The advantages and disadvantages of using reclaimed water, reported by $231$ farmers.") %>% 
    kable_styling(font_size = 8)
  out2 <- prepareSideBySideTable(out) 
  out2
}
if( knitr::is_html_output() ) {
  
  T1 <- knitr::kable(pad(AdTable,
                          surroundMaths = TRUE,
                          targetLength = c(0, 2),
                          digits = 0),
                      format = "html",
                      valign = 't',
                      align = "c",
                      linesep = "",
                      col.names = c("Advantage", "No. farmers"),
                      row.names = FALSE,
                      escape = FALSE,
                      booktabs = TRUE) %>%
    row_spec(0, bold = TRUE)
  
  
  T2 <-  knitr::kable(pad(DisadTable,
                          surroundMaths = TRUE,
                          targetLength = c(0, 2),
                          digits = 0),
                      format = "html",
                      valign = 't',
                      align = "c",
                      linesep = "",
                      col.names = c("Disadvantage", "No. farmers"),
                      row.names = FALSE,
                      escape = FALSE,
                      booktabs = TRUE) %>%
    row_spec(0, bold = TRUE)
  
  
  
  knitr::kables( list(T1, T2),
                        format = "html",
                        label = "ReclaimedWater",
                        caption = "The advantages and disadvantages of using reclaimed water, reported by $231$ farmers.") 
}
```


::: {.exercise #StudentTransport}
A study of $284$ university students in Joinville, Brazil, tabulated how students get to campus (Table\ \@ref(tab:TransportTable);  @henning2020modelling).

1. What is the mode type of active and motorised transport?
1. What is the mode type of transport overall?
1. Is a median appropriate?
1. Compute the odds that a randomly-chosen student uses motorised transport to get to campus.
   Explain what this means.
1. Compute the odds that a student walks to campus.
   Explain what this means.
1. Construct appropriate plots to display the data.
:::

```{r TransportTable}
TransportTable <- array( dim = c(5, 2) )

rownames(TransportTable) <- c("Bicycle",
                              "Walking",
                              "Car",
                              "Bus",
                              "Other")
colnames(TransportTable) <- c("Number",
                              "Percentage")
TransportTable[, 1] <- c(29, 35, 70, 117, 33)
TransportTable[, 2] <- c(10.2, 12.3, 24.7, 41.2, 11.6)

if( knitr::is_latex_output() ) {
knitr::kable(pad(TransportTable,
                 surroundMaths = TRUE,
                 targetLength = c(3, 4),
                 digits = c(0, 1)),
             format = "latex",
             longtable = FALSE,
             escape = FALSE,
             caption = "Modes of transport for students getting to campus",
             align = "c",
             booktabs = TRUE) %>%
  kable_styling(font_size = 8) %>%
  row_spec(0, bold = TRUE) %>%
  pack_rows("Active", 
            start_row = 1,
            end_row = 2) %>%
  pack_rows("Motorised", 
            start_row = 3,
            end_row = 5)
}
if( knitr::is_html_output() ) {
  
knitr::kable(pad(TransportTable,
                 surroundMaths = TRUE,
                 targetLength = c(3, 4),
                 digits = c(0, 1)),
             format = "html",
             longtable = FALSE,
             escape = FALSE,
             caption = "Modes of transport for students getting to campus",
             align = "c",
             booktabs = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  pack_rows("Active", 
            start_row = 1,
            end_row = 2) %>%
  pack_rows("Motorised", 
            start_row = 3,
            end_row = 5)
}
```






<!-- QUICK REVIEW ANSWERS -->
`r if (knitr::is_html_output()) '<!--'`
::: {.EOCanswerBox data-latex="{iconmonstr-check-mark-14-240.png}"}
**Answers to in-chapter questions:**
- \textbf{\textit{Quick Revision} questions:}
**1.** False.
**2.** True.
**3.** True.
**4.** False. Percentages are proportions multiplied by $100$, so they are similar.
:::
`r if (knitr::is_html_output()) '-->'`

