# (PART) Analysis {-}


# Introducing inference {#IntroducingInference}
\index{Inference}

<!-- Introductions; easier to separate by format -->
```{r, child = if (knitr::is_html_output()){'./introductions/21-Inference-Intro-HTML.Rmd'} else {'./introductions/21-Inference-Intro-LaTeX.Rmd'}}
```


<!-- Define colours as appropriate -->
```{r, child = if (knitr::is_html_output()) {'./children/coloursHTML.Rmd'} else {'./children/coloursLaTeX.Rmd'}}
```


After posing a RQ (Chap.\ \@ref(RQs)), a study is designed (Chaps.\ \@ref(ResearchDesign)--\@ref(Interpretation)) to gather the evidence to answer the RQ (Chap.\ \@ref(CollectingDataProcedures)).
Then the data are classified (Chap.\ \@ref(DescribingVars)) and summarised accordingly (Chaps.\ \@ref(SummariseQualData) to\ \@ref(SummariseComments)) in preparation for answering the RQ.


This Part introduces *analysis*: where the data are used to answer the research question about the population.
Answering the RQ (which is about a *parameter*) is difficult, since we only study one of the countless possible samples, and hence observe only one of the countless possible values for the *statistic*.
This is called *sampling variation* (Chap.\ \@ref(SamplingVariation)).

Analysis provides the tools for learning about a population parameter, based on observing one of the numerous possible values of a sample statistic.
The appropriate type of analysis depends upon the number and types of variables, and the purpose of the RQ (Sect.\ \@ref(TwoPurposesOfRQs)):

* *Confidence intervals* answer estimation RQ, where the interest is in how precisely a *statistic* estimates a *parameter* (Chaps.\ \@ref(CIOneProportion) to\ \@ref(OneMeanConfInterval); \@ref(AnalysisPaired) to \@ref(AnalysisOddsRatio); Sect.\ \@ref(RegressionCI)).\index{Confidence intervals}
* *Hypothesis tests* answer decision-making RQs, where *decisions* are required about a *parameter* based on the value of the *statistic* (Chaps.\ \@ref(TestOneProportion) to\ \@ref(TestOneMean);  \@ref(AnalysisPaired) to \@ref(AnalysisOddsRatio); Sects.\ \@ref(CorrelationTesting) and\ \@ref(RegressionHT).)\index{Hypothesis testing}

Different scenarios require different confidence intervals and hypothesis tests; those discussed in this book are shown in Table\ \@ref(tab:OverviewTable).

(ref:CI-OneProp) Chap.\ \@ref(CIOneProportion)

(ref:HT-OneProp) Chap.\ \@ref(TestOneProportion)

(ref:CI-OneMean) Chap.\ \@ref(OneMeanConfInterval)

(ref:HT-OneMean) Chap.\ \@ref(TestOneMean)

(ref:CI-TwoMeans) Chap.\ \@ref(AnalysisTwoMeans)

(ref:HT-TwoMeans) Chap.\ \@ref(AnalysisTwoMeans)

(ref:CI-Paired) Chap.\ \@ref(AnalysisPaired)

(ref:HT-Paired) Chap.\ \@ref(AnalysisPaired)

(ref:CI-OddsR) Chap.\ \@ref(AnalysisOddsRatio)

(ref:HT-OddsR) Chap.\ \@ref(AnalysisOddsRatio)

(ref:HT-Cor) Sect.\ \@ref(CorrelationTesting)

(ref:HT-Reg) Sect.\ \@ref(RegressionHT)

(ref:CI-Reg) Sect.\ \@ref(RegressionCI)

```{r OverviewTable}
Overview <- array(dim = c(7, 3))
colnames(Overview) <- c("",
                        "confidence intervals)",
                        "hypothesis tests)")

Overview[1, ] <- c("Single proportions",
                   "(ref:CI-OneProp)",
                   "(ref:HT-OneProp)")
Overview[2, ] <- c("Single means",
                   "(ref:CI-OneMean)",
                   "(ref:HT-OneMean)")
Overview[3, ] <- c("Mean differences (paired data)",
                   "(ref:CI-Paired)",
                   "(ref:HT-Paired)" )
Overview[4, ] <- c("Comparing two means",
                   "(ref:CI-TwoMeans)",
                   "(ref:HT-TwoMeans)")
Overview[5, ] <- c("Comparing two odds or proportions", 
                   "(ref:CI-OddsR)",
                   "(ref:HT-OddsR)")
Overview[6, ] <- c("Correlation", 
                   " ",
                   "(ref:HT-Cor)")
Overview[7, ] <- c("Regression", 
                   "(ref:CI-Reg)",
                   "(ref:HT-Reg)")

if( knitr::is_latex_output() ) {
  kable(Overview,
        format = "latex",
        booktabs = TRUE,
        align = c("l", "c","c"),
        longtable = FALSE,
        caption = "Confidence intervals and hypothesis tests for different situations.") %>%
    kable_styling("striped", 
                  font_size = 8,
                  full_width = FALSE) %>%
    row_spec(0, bold = TRUE) %>%
    pack_rows("Descriptive RQs", 
              start_row = 1, 
              end_row = 2,
              bold = FALSE, 
              italic = TRUE) %>%
    pack_rows("Repeated-measures RQs", 
              start_row = 3, 
              end_row = 3,
              bold = FALSE, 
              italic = TRUE) %>%
    pack_rows("Relational RQs", 
              start_row = 4, 
              end_row = 5,
              bold = FALSE, 
              italic = TRUE) %>%
    pack_rows("Correlational RQs", 
              start_row = 6, 
              end_row = 7,
              bold = FALSE, 
              italic = TRUE) %>%
    add_header_above( c(" " = 1,
                        "(for forming" = 1,
                        "(for conducting" = 1),
                      bold = TRUE,
                      line = FALSE) %>%
    add_header_above( c(" " = 1,
                        "Estimation RQs" = 1,
                        "Decision-making RQs" = 1),
                      bold = TRUE,
                      line = FALSE)
}
if( knitr::is_html_output() ) {
  kable(Overview,
        format = "html",
        booktabs = TRUE,
        align = c("l", "c","c"),
        longtable = FALSE,
        caption = "Confidence interrvals and hypothesis tests for different situations.") %>%
    kable_styling("striped", 
                  full_width = FALSE) %>%
    pack_rows("Descriptive RQs", 
              start_row = 1, 
              end_row = 3) %>%
    pack_rows("Repeated-measures RQs", 
              start_row = 3, 
              end_row = 3) %>%
    pack_rows("Relational RQs", 
              start_row = 4, 
              end_row = 5) %>%
    pack_rows("Correlational RQs", 
              start_row = 6, 
              end_row = 7) %>%
    add_header_above( c(" " = 1,
                        "Estimation" = 1,
                        "Decision-making" = 1),
                      bold = TRUE,
                      line = FALSE)
}
```




