# (PART) Analysis {-}


# Introducing inference {#Introducing-Inference}


<!-- Introductions; easier to separate by format -->
```{r, child = if (knitr::is_html_output()){'./introductions/19-Inference-Intro-HTML.Rmd'} else {'./introductions/19-Inference-Intro-LaTeX.Rmd'}}
```


## Introduction

After posing a RQ (Chap. XX), a study is designed (Chap. XX) to gather the evidence to answer the RQ (Chap. XX).
Then the data are described (Chap. XX) and summarised (Chap. XX) in preparation for answering the RQ.

Answering the RQ is difficult, since we only study one randomly-selected sample from all the numerous possible samples.
This is called \emph{sampling variation} (Chap.\ \ref{SamplingVariation}).


This Part introduces *analysis*: where the data are used to answer the research question.

Analysis provides the tools for learning about a population parameter, based on observing one of the numerous possible values of a sample statistic.
The appropriate type of analysis depends upon the number and types of variables, and the RQ (Table\ \@ref(tab:OverviewTable)):

* *Confidence intervals* answer [estimation-type RQ](#TypesOfRQs), where the precision of a *statistic* to estimate a *parameter* is of interest.
* *Hypothesis tests* answer [decision-type RQs](#TypesOfRQs), where *decisions* are required.


(ref:CI-OneProp) Chap.\ \@ref(CIOneProportion)

(ref:HT-OneProp) Chap.\ \@ref(TestOneProportion)

(ref:CI-OneMean) Chap.\ \@ref(OneMeanConfInterval)

(ref:HT-OneMean) Chap.\ \@ref(TestOneMean)

(ref:CI-TwoMeans) Chap.\ \@ref(CITwoMeans)

(ref:HT-TwoMeans) Chap.\ \@ref(TestTwoMeans)

(ref:CI-Paired) Chap.\ \@ref(PairedCI)

(ref:HT-Paired) Chap.\ \@ref(TestPairedMeans)

(ref:CI-OddsR) Chap.\ \@ref(OddsRatiosCI)

(ref:HT-OddsR) Chap.\ \@ref(TestsOddsRatio)

(ref:HT-Cor) Sect.\ \@ref(CorrelationTesting)

(ref:HT-Reg) Sect.\ \@ref(RegressionCI)

(ref:CI-Reg) Sect.\ \@ref(RegressionCI)

```{r OverviewTable}
Overview <- array(dim = c(7, 3))
colnames(Overview) <- c("",
                        "Estimation (CI)",
                        "Decision (Tests)")

Overview[1, ] <- c("Proportions for one sample",
                   "(ref:CI-OneProp)",
                   "(ref:HT-OneProp)")
Overview[2, ] <- c("Means for one sample",
                   "(ref:CI-OneMean)",
                   "(ref:HT-OneMean)")
Overview[3, ] <- c("Mean differences (paired data)",
                   "(ref:CI-Paired)",
                   "(ref:HT-Paired)" )
Overview[4, ] <- c("Comparing two means",
                   "(ref:CI-TwoMeans)",
                   "(ref:HT-TwoMeans)")
Overview[5, ] <- c("Comparing two odds", 
                   "(ref:CI-OddsR)",
                   "(ref:HT-OddsR)")
Overview[6, ] <- c("Correlation", 
                   " ",
                   "(ref:HT-Cor)")
Overview[7, ] <- c("Regression", 
                   "(ref:CI-Reg)",
                   "(ref:HT-Reg)")

if( knitr::is_latex_output() ) {
  kable(Overview,
        format = "latex",
        booktabs = TRUE,
        align = c("l", "c","c"),
        longtable = FALSE,
        caption = "Confidence intervals and hypothesis tests for different situations") %>%
    kable_styling("striped", 
                  font_size = 10,
                  full_width = FALSE) %>%
    row_spec(0, bold = TRUE) %>%
    pack_rows("Descriptive RQs", 
              start_row = 1, 
              end_row = 2,
              bold = FALSE, 
              italic = TRUE) %>%
    pack_rows("RQs comparing quantitative variables: within-individuals", 
              start_row = 3, 
              end_row = 3,
              bold = FALSE, 
              italic = TRUE) %>%
    pack_rows("RQs comparing variables: between-individual", 
              start_row = 4, 
              end_row = 5,
              bold = FALSE, 
              italic = TRUE) %>%
    pack_rows("RQs with a connection between quantitative variable", 
              start_row = 6, 
              end_row = 7,
              bold = FALSE, 
              italic = TRUE)
}
if( knitr::is_html_output() ) {
  kable(Overview,
        format = "html",
        booktabs = TRUE,
        align = c("l", "c","c"),
        longtable = FALSE,
        caption = "Confidence interrvals and hypothesis tests for different situations") %>%
    kable_styling("striped", 
                  full_width = FALSE) %>%
    pack_rows("Descriptive RQs", 
              start_row = 1, 
              end_row = 3) %>%
    pack_rows("RQs comparing quantitative variables: within-individuals", 
              start_row = 3, 
              end_row = 3) %>%
    pack_rows("RQs comparing variables: between-individual", 
              start_row = 4, 
              end_row = 5) %>%
    pack_rows("RQs with a connection between quantitative variable", 
              start_row = 6, 
              end_row = 7)
}
```




