# (PART) Analysis: Hypothesis testing {-}





# Introducing hypothesis tests {#HTIntro}



```{r echo=FALSE, fig.cap="", fig.align="center", fig.width=3, out.width="35%"}
SixSteps(5, "Comparison RQs: Hypothesis tests")
```



We have studied forming *confidence intervals*,
which answer estimation-type RQs,
and indicate the precision with which a statistic estimates a parameter.

Now,
we begin looking at 
[decision-type RQs](#TwoApproachesHT),
which help us make decisions about the value of unknown parameters based on the value of the statistic
(Table \@ref(tab:OverviewTable)).
The [decision-making process](#DecisionMaking)
(Chap. \@ref(MakingDecisions))
we discussed was:

1. **Assumption**:
Make an assumption about the *population*.
	      
2. **Expectation**:
Based on this assumption,
the distribution of the possible values of the sample statistic can be described.

3. **Observation**:
If *sample* information is observed that is:
    - unlikely to happen by chance, it is *contrary to* that assumption about the *population parameter*,
			and the assumption is probably **wrong**.
			There is *evidence* to suggest that the assumption is wrong (but it is not *certainly* wrong).
    - *likely* to happen by chance, it is **consistent with** that assumption about the *population parameter*,
			and the assumption may be **correct**.
			There is no *evidence* to suggest the assumption is wrong (though it may be wrong).

In this Part,
we explore decision-type relational or interventional RQs with a *comparison*.
(Decision-type RQs with a *connection* are discussed in 
Chaps. \@ref(Correlation) and \@ref(Regression).)


(ref:CI-OneProp) Chap. \@ref(CIOneProportion)

(ref:CI-OneMean) Chap. \@ref(OneMeanConfInterval)

(ref:HT-OneMean) Chap. \@ref(TestOneMean)

(ref:CI-TwoMeans) Chap. \@ref(CITwoMeans)

(ref:HT-TwoMeans) Chap. \@ref(TestTwoMeans)

(ref:CI-Paired) Chap. \@ref(PairedCI)

(ref:HT-Paired) Chap. \@ref(TestPairedMeans)

(ref:CI-OddsR) Chap. \@ref(OddsRatiosCI)

(ref:HT-OddsR) Chap. \@ref(TestsOddsRatio)


(ref:HT-Cor) Sect. \@ref(CorrelationTesting) 

(ref:HT-Reg) Sect.  \@ref(RegressionCI)

(ref:CI-Reg) Sect.  \@ref(RegressionCI)

```{r OverviewTable, echo=FALSE}
Overview <- array(dim=c(7, 3))
colnames(Overview) <- c("",
                        "Estimation type (CI)",
                        "Decision type (Tests)")

Overview[1, ] <- c("Proportions for one sample",
                   "(ref:CI-OneProp)",
                   "")
Overview[2, ] <- c("Means for one sample",
                    "(ref:CI-OneMean)",
                   "(ref:HT-OneMean)")
Overview[3, ] <- c("Mean differences (for paired data)",
                   "(ref:CI-Paired)",
                   "(ref:HT-Paired)" )
Overview[4, ] <- c("Means for two samples",
                   "(ref:CI-TwoMeans)",
                   "(ref:HT-TwoMeans)")
Overview[5, ] <- c("Odds for two samples (ORs)", 
                   "(ref:CI-OddsR)",
                   "(ref:HT-OddsR)")
Overview[6, ] <- c("Correlation", 
                   "",
                   "(ref:HT-Cor)")
Overview[7, ] <- c("Regression", 
                   "(ref:CI-Reg)",
                   "(ref:HT-Reg)")
  
if( knitr::is_html_output() ) {
  kable(Overview,
      format="html",
      booktabs=TRUE,
      longtable=FALSE,
      caption="Confidence interrvals and hypothesis tests for different situations") %>%
  kable_styling("striped", full_width = F) %>%
  pack_rows("Descriptive RQs", 1, 3) %>%
  pack_rows("Relational/Interventional RQs with a *Comparison*", 4, 5) %>%
  pack_rows("Relational/Interventional RQs with a *Connection*", 6, 7)
}
if( knitr::is_latex_output() ) {
  kable(Overview,
      format="latex",
      booktabs=TRUE,
      longtable=FALSE,
      align=c("l", "c","c"),
      caption="Confidence interrvals and hypothesis tests for different situations") %>%
  kable_styling("striped", full_width = F) %>%
  pack_rows("Descriptive RQs", 1, 3) %>%
  pack_rows("Relational/Interventional RQs (Comparison)", 4, 5) %>%
  pack_rows("Relational/Interventional RQs (Connection)", 6, 7)
}
```



