# (PART) Analysis: Hypothesis testing {-}


# Introducing hypothesis tests {#HTIntro}


```{r echo=FALSE, fig.cap="", fig.align="center", fig.width=3, out.width="35%"}
SixSteps(5, "Comparison RQs: Hypothesis tests")
```

We have studied how to construct *confidence intervals*, which answer estimation-type RQs, and indicate the precision with which a statistic estimates a parameter.
Now, we begin looking at [decision-type RQs](#TwoApproachesHT), which help us make decisions about the value of unknown parameters based on the value of the statistic (Table \@ref(tab:OverviewTable)).
This is called *hypothesis testing*.


::: {.tipBox .tip data-latex="{iconmonstr-info-6-240.png}"}
The word *hypothesis* means 'a possible explanation'.

**Scientific hypotheses** usually refer to the explanations that scientists or engineers are wishing to demonstrate are true; for example, an engineer may expect that replacing sand with glass in the manufacture of concrete produces desirable characteristics [@devaraj2021exploring].

**Statistical hypotheses** refer to the *null hypothesis* and the *alternative hypothesis* that are necessary for formal statistical hypothesis testing.
These refer to the two possible statistical explanations for the difference between the proposed population parameter and the observed sample statistic.

We are discussing *statistical hypotheses*.
:::


The [decision-making process](#DecisionMaking) (Chap. \@ref(MakingDecisions)) previously discussed was:

1. **Assumption**: Make an assumption about the *population*.
2. **Expectation**: Based on this assumption, the distribution of the possible values of the sample statistic can be described.
3. **Observation**:
   If *sample* information is observed that is:
    - unlikely to happen by chance, it is *contrary to* that assumption about the *population parameter*, and the assumption is probably **wrong**.
			There is *evidence* to suggest that the assumption is wrong (but it is not *certainly* wrong).
    - *likely* to happen by chance, it is **consistent with** that assumption about the *population parameter*, and the assumption may be **correct**.
			There is no *evidence* to suggest the assumption is wrong (though it may be wrong).

In this Part, we explore decision-type relational or interventional RQs with a *comparison*.
(Decision-type RQs with a *connection* are discussed in Chaps. \@ref(Correlation) and \@ref(Regression).)


(ref:CI-OneProp) Chap. \@ref(CIOneProportion)

(ref:HT-OneProp) Chap. \@ref(TestOneProportion)

(ref:CI-OneMean) Chap. \@ref(OneMeanConfInterval)

(ref:HT-OneMean) Chap. \@ref(TestOneMean)

(ref:CI-TwoMeans) Chap. \@ref(CITwoMeans)

(ref:HT-TwoMeans) Chap. \@ref(TestTwoMeans)

(ref:CI-Paired) Chap. \@ref(PairedCI)

(ref:HT-Paired) Chap. \@ref(TestPairedMeans)

(ref:CI-OddsR) Chap. \@ref(OddsRatiosCI)

(ref:HT-OddsR) Chap. \@ref(TestsOddsRatio)


(ref:HT-Cor) Sect. \@ref(CorrelationTesting)

(ref:HT-Reg) Sect. \@ref(RegressionCI)

(ref:CI-Reg) Sect. \@ref(RegressionCI)

```{r OverviewTable, echo=FALSE}
Overview <- array(dim = c(7, 3))
colnames(Overview) <- c("",
                        "Estimation type (CI)",
                        "Decision type (Tests)")

Overview[1, ] <- c("Proportions for one sample",
                   "(ref:CI-OneProp)",
                   "(ref:HT-OneProp)")
Overview[2, ] <- c("Means for one sample",
                    "(ref:CI-OneMean)",
                   "(ref:HT-OneMean)")
Overview[3, ] <- c("Mean differences (for paired data; within-individual)",
                   "(ref:CI-Paired)",
                   "(ref:HT-Paired)" )
Overview[4, ] <- c("Means for two samples (between-individuals)",
                   "(ref:CI-TwoMeans)",
                   "(ref:HT-TwoMeans)")
Overview[5, ] <- c("Odds for two samples, comparing ORs (between-individuals)", 
                   "(ref:CI-OddsR)",
                   "(ref:HT-OddsR)")
Overview[6, ] <- c("Correlation", 
                   " ",
                   "(ref:HT-Cor)")
Overview[7, ] <- c("Regression", 
                   "(ref:CI-Reg)",
                   "(ref:HT-Reg)")
  
if( knitr::is_latex_output() ) {
  kable(Overview,
        format = "latex",
        booktabs = TRUE,
        align = c("l", "c","c"),
        longtable = FALSE,
        caption = "Confidence intervals and hypothesis tests for different situations") %>%
  kable_styling("striped", 
                full_width = F) %>%
  pack_rows("Descriptive RQs", 1, 3) %>%
  pack_rows("Relational/Interventional RQs with a Comparison", 4, 5) %>%
  pack_rows("Relational/Interventional RQs with a Connection", 6, 7)
}
if( knitr::is_html_output(exclude = "epub") ) {
  kable(Overview,
      format = "html",
      booktabs = TRUE,
      align = c("l", "c","c"),
      longtable = FALSE,
      caption = "Confidence interrvals and hypothesis tests for different situations") %>%
    kable_styling("striped", 
		  full_width = FALSE) %>%
    pack_rows("Descriptive RQs", 1, 3) %>%
    pack_rows("Relational/Interventional RQs (Comparison)", 4, 5) %>%
    pack_rows("Relational/Interventional RQs (Connection)", 6, 7)
}
if( knitr::pandoc_to("epub3") ) {
  kable(Overview,
        format = "html",
        booktabs = TRUE,
        align = c("l", "c","c"),
        longtable = FALSE,
        caption = "Confidence interrvals and hypothesis tests for different situations") 
}

```



