# Selecting a hypothesis testing {#SelectTest}



Selecting the correct hypothesis test (or confidence interval) 
can be tricky...
and in this book only a small number of hypothesis tests
were described.
(Literally hundreds of tests exist 
[@kanji2006100].)

For the tests studied in this book,
determining if the response and explanatory *variables*
are qualitative or quantitative is important
(Table \@ref(tab:InferenceTestCI)).
So far,
only situations with a *qualitative* explanatory variable
have been considered.
In the next chapters,
cases where both the response and explanatory variables 
are *quantitative* are studied.







```{r InferenceTestCI, echo=FALSE}
Scenarios <- array( dim=c(4, 4) )
colnames(Scenarios) <- c("Graphical summary",
                         "Numerical summary",
                         "Hypothesis test",
                         "Confidence interval")

Scenarios[1, ] <- c("Histogram; stem-and-leaf plot; dot chart",
                    "Means, medians; Std. dev., IQR; etc.",
                    "One-sample $t$",
                    "CI for one mean")
Scenarios[2, ] <- c("Histogram of differences; case-profile",
                    "Mean, std. dev. etc. of differences",
                    "$t$-test for mean differences",
                    "CI for mean difference")
Scenarios[3, ] <- c("Error bar chart",
                   "Mean and std. error of the difference; mean, std. dev. etc. of each group",
                   "$t$-test comparing the difference between two means",
                   "CI of the difference between means")
Scenarios[4, ] <- c("Side-by-side bar chart; stacked bar chart",
                    "Odds; OR; percentages",
                    "Chi-square test",
                    "CI for OR")
                    
                    
                    
if( knitr::is_latex_output() ) {
  kable(Scenarios,
        booktabs = TRUE,
        longtable = FALSE,
        escape = FALSE,
        caption = "Four different scenarios studied so far",
        format = "latex") %>%
     kable_styling("striped", 
                   full_width = TRUE, 
                   font_size = 9) %>%
     pack_rows("Mean of one sample",                         
                start_row = 1, 
                end_row = 1, 
                bold = FALSE, 
                italic = TRUE) %>%
     pack_rows("Mean of differences (paired data)",          
               start_row = 2, 
               end_row = 2, 
               bold = FALSE, 
               italic = TRUE, 
               hline_before = TRUE) %>%
     pack_rows("Comparing odds/percentages in two groups",   
               start_row = 3, 
               end_row = 3, 
               bold = FALSE, 
               italic = TRUE, 
               hline_before = TRUE) %>%
     pack_rows("Comparing means in two groups",              
               start_row = 4, 
               end_row = 4, 
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE)
}                 
                    
if( knitr::is_html_output() ) {
  out <- kable(Scenarios,
               booktabs = TRUE,
               longtable = FALSE,
               escape = FALSE,
               caption = "Four different scenarios studied so far",
               format = "html")
   if ( knitr::is_html_output(excludes = "epub")) {
      kable_styling(out, 
                    "striped", 
                    full_width = TRUE, 
                    font_size = 9) %>%
      pack_rows("Mean of one sample",                         
                start_row = 1, 
                end_row = 1, 
                bold = FALSE, 
                italic = TRUE) %>%
      pack_rows("Mean of differences (paired data)",          
                start_row = 2, 
                end_row = 2, 
                bold = FALSE, 
                italic = TRUE, 
                hline_before = TRUE) %>%
      pack_rows("Comparing odds/percentages in two groups",   
                start_row = 3, 
                end_row = 3, 
                bold = FALSE, 
                italic = TRUE, 
                hline_before = TRUE) %>%
      pack_rows("Comparing means in two groups",              
                start_row = 4, 
                end_row = 4, 
                bold = FALSE, 
                italic = TRUE, 
                hline_before = TRUE)
  } else {
     out
  }
}

```



```{block2, type="rmdthinkHTML"}
<iframe src="https://h5p.org/h5p/embed/488122" width="1090" height="324" frameborder="0" allowfullscreen="allowfullscreen"></iframe><script src="https://h5p.org/sites/all/modules/h5p/library/js/h5p-resizer.js" charset="UTF-8"></script>
```


 
`r if (knitr::is_html_output()){
  'The following short video may help explain some of these concepts. Note that the test for correlation and regression have not yet been covered in this book (but they will be in the next few chapters).'
}`


<div style="text-align:center;">
```{r, echo=FALSE}
htmltools::tags$video(src ="./videos/SelectTest.mp4", 
                      width="550", 
                      autoplay="false", 
                      controls="controls", 
                      loop="loop", 
                      style="padding:5px; border: 2px solid gray;")
```


