# Selecting an analysis {#SelectTest}

Selecting the correct confidence interval or hypothesis test can be tricky... and in this book only a small number of possible scenarios are described.
For the situations studied in this book, determining if the variables are qualitative or quantitative is important (Table\ \@ref(tab:InferenceTestCI)).

So far, only situations with a *qualitative* explanatory variable have been considered.
In the next three chapters, cases where both the response and explanatory variables are *quantitative* are studied.
*Appendix\ \@ref(StatisticsAndParameters)* may also prove useful.





<!-- ```{r InferenceTestCI2} -->
<!-- Scenarios <- array( dim = c(6, 4) ) -->
<!-- colnames(Scenarios) <- c("Graphical summary", -->
<!--                          "Numerical summary", -->
<!--                          "Hypothesis test", -->
<!--                          "Confidence interval") -->

<!-- if( knitr::is_latex_output() ) { -->
<!--   Scenarios[1, ] <- c("\\stackunder{Bar chart;}{\\stackunder{pie chart;}{dotchart}}", -->
<!--                       "\\stackunder{Counts;}{\\stackunder{percentages;}{odds}}", -->
<!--                       "One-sample $z$-test", -->
<!--                       "For one proportion") -->
<!--   Scenarios[2, ] <- c("\\stackunder{Histogram;}{\\stackunder{stemplot;}{dot chart}}", -->
<!--                       "\\stackunder{Means, medians;}{\\stackunder{Std. dev., IQR;}{outliers, etc.}}", -->
<!--                       "One-sample $t$-test", -->
<!--                       "For one mean") -->
<!--   Scenarios[3, ] <- c("\\stackunder{Histogram of \\emph{diffs.};}{case-profile}", -->
<!--                       "\\stackunder{Mean, median of diffs.;}{\\stackunder{Std. dev., IQR of diffs.;}{Outliers, etc.}}", -->
<!--                       "\\stackunder{$t$-test for}{mean diffs.}", -->
<!--                       "For mean diff") -->
<!--   Scenarios[4, ] <- c("\\stackunder{Boxplot;}{error bar chart}", -->
<!--                       "\\stackunder{Diff. between means;}{\\stackunder{SE of difference;}{Summary of both groups}}", -->
<!--                       "\\stackunder{$t$-test for the diff.}{between two means}", -->
<!--                       "\\stackunder{For diff. between}{two means}") -->
<!--   Scenarios[5, ] <- c("\\stackunder{Side-by-side bar chart;}{stacked bar chart}", -->
<!--                       "\\stackunder{Odds;}{\\stackunder{OR;}{percentages}}", -->
<!--                       "Chi-square test", -->
<!--                       "For ORs") -->
<!--   Scenarios[6, ] <- c("Scatterplot", -->
<!--                       "Correlation", -->
<!--                       "\\stackunder{Correlation;}{\\stackunder{$t$-test for regression}{parameters}}", -->
<!--                       "For regression parameters") -->

<!--   kable(Scenarios, -->
<!--         booktabs = TRUE, -->
<!--         longtable = FALSE, -->
<!--         escape = FALSE, -->
<!--         caption = "Five scenarios studied so far, plus one to follow", -->
<!--         format = "latex") %>% -->
<!--      kable_styling("striped",  -->
<!--                    full_width = TRUE,  -->
<!--                    font_size = 9) %>% -->
<!--      pack_rows("Proportion in one sample",                          -->
<!--                 start_row = 1,  -->
<!--                 end_row = 1,  -->
<!--                 bold = FALSE,  -->
<!--                 italic = TRUE) %>% -->
<!--      pack_rows("Mean of one sample",                          -->
<!--                 start_row = 2,  -->
<!--                 end_row = 2,  -->
<!--                 bold = FALSE,  -->
<!--                 italic = TRUE, -->
<!--                 hline_before = TRUE) %>% -->
<!--      pack_rows("Comparing quantitative variables: within-individuals (paired data)",           -->
<!--                start_row = 3,  -->
<!--                end_row = 3,  -->
<!--                bold = FALSE,  -->
<!--                italic = TRUE,  -->
<!--                hline_before = TRUE) %>% -->
<!--      pack_rows("RQs comparing quantitative variables: between-individuals",    -->
<!--                start_row = 4,  -->
<!--                end_row = 4,  -->
<!--                bold = FALSE,  -->
<!--                italic = TRUE,  -->
<!--                hline_before = TRUE) %>% -->
<!--      pack_rows("RQs comparing qualitative variables: between-individuals",               -->
<!--                start_row = 5,  -->
<!--                end_row = 5,  -->
<!--                bold = FALSE, -->
<!--                italic = TRUE, -->
<!--                hline_before = TRUE) %>% -->
<!--     pack_rows("RQs with a connection between quantitative variable", -->
<!--                start_row = 6,  -->
<!--                end_row = 6,  -->
<!--                bold = FALSE, -->
<!--                italic = TRUE, -->
<!--                hline_before = TRUE) %>% -->
<!--      row_spec(0, bold = TRUE) -->

<!-- }                  -->

<!-- if( knitr::is_html_output() ) { -->

<!--   Scenarios[1, ] <- c("[Bar charts; pie chart](#GraphsOneQual)", -->
<!--                       "[Counts; percentages; odds](#NumericalQual)", -->
<!--                       "[One-sample $z$](#TestOneProportion)", -->
<!--                       "[CI for one mean](#CIOneProportion)") -->
<!--   Scenarios[2, ] <- c("[Histogram; stemplot; dot chart](#GraphsOneQuant)", -->
<!--                       "[Means, medians; Std. dev., IQR; etc.](#NumericalQuant)", -->
<!--                       "[One-sample $t$](#TestOneMean)", -->
<!--                       "[CI for one mean](#OneMeanConfInterval)") -->
<!--   Scenarios[3, ] <- c("[Histogram of *differences*](#HistoDiffPlot); [case-profile](#CaseProfilePlot)", -->
<!--                       "[Mean, std. dev. etc. of *differences*](#NumericalQuant)", -->
<!--                       "[$t$-test for mean differences](#TestPairedMeans)", -->
<!--                       "[CI for mean difference](#PairedCI)") -->
<!--   Scenarios[4, ] <- c("[Error bar chart](#ErrorBarCharts)", -->
<!--                      "[Mean and std. error of the difference; mean, std. dev. etc. of *each* group](#NumericalQuant)", -->
<!--                      "[$t$-test for the difference between two means](#TestTwoMeans)", -->
<!--                      "[CI of the difference between two means](#CITwoMeans)") -->
<!--   Scenarios[5, ] <- c("[Side-by-side bar chart; stacked bar chart](#TwoQualVars)", -->
<!--                       "[Odds; OR; percentages](#NumericalQual)", -->
<!--                       "[Chi-square test](#TestsOddsRatio)", -->
<!--                       "[CI for ORs](#OddsRatiosCI)") -->
<!--   Scenarios[6, ] <- c("[Scatterplot](#TwoQualVars)", -->
<!--                       "[Correlation](#NumericalQual)", -->
<!--                       "[Correlation; regression](#TestsOddsRatio)", -->
<!--                       "[CI for regression parameters](#OddsRatiosCI)") -->


<!--   kable(Scenarios, -->
<!--                booktabs = TRUE, -->
<!--                longtable = FALSE, -->
<!--                escape = FALSE, -->
<!--                caption = "Five different scenarios studied so far, plus one to follow", -->
<!--                format = "html") %>% -->
<!--       kable_styling("striped",  -->
<!--                     full_width = TRUE) %>% -->
<!--       column_spec(column = 1, width = "25mm") %>% -->
<!--       column_spec(column = 2, width = "25mm") %>% -->
<!--       column_spec(column = 3, width = "25mm") %>% -->
<!--       column_spec(column = 4, width = "25mm") %>% -->
<!--      pack_rows("Proportion in one sample",                          -->
<!--                 start_row = 1,  -->
<!--                 end_row = 1,  -->
<!--                 bold = FALSE,  -->
<!--                 italic = TRUE) %>% -->
<!--       pack_rows("Mean of one sample",                          -->
<!--                 start_row = 2,  -->
<!--                 end_row = 2,  -->
<!--                 bold = FALSE,  -->
<!--                 italic = TRUE) %>% -->
<!--      pack_rows("Comparing quantitative variables: within-individuals (paired data)",           -->
<!--                start_row = 3,  -->
<!--                end_row = 3,  -->
<!--                bold = FALSE,  -->
<!--                italic = TRUE,  -->
<!--                hline_before = TRUE) %>% -->
<!--      pack_rows("RQs comparing quantitative variables: between-individuals",    -->
<!--                start_row = 4,  -->
<!--                end_row = 4,  -->
<!--                bold = FALSE,  -->
<!--                italic = TRUE,  -->
<!--                hline_before = TRUE) %>% -->
<!--      pack_rows("RQs comparing qualitative variables: between-individuals",               -->
<!--                start_row = 5,  -->
<!--                end_row = 5,  -->
<!--                bold = FALSE, -->
<!--                italic = TRUE, -->
<!--                hline_before = TRUE) -->
<!-- } -->

<!-- ``` -->






```{r InferenceTestCI}
Scenarios <- array( dim = c(6, 5) )
colnames(Scenarios) <- c("Scenario",
                         "Graphical",
                         "Numerical",
                         "\\stackunder{Hypothesis}{test}",
                         "\\stackunder{Confidence}{interval}")


if( knitr::is_latex_output() ) {
  Scenarios[1, ] <- c("\\stackunder{Proportion in}{one sample}",
                      "\\stackunder{Bar chart;}{\\stackunder{pie chart;}{dotchart}}",
                      "\\stackunder{Counts;}{\\stackunder{percentages;}{odds}}",
                      "\\stackunder{One-sample}{$z$-test}",
                      "\\stackunder{For one}{proportion}")
  Scenarios[2, ] <- c("\\stackunder{Means in}{one sample}",
                      "\\stackunder{Histogram;}{\\stackunder{stemplot;}{dot chart}}",
                      "\\stackunder{Means, medians;}{\\stackunder{Std. dev., IQR;}{outliers, etc.}}",
                      "\\stackunder{One-sample}{$t$-test}",
                      "For one mean")
  Scenarios[3, ] <- c("\\stackunder{Comparing quant.}{\\stackunder{vars: within}{-individuals (paired)}}",
                      "\\stackunder{Histogram of \\emph{diffs.};}{case-profile}",
                      "\\stackunder{Of the}{\\stackunder{difference, and}{for each group}}",
                      "\\stackunder{$t$-test for}{mean diffs.}",
                      "For mean diff.")
  Scenarios[4, ] <- c("\\stackunder{Comparing quant.}{\\stackunder{vars: between}{-individuals}}",
                      "\\stackunder{Boxplot;}{error bar chart}",
                      "\\stackunder{For each group;}{\\stackunder{difference between}{\\stackunder{means; std.\\ error of}{of diff. between means}}}",
                      "\\stackunder{$t$-test for the}{\\stackunder{diff. between}{two means}}", 
                      "\\stackunder{For diff.}{\\stackunder{between two}{means}}")
  Scenarios[5, ] <- c("\\stackunder{Comparing qual.}{\\stackunder{vars: between}{-individuals}}",
                      "\\stackunder{Side-by-side}{\\stackunder{bar chart;}{stacked bar chart}}",
                      "\\stackunder{Odds;}{\\stackunder{OR;}{percentages}}",
                      "Chi-square test",
                      "For ORs")
  Scenarios[6, ] <- c("\\stackunder{Connection}{\\stackunder{between}{quant. vars}}",
                      "Scatterplot",
                      "Correlation",
                      "\\stackunder{Correlation;}{\\stackunder{$t$-test for}{\\stackunder{regression}{parameters}}}",
                      "\\stackunder{For regression}{parameters}")

  kable(Scenarios,
        booktabs = TRUE,
        longtable = FALSE,
        escape = FALSE,
        linesep = "\\addlinespace\\addlinespace",
        # align = c("p{20mm}", "p{25mm}", "p{25mm}", "p{25mm}", "p{20mm}"),
        caption = "Five scenarios studied so far, plus one to follow",
        format = "latex") %>%
    kable_styling(font_size = 9) %>%
    row_spec(0, bold = TRUE) %>%
    column_spec(1, italic = TRUE) %>%
    add_header_above( c(" ", 
                        "Summaries" = 2,
                        "Analysis" = 2),
                      bold = TRUE)
}                 
if( knitr::is_html_output() ) {
  Scenarios[1, ] <- c("Proportion in one sample",
                      "Bar chart; pie chart; dotchart",
                      "Counts; percentages; odds",
                      "One-sample $z$-test",
                      "For one proportion")
  Scenarios[2, ] <- c("Means in one sample",
                      "Histogram; stemplot; dot chart",
                      "Means, medians; Std. dev., IQR; outliers, etc.",
                      "One-sample $t$-test",
                      "For one mean")
  Scenarios[3, ] <- c("Comparing quant. vars: within-individuals (paired)",
                      "Histogram of diffs.; case-profile",
                      "Of the difference, and each group",
                      "$t$-test for mean diffs.",
                      "For mean diff")
  Scenarios[4, ] <- c("Comparing quant. vars: between-individuals",
                      "Boxplot; error bar chart",
                      "Diff. between means; std. error of diff.; For each group",
                      "$t$-test for the diff. between two means",
                      "For diff. between two means")
  Scenarios[5, ] <- c("Comparing qual. vars: between-individuals",
                      "Side-by-side bar chart; stacked bar chart",
                      "Odds; OR; percentages",
                      "Chi-square test",
                      "For ORs")
  Scenarios[6, ] <- c("Connection between quant. vars",
                      "Scatterplot",
                      "Correlation",
                      "Correlation; $t$-test for regression parameters",
                      "For regression parameters")

  kable(Scenarios,
        booktabs = TRUE,
        longtable = FALSE,
        escape = FALSE,
        linesep = "\\addlinespace\\addlinespace",
        # align = c("p{20mm}", "p{25mm}", "p{25mm}", "p{25mm}", "p{20mm}"),
        caption = "Five scenarios studied so far, plus one to follow",
        format = "latex") %>%
    kable_styling(font_size = 9) %>%
    row_spec(0, bold = TRUE) %>%
    column_spec(1, italic = TRUE) %>%
    add_header_above( c(" ", 
                        "Summaries" = 2,
                        "Analysis" = 2),
                      bold = TRUE)
}                 
  
```

::: {.thinkBox .think data-latex="{iconmonstr-light-bulb-2-240.png}"}
1. Suppose researchers compare the average number of hours of exercise per week for office workers, both in summer and in winter, to see if the averages are different.  
   What would be a suitable analysis?
`r if( knitr::is_html_output() ) {
	 mcq( c(answer = "A paired samples t-test (for a mean difference)",
 	        "A chi-squared test (to compare two proportions)",
  	      "A two-sample t-test (to compare the means of the two groups)"))}`   


2. Suppose we wish to estimate difference between the mean number of hours of sunlight exposure per day for female and male teachers.  
   What would be a suitable analysis?
`r if( knitr::is_html_output() ) {
	 mcq( c("A paired samples CI (for a mean difference)",
 	        "A CI for the odds ratio (to compare two odds)",
  	      answer = "A CI for the difference between two means"))}`   

3. Suppose researchers wish to compare the proportion of trees with koalas in them, comparing trees more than $10$\ m tall with trees $10$\ m or shorter.  
   What would be a suitable analysis?
`r if( knitr::is_html_output() ) {
	 mcq( c("A paired samples t-test (for a mean difference)",
 	        answer = "A chi-squared test (to compare two proportions)",
  	      "A two-sample t-test (to compare the means of the two groups)"))}`   

4. Suppose researchers are wanting to compare the number of hours spend on social media for people aged over $30$, to people aged $30$ and under.  
   What would be a suitable analysis?
`r if( knitr::is_html_output() ) {
	 mcq( c("A paired samples t-test (for a mean difference)",
 	        "A chi-squared test (to compare two proportions)",
  	      answer = "A two-sample t-test (to compare the means of the two groups)"))}`   

`r if (!knitr::is_html_output()) '<!--'`
`r webexercises::hide()`
To select the correct test, it is important to know how many 
`r if( knitr::is_html_output() ) { mcq( c("observations", answer = "variables"))}`
are measured, observed, or recorded on each unit of 
`r if( knitr::is_html_output() ) { mcq( c("observation", answer = "analysis"))}`, and what type they are.  
If one quantitative variable is recorded, we can conduct a test about the 
`r if( knitr::is_html_output() ) { mcq( c(answer = "mean", "proportions"))}`.

If two variables are recorded, there are a lot of possible options.

If **both** variables are qualitative, we could use a
`r if( knitr::is_html_output() ) { mcq( c("t-test", answer = "chi-squared test"))}`
to compare the odds (or the proportions) in the two groups.

If one variable is qualitative and one is quantitative, we could use a
`r if( knitr::is_html_output() ) { mcq( c(answer = "t-test", "chi-squared test"))}` to compare the 
`r if( knitr::is_html_output() ) { mcq( c(answer = "means", "odds"))}` in both groups.

If the **change** in the value of a quantitative variable is of interest, we have **paired** data so we could use a $t$-test, based on the 
`r if( knitr::is_html_output() ) { mcq( c( "means", answer = "mean difference", "odds"))}`.
`r if (!knitr::is_html_output()) '-->'`
`r webexercises::unhide()`
:::



 
`r if (knitr::is_html_output()){
  'The following short video may help explain some of these concepts. Note that the test for correlation and regression have not yet been covered in this book (but they will be in the next few chapters).'
}`


<div style="text-align:center;">
```{r}
htmltools::tags$video(src = "./videos/SelectTest.mp4", 
                      width = "550", 
                      controls = "controls", 
                      loop = "loop", 
                      style = "padding:5px; border: 2px solid gray;")
```
</div>


