# Selecting an analysis {#SelectTest}
\index{Hypothesis testing!selecting}\index{Confidence intervals!selecting}


<!-- Define colours as appropriate -->
```{r, child = if (knitr::is_html_output()) {'./children/coloursHTML.Rmd'} else {'./children/coloursLaTeX.Rmd'}}
```


Selecting the correct confidence interval or hypothesis test can be tricky.
This book only describes a small number of possible scenarios.
For the situations studied in this book, determining if the variables are qualitative or quantitative is important (Table\ \@ref(tab:InferenceTestCI2)).
To compare more than two means, a situation not covered in this book, *analysis of variance*\index{Analysis of variance} (ANOVA) is used.

So far,  only descriptive and relational RQs have been studied.
The next two chapters consider *correlational RQs* (Sect.\ \@ref(RQsCorrelational)).
*Appendix\ \@ref(StatisticsAndParameters)* may also prove useful.





```{r InferenceTestCI2}
Scenarios <- array( dim = c(6, 4) )
colnames(Scenarios) <- c("Graphical display",
                         "Numerical summary",
                         "Hypothesis test",
                         "Confidence interval")

if( knitr::is_latex_output() ) {
  Scenarios[1, ] <- c("\\stackunder{Bar chart}{\\stackunder{Pie chart}{Dot chart}}",
                      "\\stackunder{Counts}{\\stackunder{Percentages}{Odds}}",
                      "One-sample $z$-test",
                      "For one proportion")
  Scenarios[2, ] <- c("\\stackunder{Histogram}{\\stackunder{Stemplot}{Dot chart}}",
                      "\\stackunder{Means, medians}{\\stackunder{Std. deviation, IQR}{Outliers}}",
                      "One-sample $t$-test",
                      "For one mean")
  Scenarios[3, ] <- c("\\stackunder{Histogram of}{\\stackunder{\\quad differences}{Case-profile}}",
                      "\\stackunder{Mean, median of diffs.}{\\stackunder{Std. dev., IQR of diffs.}{Outliers, etc.}}",
                      "\\stackunder{$t$-test for mean}{\\quad differences}",
                      "For mean difference")
  #"\\stackunder{For mean}{\\enskip difference}")
  Scenarios[4, ] <- c("\\stackunder{Boxplot}{\\stackunder{Dot chart}{Error bar chart}}",
                      "\\stackunder{Diff. between means}{\\stackunder{Std. error of difference}{Summary of both groups}}",
                      "\\stackunder{$t$-test for difference}{\\quad between two means}",
                      "\\stackunder{For difference between}{\\quad two means}")
  Scenarios[5, ] <- c("\\stackunder{Side-by-side bar}{\\stackunder{Stacked bar}{Dot chart}}",
                      "\\stackunder{Odds}{\\stackunder{Odds ratio}{Percentages}}",
                      "$\\chi^2$-test for OR",
                      "For odds ratios")
  Scenarios[6, ] <- c("Scatterplot",
                      "Correlation",
                      "\\stackunder{Correlation}{\\stackunder{$t$-test for regression}{\\quad parameters}}",
                      "\\stackunder{\\null}{\\stackunder{For regression}{\\quad parameters}}")
  
  kable(Scenarios,
        booktabs = TRUE,
        escape = FALSE,
        linesep = "\\addlinespace\\addlinespace",
        caption = "Five scenarios studied so far, plus one to follow",
        format = "latex") %>%
    row_spec(0, bold = TRUE) %>%
    column_spec(column = 1, width = "28mm") %>%
    column_spec(column = 2, width = "35mm") %>%
    column_spec(column = 3, width = "29mm") %>%
    column_spec(column = 4, width = "30mm") %>%
    kable_styling("striped",
                  font_size = 8)  %>%
    pack_rows("Descriptive RQ: Proportion in one sample (i.e., one qualitative variable)",
              start_row = 1,
              end_row = 1,
              bold = FALSE,
              italic = TRUE) %>%
    pack_rows("Descriptive RQ: Mean of one sample (i.e., one quantitative variable)",
              start_row = 2,
              end_row = 2,
              bold = FALSE,
              italic = TRUE,
              hline_before = TRUE) %>%
    pack_rows("Repeated-measures RQ: paired quantitative data",
              start_row = 3,
              end_row = 3,
              bold = FALSE,
              italic = TRUE,
              hline_before = TRUE) %>%
    pack_rows("Relational RQs: comparing quantitative variables",
              start_row = 4,
              end_row = 4,
              bold = FALSE,
              italic = TRUE,
              hline_before = TRUE) %>%
    pack_rows("Relational RQs: comparing qualitative variables",
              start_row = 5,
              end_row = 5,
              bold = FALSE,
              italic = TRUE,
              hline_before = TRUE) %>%
    pack_rows("Correlational RQs",
              start_row = 6,
              end_row = 6,
              bold = FALSE,
              italic = TRUE,
              hline_before = TRUE) %>%
    add_header_above( c("Summarise" = 2,
                        "Analyse" = 2),
                      line = TRUE,
                      bold = TRUE)
  
}

if( knitr::is_html_output() ) {

  Scenarios[1, ] <- c("[Bar charts; pie chart](#GraphsOneQual)",
                      "[Counts; percentages; odds](#NumericalQual)",
                      "[One-sample $z$](#TestOneProportion)",
                      "[CI for one mean](#CIOneProportion)")
  Scenarios[2, ] <- c("[Histogram; stemplot; dot chart](#GraphsOneQuant)",
                      "[Means, medians; Std. dev., IQR; outliers](#NumericalQuant)",
                      "[One-sample $t$](#TestOneMean)",
                      "[CI for one mean](#OneMeanConfInterval)")
  Scenarios[3, ] <- c("[Histogram of *differences*](#HistoDiffPlot); [case-profile](#CaseProfilePlot)",
                      "[Mean, std. dev. etc. of *differences*](#NumericalQuant)",
                      "[$t$-test for mean differences](#TestPairedMeans)",
                      "[CI for mean difference](#PairedCI)")
  Scenarios[4, ] <- c("[Error bar chart](#ErrorBarCharts)",
                     "[Mean and std. error of the difference; mean, std. dev. etc. of *each* group](#NumericalQuant)",
                     "[$t$-test for the difference between two means](#TestTwoMeans)",
                     "[CI of the difference between two means](#CITwoMeans)")
  Scenarios[5, ] <- c("[Side-by-side bar chart; stacked bar chart](#TwoQualVars)",
                      "[Odds; OR; percentages](#NumericalQual)",
                      "[$\\chi^2$-test for OR](#TestsOddsRatio)",
                      "[CI for odds ratios](#OddsRatiosCI)")
  Scenarios[6, ] <- c("[Scatterplot](#TwoQualVars)",
                      "[Correlation](#NumericalQual)",
                      "[Correlation; regression](#TestsOddsRatio)",
                      "[CI for regression parameters](#OddsRatiosCI)")


  kable(Scenarios,
               booktabs = TRUE,
               longtable = FALSE,
               escape = FALSE,
               caption = "Five different scenarios studied so far, plus one to follow",
               format = "html") %>%
      kable_styling("striped",
                    full_width = TRUE) %>%
      column_spec(column = 1, width = "25mm") %>%
      column_spec(column = 2, width = "25mm") %>%
      column_spec(column = 3, width = "25mm") %>%
      column_spec(column = 4, width = "25mm") %>%
     pack_rows("Descriptive RQ: Proportion in one sample (i.e., one qualitative variable)",
                start_row = 1,
                end_row = 1,
                bold = FALSE,
                italic = TRUE) %>%
     pack_rows("Descriptive RQ: Mean of one sample (i.e., one quantitative variable)",
                start_row = 2,
                end_row = 2,
                bold = FALSE,
                italic = TRUE,
                hline_before = TRUE) %>%
     pack_rows("Repeated-measures RQ: paired quantitative data",
               start_row = 3,
               end_row = 3,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
     pack_rows("Relational RQs: comparing quantitative variables",
               start_row = 4,
               end_row = 4,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
     pack_rows("Relational RQs: comparing qualitative variables",
               start_row = 5,
               end_row = 5,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
    pack_rows("Correlational RQs",
               start_row = 6,
               end_row = 6,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
    add_header_above( c("Summaries" = 2,
                        "Analysis" = 2),
                      line = TRUE,
                      bold = TRUE)
}

```


\pagebreak

## Exercises {#SelectTestExercises}

Answers to odd-numbered exercises are available in App.\ \@ref(Answers).


:::{.exercise #Method1}
Suppose researchers compare the average number of hours of exercise per week for office workers, both in summer and in winter, to see if the averages are different.  

What methods would be a suitable for creating a summary and performing analyses?
:::

:::{.exercise #Method2}
Suppose we wish to *estimate* the difference between the mean number of hours of sunlight exposure per day for female and male teachers.  

What methods would be a suitable for creating a summary and performing analyses?
:::

:::{.exercise #Method3}
Suppose researchers wish to compare the proportion of trees with koalas in them, comparing trees more than $10$\ m tall with trees $10$\ m or shorter.  

What methods would be a suitable for creating a summary and performing analyses?
:::

:::{.exercise #Method4}
Suppose researchers are wanting to *estimate* the difference between the mean number of hours spent on social media for people aged over $30$, to people aged $30$ and under.  

What methods would be a suitable for creating a summary and performing analyses?
:::

 
`r if (knitr::is_html_output()){
  'The following short video may help explain some of these concepts. Note that the test for correlation and regression have not yet been covered in this book (but they will be in the next few chapters).'
}`


<div style="text-align:center;">
```{r}
htmltools::tags$video(src = "./videos/SelectTest.mp4", 
                      width = "550", 
                      controls = "controls", 
                      loop = "loop", 
                      style = "padding:5px; border: 2px solid gray;")
```
</div>


