# Selecting an analysis {#SelectTest}
\index{Hypothesis testing!selecting}\index{Confidence intervals!selecting}

Selecting the correct confidence interval or hypothesis test can be tricky.
This book only describes a small number of possible scenarios.
For the situations studied in this book, determining if the variables are qualitative or quantitative is important (Table\ \@ref(tab:InferenceTestCI2)).

So far,  only descriptive and relational RQs have been studied.
The next two chapters consider *correlational RQs* (Sect.\ \@ref(RQsCorrelational)).
*Appendix\ \@ref(StatisticsAndParameters)* may also prove useful.





```{r InferenceTestCI2}
Scenarios <- array( dim = c(6, 4) )
colnames(Scenarios) <- c("Graphical",
                         "Numerical",
                         "Hypothesis\\hspace{1cm} test",
                         "Confidence interval")

if( knitr::is_latex_output() ) {
  Scenarios[1, ] <- c("\\stackunder{Bar chart}{\\stackunder{Pie chart}{Dotchart}}",
                      "\\stackunder{Counts}{\\stackunder{Percentages}{Odds}}",
                      "One-sample $z$-test",
                      "For one proportion")
  Scenarios[2, ] <- c("\\stackunder{Histogram}{\\stackunder{Stemplot}{Dot chart}}",
                      "\\stackunder{Means, medians}{\\stackunder{Std. dev., IQR}{Outliers, etc.}}",
                      "One-sample $t$-test",
                      "For one mean")
  Scenarios[3, ] <- c("\\stackunder{Histogram of}{\\stackunder{\\enskip differences}{Case-profile}}",
                      "\\stackunder{Mean, median of diffs.}{\\stackunder{Std. dev., IQR of diffs.}{Outliers, etc.}}",
                      "\\stackunder{$t$-test for mean}{\\enskip differences}",
                      "\\stackunder{For mean}{\\enskip difference}")
  Scenarios[4, ] <- c("\\stackunder{Boxplot}{Error bar chart}",
                      "\\stackunder{Diff. between means}{\\stackunder{Std. err. of difference}{Summary of both groups}}",
                      "\\stackunder{$t$-test for the}{\\stackunder{\\enskip difference between}{\\enskip two means}}",
                      "\\stackunder{For diff. between}{\\enskip two means}")
  Scenarios[5, ] <- c("\\stackunder{Side-by-side bar}{Stacked bar}",
                      "\\stackunder{Odds}{\\stackunder{Odds ratio}{Percentages}}",
                      "Chi-square test",
                      "For ORs")
  Scenarios[6, ] <- c("Scatterplot",
                      "Correlation",
                      "\\stackunder{Correlation}{\\stackunder{$t$-test for regression}{\\enskip parameters}}",
                      "\\stackunder{For regression}{\\enskip parameters}")

  kable(Scenarios,
        booktabs = TRUE,
        escape = FALSE,
        caption = "Five scenarios studied so far, plus one to follow",
        format = "latex") %>%
    row_spec(0, bold = TRUE) %>%
    column_spec(column = 1, width = "24mm") %>%
    column_spec(column = 2, width = "30mm") %>%
    column_spec(column = 3, width = "24mm") %>%
    column_spec(column = 4, width = "25mm") %>%
    kable_styling("striped",
                   font_size = 8)  %>%
     pack_rows("Descriptive RQ: Proportion in one sample",
                start_row = 1,
                end_row = 1,
                bold = FALSE,
                italic = TRUE) %>%
     pack_rows("Descriptive RQ: Mean of one sample",
                start_row = 2,
                end_row = 2,
                bold = FALSE,
                italic = TRUE,
                hline_before = TRUE) %>%
     pack_rows("Repeated-measures RQ: paired quantitative data",
               start_row = 3,
               end_row = 3,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
     pack_rows("Relational RQs: comparing quantitative variables",
               start_row = 4,
               end_row = 4,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
     pack_rows("Relational RQs: comparing qualitative variables",
               start_row = 5,
               end_row = 5,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
    pack_rows("Correlational RQs",
               start_row = 6,
               end_row = 6,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
    add_header_above( c("Summaries" = 2,
                        "Analysis" = 2),
                      line = TRUE,
                      bold = TRUE)
    
}

if( knitr::is_html_output() ) {

  Scenarios[1, ] <- c("[Bar charts; pie chart](#GraphsOneQual)",
                      "[Counts; percentages; odds](#NumericalQual)",
                      "[One-sample $z$](#TestOneProportion)",
                      "[CI for one mean](#CIOneProportion)")
  Scenarios[2, ] <- c("[Histogram; stemplot; dot chart](#GraphsOneQuant)",
                      "[Means, medians; Std. dev., IQR; etc.](#NumericalQuant)",
                      "[One-sample $t$](#TestOneMean)",
                      "[CI for one mean](#OneMeanConfInterval)")
  Scenarios[3, ] <- c("[Histogram of *differences*](#HistoDiffPlot); [case-profile](#CaseProfilePlot)",
                      "[Mean, std. dev. etc. of *differences*](#NumericalQuant)",
                      "[$t$-test for mean differences](#TestPairedMeans)",
                      "[CI for mean difference](#PairedCI)")
  Scenarios[4, ] <- c("[Error bar chart](#ErrorBarCharts)",
                     "[Mean and std. error of the difference; mean, std. dev. etc. of *each* group](#NumericalQuant)",
                     "[$t$-test for the difference between two means](#TestTwoMeans)",
                     "[CI of the difference between two means](#CITwoMeans)")
  Scenarios[5, ] <- c("[Side-by-side bar chart; stacked bar chart](#TwoQualVars)",
                      "[Odds; OR; percentages](#NumericalQual)",
                      "[Chi-square test](#TestsOddsRatio)",
                      "[CI for ORs](#OddsRatiosCI)")
  Scenarios[6, ] <- c("[Scatterplot](#TwoQualVars)",
                      "[Correlation](#NumericalQual)",
                      "[Correlation; regression](#TestsOddsRatio)",
                      "[CI for regression parameters](#OddsRatiosCI)")


  kable(Scenarios,
               booktabs = TRUE,
               longtable = FALSE,
               escape = FALSE,
               caption = "Five different scenarios studied so far, plus one to follow",
               format = "html") %>%
      kable_styling("striped",
                    full_width = TRUE) %>%
      column_spec(column = 1, width = "25mm") %>%
      column_spec(column = 2, width = "25mm") %>%
      column_spec(column = 3, width = "25mm") %>%
      column_spec(column = 4, width = "25mm") %>%
     pack_rows("Descriptive RQ: Proportion in one sample",
                start_row = 1,
                end_row = 1,
                bold = FALSE,
                italic = TRUE) %>%
     pack_rows("Descriptive RQ: Mean of one sample",
                start_row = 2,
                end_row = 2,
                bold = FALSE,
                italic = TRUE,
                hline_before = TRUE) %>%
     pack_rows("Repeated-measures RQ: paired quantitative data",
               start_row = 3,
               end_row = 3,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
     pack_rows("Relational RQs: comparing quantitative variables",
               start_row = 4,
               end_row = 4,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
     pack_rows("Relational RQs: comparing qualitative variables",
               start_row = 5,
               end_row = 5,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
    pack_rows("Correlational RQs",
               start_row = 6,
               end_row = 6,
               bold = FALSE,
               italic = TRUE,
               hline_before = TRUE) %>%
    add_header_above( c("Summaries" = 2,
                        "Analysis" = 2),
                      line = TRUE,
                      bold = TRUE)
}

```






<!-- ```{r InferenceTestCI} -->
<!-- Scenarios <- array( dim = c(6, 5) ) -->
<!-- colnames(Scenarios) <- c("Scenario", -->
<!--                          "Graphical", -->
<!--                          "Numerical", -->
<!--                          "\\stackunder{Hypothesis}{test}", -->
<!--                          "\\stackunder{Confidence}{interval}") -->


<!-- if( knitr::is_latex_output() ) { -->
<!--   Scenarios[1, ] <- c("\\stackunder{Proportion in}{one sample}", -->
<!--                       "\\stackunder{Bar chart;}{\\stackunder{pie chart;}{dotchart}}", -->
<!--                       "\\stackunder{Counts;}{\\stackunder{percentages;}{odds}}", -->
<!--                       "\\stackunder{One-sample}{$z$-test}", -->
<!--                       "\\stackunder{For one}{proportion}") -->
<!--   Scenarios[2, ] <- c("\\stackunder{Means in}{one sample}", -->
<!--                       "\\stackunder{Histogram;}{\\stackunder{stemplot;}{dot chart}}", -->
<!--                       "\\stackunder{Means, medians;}{\\stackunder{Std. dev., IQR;}{outliers, etc.}}", -->
<!--                       "\\stackunder{One-sample}{$t$-test}", -->
<!--                       "For one mean") -->
<!--   Scenarios[3, ] <- c("\\stackunder{Comparing quant.}{\\stackunder{vars: within}{-individuals (paired)}}", -->
<!--                       "\\stackunder{Histogram of \\emph{diffs.};}{case-profile}", -->
<!--                       "\\stackunder{Of the}{\\stackunder{difference, and}{for each group}}", -->
<!--                       "\\stackunder{$t$-test for}{mean diffs.}", -->
<!--                       "For mean diff.") -->
<!--   Scenarios[4, ] <- c("\\stackunder{Comparing quant.}{\\stackunder{vars: between}{-individuals}}", -->
<!--                       "\\stackunder{Boxplot;}{error bar chart}", -->
<!--                       "\\stackunder{For each group;}{\\stackunder{difference between}{\\stackunder{means; std.\\ error of}{of diff. between means}}}", -->
<!--                       "\\stackunder{$t$-test for the}{\\stackunder{diff. between}{two means}}",  -->
<!--                       "\\stackunder{For diff.}{\\stackunder{between two}{means}}") -->
<!--   Scenarios[5, ] <- c("\\stackunder{Comparing qual.}{\\stackunder{vars: between}{-individuals}}", -->
<!--                       "\\stackunder{Side-by-side}{\\stackunder{bar chart;}{stacked bar chart}}", -->
<!--                       "\\stackunder{Odds;}{\\stackunder{OR;}{percentages}}", -->
<!--                       "Chi-square test", -->
<!--                       "For ORs") -->
<!--   Scenarios[6, ] <- c("\\stackunder{Connection}{\\stackunder{between}{quant. vars}}", -->
<!--                       "Scatterplot", -->
<!--                       "Correlation", -->
<!--                       "\\stackunder{Correlation;}{\\stackunder{$t$-test for}{\\stackunder{regression}{parameters}}}", -->
<!--                       "\\stackunder{For regression}{parameters}") -->

<!--   kable(Scenarios, -->
<!--         booktabs = TRUE, -->
<!--         longtable = FALSE, -->
<!--         escape = FALSE, -->
<!--         linesep = "\\addlinespace\\addlinespace", -->
<!--         # align = c("p{20mm}", "p{25mm}", "p{25mm}", "p{25mm}", "p{20mm}"), -->
<!--         caption = "Five scenarios studied so far, plus one to follow", -->
<!--         format = "latex") %>% -->
<!--     kable_styling(font_size = 9) %>% -->
<!--     row_spec(0, bold = TRUE) %>% -->
<!--     column_spec(1, italic = TRUE) %>% -->
<!--     add_header_above( c(" ",  -->
<!--                         "Summaries" = 2, -->
<!--                         "Analysis" = 2), -->
<!--                       bold = TRUE) -->
<!-- }                  -->
<!-- if( knitr::is_html_output() ) { -->
<!--   Scenarios[1, ] <- c("Proportion in one sample", -->
<!--                       "Bar chart; pie chart; dotchart", -->
<!--                       "Counts; percentages; odds", -->
<!--                       "One-sample $z$-test", -->
<!--                       "For one proportion") -->
<!--   Scenarios[2, ] <- c("Means in one sample", -->
<!--                       "Histogram; stemplot; dot chart", -->
<!--                       "Means, medians; Std. dev., IQR; outliers, etc.", -->
<!--                       "One-sample $t$-test", -->
<!--                       "For one mean") -->
<!--   Scenarios[3, ] <- c("Comparing quant. vars: within-individuals (paired)", -->
<!--                       "Histogram of diffs.; case-profile", -->
<!--                       "Of the difference, and each group", -->
<!--                       "$t$-test for mean diffs.", -->
<!--                       "For mean diff") -->
<!--   Scenarios[4, ] <- c("Comparing quant. vars: between-individuals", -->
<!--                       "Boxplot; error bar chart", -->
<!--                       "Diff. between means; std. error of diff.; For each group", -->
<!--                       "$t$-test for the diff. between two means", -->
<!--                       "For diff. between two means") -->
<!--   Scenarios[5, ] <- c("Comparing qual. vars: between-individuals", -->
<!--                       "Side-by-side bar chart; stacked bar chart", -->
<!--                       "Odds; OR; percentages", -->
<!--                       "Chi-square test", -->
<!--                       "For ORs") -->
<!--   Scenarios[6, ] <- c("Connection between quant. vars", -->
<!--                       "Scatterplot", -->
<!--                       "Correlation", -->
<!--                       "Correlation; $t$-test for regression parameters", -->
<!--                       "For regression parameters") -->

<!--   kable(Scenarios, -->
<!--         booktabs = TRUE, -->
<!--         longtable = FALSE, -->
<!--         escape = FALSE, -->
<!--         linesep = "\\addlinespace\\addlinespace", -->
<!--         # align = c("p{20mm}", "p{25mm}", "p{25mm}", "p{25mm}", "p{20mm}"), -->
<!--         caption = "Five scenarios studied so far, plus one to follow", -->
<!--         format = "latex") %>% -->
<!--     kable_styling(font_size = 9) %>% -->
<!--     row_spec(0, bold = TRUE) %>% -->
<!--     column_spec(1, italic = TRUE) %>% -->
<!--     add_header_above( c(" ",  -->
<!--                         "Summaries" = 2, -->
<!--                         "Analysis" = 2), -->
<!--                       bold = TRUE) -->
<!-- }                  -->

<!-- ``` -->



## Exercises {#SelectTestExercises}

Answers to odd-numbered questions are available in App.\ \@ref(Answers).


:::{.exercise #Method1}
Suppose researchers compare the average number of hours of exercise per week for office workers, both in summer and in winter, to see if the averages are different.  

What methods would be a suitable for creating a summary and performing analyses?
:::

:::{.exercise #Method2}
Suppose we wish to estimate difference between the mean number of hours of sunlight exposure per day for female and male teachers.  

What methods would be a suitable for creating a summary and performing analyses?
:::

:::{.exercise #Method3}
Suppose researchers wish to compare the proportion of trees with koalas in them, comparing trees more than $10$\ m tall with trees $10$\ m or shorter.  

What methods would be a suitable for creating a summary and performing analyses?
:::

:::{.exercise #Method4}
Suppose researchers are wanting to compare the number of hours spend on social media for people aged over $30$, to people aged $30$ and under.  

What methods would be a suitable for creating a summary and performing analyses?
:::

 
`r if (knitr::is_html_output()){
  'The following short video may help explain some of these concepts. Note that the test for correlation and regression have not yet been covered in this book (but they will be in the next few chapters).'
}`


<div style="text-align:center;">
```{r}
htmltools::tags$video(src = "./videos/SelectTest.mp4", 
                      width = "550", 
                      controls = "controls", 
                      loop = "loop", 
                      style = "padding:5px; border: 2px solid gray;")
```
</div>


