

# Tables

This Appendix contains tables that may be useful:

* Random numbers (Appendix \@ref(AppendixRandomNumbers)).
* $z$-tables: Find the area associated with a normal distribution *given* the $z$-score
`r if (knitr::is_html_output(exclude = "epub3")){
   '(Appendix \\@ref(ZTablesOnline)).'
} else {
   '(Appendices \\@ref(ZTablesNEG) and \\@ref(ZTablesPOS)).'
}`
`r if (knitr::is_html_output(exclude = "epub3")){
'* $z$-tables: Find the $z$-score when an area under a normal distribution is known
  (Appendix \\@ref(ZTablesOnlineBackwards)).'
}`



## Random numbers {#AppendixRandomNumbers}
 
 
```{r , echo=FALSE}
num.rows <- 26
num.cols <- 5
numbers.per.cell <- 5

total.numbers <- num.rows * num.cols * numbers.per.cell
random.numbers <- sample(0:9, 
                         total.numbers, 
                         replace = TRUE)

rn.array <- array(random.numbers, 
                  dim = c(num.rows, num.cols) )
rownames(rn.array) <- paste( "Row", 
                             LETTERS)
colnames(rn.array) <- paste("Column", 
                            c("I","II","III","IV","V") )

for (i in (1 : num.rows)){
  for (j in (1 : num.cols)){
     rn.array[i,j] <- paste( sample(0:9, 
                                    size = numbers.per.cell, 
                                    replace = TRUE), 
                             collapse = " " )
  }
}


if( knitr::is_latex_output() ) {
  knitr::kable(rn.array,
               format = "latex",
               longtable = FALSE,
               caption = "Some random numbers",
               booktabs = TRUE) %>%
  kable_styling(font_size = 9) %>%
  row_spec(0, bold = TRUE) 
}1

if( knitr::is_html_output() ) {
  out <- knitr::kable(rn.array,
               format = "html",
               longtable = FALSE,
               caption = "Some random numbers",
               booktabs = TRUE) %>%
    row_spec(row = 0, 
             bold = TRUE) 
}

if( knitr::pandoc_to("epub3") ) {
  knitr::kable(rn.array,
               format = "html",
               longtable = FALSE,
               caption = "Some random numbers",
               booktabs = TRUE)
}
```







<!-- Appendix B: Table -->
<!-- Forward tables -->




```{r, child = if (knitr::is_html_output(excludes = "epub")) './children/ZtablesHTML.Rmd'}
```

```{r, child = if (knitr::is_html_output(excludes = "epub")) './children/InstructionsZtablesHTML.Rmd'}
```


<!-- The  LaTeX file contains *both* forwards and backwards tables -->
```{r, child = if (knitr::is_latex_output()) './children/ZtablesLaTeX.Rmd'}
```




```{r, child = if (knitr::pandoc_to("epub3")) './children/ZtablesLaTeX.Rmd'}
```



<!-- Backwards tables for HTML only -->

```{r, child = if (knitr::is_html_output(excludes = "epub")) './children/ZtablesHTMLBackwards.Rmd'}
```

<!-- The  LaTeX file contains *both* forwards and backwards tables -->


```{r, child = if (knitr::is_html_output(excludes = "epub")) './children/InstructionsZtablesHTMLBackwards.Rmd'}
```





