# Tables {#Tables}

<!-- APPENDIX B: Table -->


This Appendix contains tables that may be useful:

* Random numbers (Appendix \@ref(AppendixRandomNumbers)).
* $z$-tables: Find the area associated with a normal distribution *given* the $z$-score
`r if (knitr::is_html_output()){
   '(Appendix \\@ref(ZTablesOnline)).'
} else {
   '(Appendices \\@ref(ZTablesNEG) and \\@ref(ZTablesPOS)).'
}`
`r if (knitr::is_html_output()){
'* $z$-tables: Find the $z$-score when an area under a normal distribution is known
  (Appendix \\@ref(ZTablesOnlineBackwards)).'
}`


## Random numbers {#AppendixRandomNumbers}
 
 
```{r , echo=FALSE}
num.rows <- 26
num.cols <- 5
numbers.per.cell <- 5

total.numbers <- num.rows * num.cols * numbers.per.cell
random.numbers <- sample(0:9, 
                         total.numbers, 
                         replace = TRUE)

rn.array <- array(random.numbers, 
                  dim = c(num.rows, num.cols) )
rownames(rn.array) <- paste( "Row", 
                             LETTERS)
colnames(rn.array) <- paste("Column", 
                            c("I","II","III","IV","V") )

for (i in (1 : num.rows)){
  for (j in (1 : num.cols)){
     rn.array[i,j] <- paste( sample(0:9, 
                                    size = numbers.per.cell, 
                                    replace = TRUE), 
                             collapse = " " )
  }
}


if( knitr::is_latex_output() ) {
  knitr::kable(rn.array,
               format = "latex",
               longtable = FALSE,
               caption = "Some random numbers",
               booktabs = TRUE) %>%
  kable_styling(font_size = 9) %>%
  row_spec(0, bold = TRUE) 
}

if( knitr::is_html_output() ) {
  knitr::kable(rn.array,
               format = "html",
               longtable = FALSE,
               caption = "Some random numbers",
               booktabs = TRUE) %>%
    row_spec(row = 0, 
             bold = TRUE) 
}
```






<!-- The  LaTeX files: Only forwards tables, in two bits (positive z and negative z) -->
```{r, child = if (knitr::is_latex_output()) './Tables/Ztables-LaTeX.Rmd'}
```



<!-- The HTML tables (plus instructions) -->
<!-- Forward (z to Area) tables -->

```{r, child = if (knitr::is_html_output()) './Tables/Ztables-HTML.Rmd'}
```

```{r, child = if (knitr::is_html_output()) './Tables/Ztables-HTML-Instructions.Rmd'}
```


<!-- The HT<L tables also includes backwards (Area to z) tables (NOT LATEX) -->
```{r, child = if (knitr::is_html_output()  ) './Tables/Ztables-Backwards-HTML.Rmd' }
```

```{r, child = if (knitr::is_html_output() ) './Tables/Ztables-Backwards-Instructions-HTML.Rmd'} 
```


