# $z$-score tables {#Tables}

<!-- APPENDIX B: Table -->

<!-- Different forwards-z tables pointers for ONLINE and PDF: -->
<!-- * ONLINE: Only have *one* link to the forward tables  -->
<!-- * PDF: Have *two* links to the forward tables: Negative and positive z-values  -->

<!-- Different backwards z-table pointers for ONLINE and PDF: -->
<!-- * ONLINE: Point to the backwards-z tables  -->
<!-- * PDF: NO pointer, as we just use the forward-tables  -->



<!-- Define colours as appropriate -->
```{r, child = if (knitr::is_html_output()) {'./children/coloursHTML.Rmd'} else {'./children/coloursLaTeX.Rmd'}}
```



This Appendix contains $z$-score tables.

<!-- * Random numbers (Appendix\ \@ref(AppendixRandomNumbers)). -->
These tables provide the area *to the left* of a given $z$-score associated with a normal
<!-- HTML: One set of tables only is needed -->
<!-- PDF: Two set of tables: NEG and POS z-scores -->
`r if (knitr::is_html_output()){
   'distribution (Appendix\\ \\@ref(ZTablesOnline)),'  
} else {
   'distribution: Appendices\\ \\@ref(ZTablesNEG) (for negative values of\ $z$) and\\ \\@ref(ZTablesPOS) (for positive values of\ $z$).'   
}`
<!-- Only for online: PDF uses forward tables -->
`r if (knitr::is_html_output()){
' or the $z$-score such that a given area is to the left (Appendix\\ \\@ref(ZTablesOnlineBackwardsHTML)).'
}`

`r if (knitr::is_latex_output()) {
  'The online version of this book (https://bookdown.org/pkaldunn/SRM-Textbook)'
} else {
   'The [online version of this book](https://bookdown.org/pkaldunn/SRM-Textbook/AppendixDataSets.html) '
}`
has online tables, which are easier to use.



```{r, child = if (knitr::is_latex_output()) './Tables/Ztables-Using-Hardcopy-Appendix.Rmd'}
```

\pagebreak

<!-- ## Random numbers {#AppendixRandomNumbers} -->


<!-- ```{r} -->
<!-- num.rows <- 26 -->
<!-- num.cols <- 5 -->
<!-- numbers.per.cell <- 5 -->

<!-- total.numbers <- num.rows * num.cols * numbers.per.cell -->
<!-- random.numbers <- sample(0:9,  -->
<!--                          total.numbers,  -->
<!--                          replace = TRUE) -->

<!-- rn.array <- array(random.numbers,  -->
<!--                   dim = c(num.rows, num.cols) ) -->
<!-- rownames(rn.array) <- paste( "Row",  -->
<!--                              LETTERS) -->
<!-- colnames(rn.array) <- paste("Column",  -->
<!--                             c("I","II","III","IV","V") ) -->

<!-- for (i in (1 : num.rows)){ -->
<!--   for (j in (1 : num.cols)){ -->
<!--      rn.array[i,j] <- paste( paste0("$", -->
<!--                                     sample(0:9,  -->
<!--                                            size = numbers.per.cell,  -->
<!--                                            replace = TRUE), -->
<!--                                     "$"), -->
<!--                              collapse = " " ) -->
<!--   } -->
<!-- } -->


<!-- if( knitr::is_latex_output() ) { -->
<!--   knitr::kable(surroundMaths(rn.array), -->
<!--                format = "latex", -->
<!--                longtable = FALSE, -->
<!--                escape = FALSE, -->
<!--                #caption = "Some random numbers", -->
<!--                booktabs = TRUE) %>% -->
<!--   kable_styling(font_size = 9) %>% -->
<!--   row_spec(0, bold = TRUE)  -->
<!-- } -->

<!-- if( knitr::is_html_output() ) { -->
<!--   knitr::kable(rn.array, -->
<!--                format = "html", -->
<!--                longtable = FALSE, -->
<!--                escape = FALSE, -->
<!--                #caption = "Some random numbers", -->
<!--                booktabs = TRUE) %>% -->
<!--     row_spec(row = 0,  -->
<!--              bold = TRUE)  -->
<!-- } -->
<!-- ``` -->



<!-- ********* LaTeX *********** -->
<!-- The  LaTeX files: Only forwards tables, in two bits (positive z and negative z) -->
```{r, child = if (knitr::is_latex_output()) './Tables/Ztables-LaTeX.Rmd'}
```


<!-- ********* HTML *********** -->
<!-- The HTML tables (plus instructions) -->
<!-- Forward (z to Area) tables -->

```{r, child = if (knitr::is_html_output()) './Tables/Ztables-Online.Rmd'}
```

```{r, child = if (knitr::is_html_output()) './Tables/Ztables-Using-Online-Instructions.Rmd'}
```


<!-- The above HTML tables also includes backwards (Area to z) tables (NOT LATEX) -->
```{r, child = if (knitr::is_html_output()  ) './Tables/Ztables-Backwards-Online.Rmd' }
```

```{r, child = if (knitr::is_html_output() ) './Tables/Ztables-Backwards-Using-Online-Instructions.Rmd'} 
```


