
# Symbols, formulas, statistics and parameters {#StatisticsAndParameters}

## Symbols and standard errors {#Symbols}

* The following table lists the statistics used to estimate unknown population parameters. \tightlist
* The sampling distribution is given for each statistic.
* When the sampling distribution is approximately normally distributed, under appropriate statistical validity conditions, this is indicated by\ \ding{51}.\index{Sampling distribution}
* The value of the mean of the sampling distribution (the *sampling mean*) is\index{Sampling mean}:

  - unknown, for *confidence intervals*.
  - assumed to be the value given in the null hypothesis, for *hypothesis tests*.
  


(ref:SamplingDistributionPropCI) Ch.\ \@ref(CIOneProportion)

(ref:SamplingDistributionPropHT) Ch.\ \@ref(TestOneProportion)

(ref:SamplingDistributionXbarCI) Ch.\ \@ref(OneMeanConfInterval)

(ref:SamplingDistributionXbarHT) Ch.\ \@ref(TestOneMean)

(ref:SamplingDistributionDbarCI) Ch.\ \@ref(PairedCI)

(ref:SamplingDistributionDbarHT) Ch.\ \@ref(TestPairedMeans)

(ref:SamplingDistributionTwoMeansCI) Ch.\ \@ref(CITwoMeans)

(ref:SamplingDistributionTwoMeansHT) Ch.\ \@ref(TestTwoMeans)

(ref:SamplingDistributionTwoPropsCI) Ch.\ \@ref(OddsRatiosCI)

(ref:SamplingDistributionTwoPropsHT) Ch.\ \@ref(TestsOddsRatio)

(ref:SamplingDistributionORCI) Ch.\ \@ref(OddsRatiosCI)

(ref:SamplingDistributionORHT) Ch.\ \@ref(TestsOddsRatio)

(ref:SamplingDistributionCorrelationHT) Ch.\ \@ref(Correlation)

(ref:SamplingDistributionRegression) Ch.\ \@ref(Regression)


```{r ParametersStatistics2}
ParStat <- array( dim = c(18, 6))

colnames(ParStat) <- c("",
                       #"Param.",
                       "Statistic",
                       "Mean",
                       "Std error",
                       "Normal?",
                       "Ref.")
rownames(ParStat) <- c("Proportion",
                       "Proportion",
                       "Mean",
                       "Mean",
                       "Mean difference",
                       "Mean difference",
                       "Difference between means",
                       "Difference between means",
                       "Diff. between proportions",
                       "Diff. between proportions",
                       "Odds ratio",
                       "Odds ratio",
                       "Correlation",
                       "Correlation",
                       "Regression: slope",
                       "Regression: slope",
                       "Regression: intercept",
                       "Regression: intercept")

ParStat[1, ] <- c("CI:",
#                  "$p$",
                  "$\\hat{p}$", 
                  "$p$", 
                  "CI: $\\displaystyle \\sqrt{\\frac{ \\hat{p} \\times (1 - \\hat{p})}{n}}$",
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                  "CI: (ref:SamplingDistributionPropCI)")
ParStat[2, ] <- c("Test:",
#                  "$p$",
                  "$\\hat{p}$", 
                  "$p$", 
                  "HT: $\\displaystyle \\sqrt{\\frac{ p \\times (1 - p)}{n}}$",
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                  "HT: (ref:SamplingDistributionPropHT)")

ParStat[3, ] <- c("CI:",
#                  "$\\mu$", 
                 "$\\bar{x}$",
                 "$\\mu$",
                 "$\\displaystyle \\frac{s}{\\sqrt{n}}$", 
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                 "CI: (ref:SamplingDistributionXbarCI)")
ParStat[4, ] <- c("Test:",
#                  "$\\mu$", 
                 "$\\bar{x}$", 
                 "$\\mu$",
                 "$\\displaystyle \\frac{s}{\\sqrt{n}}$", 
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                 "HT: (ref:SamplingDistributionXbarHT)")
ParStat[5, ] <- c("CI:",
#                  "$\\mu_d$", 
                 "$\\bar{d}$",
                 "$\\mu_d$",
                 "$\\displaystyle \\frac{s_d}{\\sqrt{n}}$", 
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                 "CI: (ref:SamplingDistributionDbarCI)")
ParStat[6, ] <- c("Test:",
#                  "$\\mu_d$", 
                 "$\\bar{d}$", 
                 "$\\mu_d$",
                 "$\\displaystyle \\frac{s_d}{\\sqrt{n}}$", 
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                 "HT: (ref:SamplingDistributionDbarHT)")
ParStat[7, ] <- c("CI:",
#                  "$\\mu_1 - \\mu_2$", 
                 "$\\bar{x}_1 - \\bar{x}_2$",
                 "$\\mu_1 - \\mu_2$",
                 "$\\displaystyle \\sqrt{\\text{s.e.}(\\bar{x}_1) + \\text{s.e.}(\\bar{x}_2)}$", 
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                 "CI: (ref:SamplingDistributionTwoMeansCI)")
ParStat[8, ] <- c("Test:",
#                  "$\\mu_1 - \\mu_2$", 
                 "$\\bar{x}_1 - \\bar{x}_2$",
                 "$\\mu_1 - \\mu_2$",
                 "$\\displaystyle \\sqrt{\\text{s.e.}(\\bar{x}_1) + \\text{s.e.}(\\bar{x}_2)}$", 
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                 "HT: (ref:SamplingDistributionTwoMeansHT)")# ParStat[6, ] <- c("$\\sigma$", 

ParStat[9, ] <- c("CI:",
#                 "$p_1 - p_2$",
                 "$\\hat{p}_1 - \\hat{p}_2$", 
                 "$p_1 - p_2$",
                 "CI: $\\displaystyle \\sqrt{\\text{s.e.}(\\hat{p}_1) + \\text{s.e.}(\\hat{p}_2)}$", 
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                 "CI: (ref:SamplingDistributionTwoPropsCI)")
ParStat[10, ] <- c("Test:",
#                  "$p_1 - p_2$", 
                 "$\\hat{p}_1 - \\hat{p}_2$",
                 "$p_1 - p_2$",
                 "HT: $\\displaystyle \\sqrt{\\text{s.e.}(\\hat{p}) + \\text{s.e.}(\\hat{p})}$ for common proportion $\\hat{p}$", 
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                 "HT: (ref:SamplingDistributionTwoPropsHT)")# ParStat[6, ] <- c("$\\sigma$", 
ParStat[11, ] <- c("CI:",
                 "Sample OR", 
                 "Pop. OR",
                 "(Not given)",
                  ifelse( knitr::is_latex_output(), "\\ding{55}", "&#10008;"),
                 "CI: (ref:SamplingDistributionORCI)")
ParStat[12, ] <- c("Test:",
                 "Sample OR", 
                 "Pop. OR",
                 "(Not given)",
                  ifelse( knitr::is_latex_output(), "\\ding{55}", "&#10008;"),
                 "HT: (ref:SamplingDistributionORHT)")
ParStat[13, ] <- c("Test:",
                  "$r$",
                  "$\\rho$",
                  "(Not given)",
                  ifelse( knitr::is_latex_output(), "\\ding{55}", "&#10008;"),
                  " ")
ParStat[14, ] <- c("CI:",
                  "$r$",
                  "$\\rho$",
                  "(Not given)",
                  ifelse( knitr::is_latex_output(), "\\ding{55}", "&#10008;"),
                  "HT: (ref:SamplingDistributionCorrelationHT)")
ParStat[15, ] <- c("Test:",
                  "$b_1$",
                  "$\\beta_1$",
                  "$\\text{s.e.}(b_1)$ (value from software)",
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                  "CI: (ref:SamplingDistributionRegression)")
ParStat[16, ] <- c("CI:",
                  "$b_1$",
                  "$\\beta_1$",
                  "$\\text{s.e.}(b_1)$ (value from software)",
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                  "HT: (ref:SamplingDistributionRegression)")
ParStat[17, ] <- c("Test:",
                  "$b_0$",
                  "$\\beta_0$",
                  "$\\text{s.e.}(b_0)$ (value from software)",
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                  "CI: (ref:SamplingDistributionRegression)")
ParStat[18, ] <- c("CI:",
                  "$b_0$",
                  "$\\beta_0$",
                  "$\\text{s.e.}(b_0)$ (value from software)",
                  ifelse( knitr::is_latex_output(), "\\ding{52}", "&#10004;"),
                  "HT: (ref:SamplingDistributionRegression)")

if( knitr::is_latex_output() ) {
  kable( ParStat[ -13, c(2, 3, 5, 4, 6)],
         format = "latex",
         booktabs = TRUE,
         escape = FALSE,
         align = c("c", "c", "c", "c", "l"),
         col.names = c("Statistic",
                       "sampling mean",
                       "distn?",
                       "error",
                       "Ref."),
         linesep = "\\addlinespace",
         caption = "Sample statistics used to estimate population parameters. For statistics with standard errors given, the sampling distribution is approximately normally distributed under appropriate (statistical validity) conditions."
         ) %>%
	  row_spec(0, bold = TRUE) %>%
    #row_spec(0, angle = c(0, 0, 90, 90, 0, 0, 0, 0)) %>%
   column_spec(column = 1, width = "21mm") %>% 
   #column_spec(column = 2, width = "14mm") %>%
   #column_spec(column = 3, width = "15mm") %>%
   #column_spec(column = 4, width = "39mm") %>%
   column_spec(column = 5, width = "30mm") %>%
#   column_spec(column = 6, width = "24mm") %>%
    add_header_above( c(" " = 2, 
                        "Parameter, and" = 1,
                        "Normal" = 1,
                        "Standard" = 1,
                        " " = 1),
                      bold = TRUE,
                      line = FALSE) %>%
    add_header_above( c(" " = 2,
                        "Sampling distribution" = 3,
                        " " = 1),
                      bold = TRUE,
                      line = TRUE) %>%
    kable_styling(font_size = 8) %>% 
    collapse_rows(latex_hline = "custom",
                  target = 3,
                  custom_latex_hline = 6)
}
if( knitr::is_html_output() ) {
  kable( ParStat[ -13, c(2, 3, 5, 4, 6)],
         format = "html",
         col.names = c("Statistic",
                       "sampling mean",
                       "distn?",
                       "error",
                       "Ref."),
         booktabs = TRUE,
         escape = FALSE,
         align = c("l", "c", "c", "c", "c") ) %>%
         #caption = "Sample statistics used to estimate population parameters. Empty table cells means that these are not studied in this book. For statistics with standard errors given, the sampling distribution is approximately normally distributed under certain (statistical validity) conditions.") %>%
    row_spec(0, bold = TRUE) %>%
     add_header_above( c(" " = 2, 
                        "Parameter, and" = 1,
                        "Normal" = 1,
                        "Standard" = 1,
                        " " = 1),
                      bold = TRUE,
                      line = FALSE) %>%
    add_header_above( c(" " = 2,
                        "Sampling distribution" = 3,
                        " " = 1),
                      bold = TRUE,
                      line = TRUE) %>% 
    collapse_rows(latex_hline = "custom",
                  columns = 1:3,
                  target = 2,
                  custom_latex_hline = 6)
#                  custom_latex_hline = 6)
}
```





## Confidence intervals {#FormulasCI}

For statistics that have an approximate normal distribution, **confidence intervals** have the form
$$ 
    \text{statistic} \pm ( \text{multiplier} \times \text{s.e.}(\text{statistic})).
$$

**Notes:**

* The multiplier is *approximately*\ $2$ to create an *approximate* $95$% CI (based on the $68$--$95$--$99.7$ rule).
* The quantity '$\text{multiplier} \times \text{s.e.}(\text{statistic})$' is called the *margin of error*.
* When the sampling distribution for the statistic does *not* have an approximate normal distribution (e.g., for *odds ratios* and *correlation coefficients*), **this formula does not apply**.



\pagebreak

## Hypothesis testing {#FormulasTest}

For statistics that have an approximate normal distribution, the *test statistic* has the form:
$$
  \frac{\text{statistic} - \text{parameter}}{\text{s.e.}(\text{statistic})}.
$$
This quantity is a $t$-score for most hypothesis tests in this book.
However, this quantity is a $z$-score for a hypothesis test involving one or two *proportions*.



**Notes:**

* Since $t$-scores are approximately like $z$-scores (Sect.\ \@ref(TestStatistic)), the $68$--$95$--$99.7$ rule can be used to *approximate* $P$-values.
* When the sampling distribution for the statistic does not have an approximate normal distribution (e.g., for *odds ratios* and *correlation coefficients*), **this formula does not apply**.
* A hypothesis test about **odds ratios** uses a $\chi^2$ test statistic.
  For $2\times 2$ tables only, the $\chi^2$ value is equivalent to a $z$-score with a value of $\sqrt{\chi^2}$.





## Sample size estimation {#FormulasSampleSize}

All of the following formulas compute the approximate minimum sample size needed to produce a $95$% confidence interval with a specified margin of error.

* To estimate the sample size needed for **estimating a proportion** (Sect.\ \@ref(SampleSizeProportions)):
$$
   n = \frac{1}{(\text{Margin of error})^2}.
$$
* To estimate the sample size needed for **estimating a mean** (Sect.\ \@ref(SampleSizeOneMean)):
$$
   n = \left( \frac{2\times s}{\text{Margin of error}}\right)^2.
$$
* To estimate the sample size needed for **estimating a mean difference** (Sect.\ \@ref(SampleSizeMeanDifferences)):
$$
   n = \left( \frac{2 \times s_d}{\text{Margin of error}}\right)^2.
$$
* To estimate the sample size needed for **estimating the difference between two means** (Sect.\ \@ref(SampleSizeDifferenceTwoMeans)):
$$
   n = 2\times \left( \frac{2 \times s}{\text{Margin of error}}\right)^2
$$
for *each* sample, where $s$ is an estimate of the common standard deviation in the population for both groups.
This formula assumes:

  * the sample size in each group is the same; and
  * the standard deviation in each group is the same.





**Notes:**

* In **sample size calculations**, always **round up** the sample size found from the above formulas.


\pagebreak

## Other formulas {#FormulasOther}

* To **calculate $z$-scores** (Sect.\ \@ref(zScores)), use
$$
   z = \frac{\text{value of variable} - \text{mean of the distribution of the variable}}{\text{standard deviation of the distribution of the variable}}.
$$
  $t$-scores are like $z$-scores.
  When the 'variable' is a sample estimate, the 'standard deviation of the distribution' is a standard error.
* The **unstandardizing formula** (Sect.\ \@ref(Unstandardising)) is $x = \mu + (z\times \sigma)$.
* The **interquartile range** (IQR) is $Q_3 - Q_1$, where $Q_1$ and $Q_3$ are the first and third quartiles respectively (or, equivalently, the $25$th and $75$th percentiles).
* The smallest expected value (for assessing statistical validity when forming CIs and conducting hypothesis tests with proportions or odds ratio) is
$$
  \frac{(\text{Smallest row total})\times(\text{Smallest column total})}{\text{Overall total}}.
$$
* The **regression equation** in the *sample* is $\hat{y} = b_0 + b_1 x$, where $b_0$ is the sample intercept and $b_1$ is the sample slope.
* The **regression equation** in the *population* is $\hat{y} = \beta_0 + \beta_1 x$, where $\beta_0$ is the intercept and $\beta_1$ is the slope.




## Other symbols and abbreviations used {#OtherSymbols}

(ref:RQs) Chap.\ \@ref(RQs)

(ref:AboutHypotheses) Sect.\ \@ref(AboutHypotheses)

(ref:TestStatObs) Sect.\ \@ref(TestStatObs)

(ref:AboutCIs) Chap.\ \@ref(AboutCIs)

(ref:StandardError) Def.\ \@ref(def:StandardError)

(ref:Rsquared) Sect.\ \@ref(Rsquared)

(ref:VariationStdDev) Sect.\ \@ref(VariationStdDev)



```{r OtherSymbols}
SymbolsTab <- array( dim = c(12, 3))

colnames(SymbolsTab) <- c("Symbol", 
                          "Meaning", 
                          "Reference")

SymbolsTab[1, ] <- c("RQ", 
                     "Research question", 
                     "(ref:RQs)")
SymbolsTab[2, ] <- c("$s$", 
                     "Sample standard deviation", 
                     "(ref:VariationStdDev)")
SymbolsTab[3, ] <- c("$\\sigma$", 
                     "Population standard deviation", 
                     "(ref:VariationStdDev)")
SymbolsTab[4, ] <- c("$s_d$", 
                     "Sample standard deviation of differences", 
                     "(ref:VariationStdDev)")
SymbolsTab[5, ] <- c("$\\sigma_d$", 
                     "Population standard deviation of differences", 
                     "(ref:VariationStdDev)")
SymbolsTab[6, ] <- c("$R^2$", 
                     "R-squared", 
                     "(ref:Rsquared)")
SymbolsTab[7, ] <- c("$H_0$", 
                     "Null hypothesis", 
                     "(ref:AboutHypotheses)")
SymbolsTab[8, ] <- c("$H_1$", 
                     "Alternative hypothesis", 
                     "(ref:AboutHypotheses)")
SymbolsTab[9, ] <- c("CI", 
                     "Confidence interval", 
                     "(ref:AboutCIs) ")
SymbolsTab[10, ] <- c("s.e.", 
                     "Standard error" , 
                     "(ref:StandardError)")
SymbolsTab[11, ] <- c("$n$", 
                     "Sample size", "")
SymbolsTab[12, ] <- c("$\\chi^2$", 
                     "The chi-squared test statistic", 
                     "(ref:TestStatObs) ")


if( knitr::is_latex_output() ) {
  kable( SymbolsTab,
         format = "latex",
         booktabs = TRUE,
         escape = FALSE,
         align = c("c", "l", "c"),
         linesep = c("\\addlinespace", "", "\\addlinespace", "",  "\\addlinespace", "\\addlinespace", "", "\\addlinespace", "", "")) %>%  # Else adds a space every 5 lines...
        # caption = "Some symbols used") %>%
	row_spec(0, bold = TRUE) %>%
  kable_styling(font_size = 9)
}
if( knitr::is_html_output() ) {
  kable( SymbolsTab,
                format = "html",
                booktabs = TRUE,
                align = c("c", "l", "c")) %>%
                #caption = "Some symbols used")
  row_spec(0, bold = TRUE)
}
```

