# Glossary {#Glossary}

<!-- Add things like boxplot, histogram, error bar chart, etc? -->


<!-- Define colours as appropriate -->
```{r, child = if (knitr::is_html_output()) {'./children/coloursHTML.Rmd'} else {'./children/coloursLaTeX.Rmd'}}
```



```{=latex}
\begin{small}
```


[$68$--$95$--$99.7$ rule](#def:EmpiricalRule)\index{68@$68$--$95$--$99.7$ rule|textbf}
: For *any* bell-shaped distribution, *approximately* $68$% of observations lie within one standard deviation of the mean, $95$% of observations lie within two standard deviations of the mean, and $99.7$% of observations lie within three standard deviations of the mean.
Also called the *empirical rule*.


[Accuracy](#PrecisionAccuracy)\index{Accuracy|textbf}
: *Accuracy* refers to how close a *sample* estimate is likely to be to the *population* value, on average.


[Alternative hypothesis](#def:AltHypothesis)\index{Hypotheses!alternative|textbf}
: The *alternative hypothesis* proposes that any difference, change or relationship observed in the sample is because a difference, change or relationship exists the *population* (that is, the difference, change or relationship cannot be explained by sampling variation).


[Bell-shaped distributions](#DistributionsExample)
: See *Normal distributions*.


[Between-individual comparisons](#def:ComparisonBetween)\index{Comparison!between individuals|textbf}
: See *Comparison (between individuals)*, *Comparison*.


[Bias](#def:Bias)\index{Bias|textbf}
: *Bias* refers to any systematic misrepresentation of the target population by the sample.


[Blinding](#DescribingBlinding)\index{Blinding|textbf}
: *Blinding* when those involved in the study do not know which comparison group the individuals are in.
A study can blind the *researcher* to knowing what comparison group the individuals are in, the *participants* to knowing what comparison group they are in, and/or the *analysts* to knowing what comparison group the individuals are in during analysis.


[Blocking](#def:Blocking)\index{Blocking|textbf}
: *Blocking* occurs when units of analysis are arranged or analysed as separate groups of similar units (called *blocks*).


[Carry-over effect](#def:CarryoverEffect)\index{Carry-over effect|textbf}
: The carry-over effect occurs when the influence of one treatment or condition influences individuals' responses to subsequent treatments or conditions.
(Sect.\ \@ref(CarryOverEffect)).



[Cases](#def:UnitOfAnalysis)\index{Cases|textbf}
: The individual units in the population; the [*units of analysis*](#def:UnitOfAnalysis).
Also called *individuals*, or (when the individuals are people) *subjects*.



[Categorical data](#def:QuantitativeData)
: See *qualitative data*.


[Chi-square ($\chi^2$) score](#TestStatObs)
: The test-statistic used when comparing odds (or proportions).\index{chi@$\chi^2$-score|textbf}
The $\chi^2$ statistic measures the overall size of the differences between the expected counts and observed counts, over the entire two-way table.


[Classical approach to probability](#def:ClassicalApproachToProbability)\index{Probability!classical approach|textbf}
: In the *classical approach to probability*, the probability of an event occurring is the number of elements of the sample space included in the event, divided by the total number of elements in the sample space, *when all outcomes are equally likely*.



[Cluster sampling](#ClusterSampling)
: A sample where the population is split into a large number of small groups called *clusters*, then a *simple random sample* of clusters is selected and *every* member of the chosen small groups is part of the sample.


Comparison
: In a RQ, a *comparison* refers to groups of individuals being compared.
The comparison may be *within* individuals, or *between* groups of individuals.
Also see *Comparison (between individuals)*, *Comparison (within individuals)*.


[Comparison (between individuals)](#def:ComparisonBetween)\index{Comparison!between individuals|textbf}
: The *between-individuals comparison* in the RQ identifies the small number of groups of different individuals for which the outcome is compared.
Also see *Comparison (within individuals)*, *Comparison*.


[Comparison (within individuals)](#def:ComparisonWithin)
: The *within-individuals comparison* in the RQ identifies the small number of different, distinct situations for which the outcome is compared for each individual.
Also see *Comparison (between individuals)*, *Comparison*.


[Compound event](#def:CompoundEvent)\index{Compound event|textbf}
: A *compound event* is any combination of *simple events*.


[Conceptual definition](#def:ConceptualDefinition)\index{Definitions!conceptual|textbf}
: A *conceptual definition* articulates precisely *what* words or phrases mean in a study.


[Conditions](#def:Conditions)\index{Conditions|textbf}
: The *conditions* are the values of the comparison that those in the *observational* study experience, but are not imposed by the researchers.


[Confidence interval](#def:ConfidenceInterval)\index{Confidence intervals|textbf}
: Informally: a *confidence interval* (CI) is an interval likely to contain the unknown value of the [*parameter*](#def:Parameter).
More formally, a CI is an interval which contains the unknown [*parameter*](#def:Parameter) a given percentage of the time (over repeated sampling).
We studied CIs in some specific situations (see Sect.\ \@ref(Symbols)); there are hundreds more.


[Confounding](#def:Confounding)\index{Confounding|textbf}
: *Confounding* is when a third variable influences the observed relationship between the response and explanatory variable. 
 
 
[Confounding variable](#def:ConfoundingVariable)\index{Variables!confounding|textbf}
: A *confounding variable* (or a *confounder*) is an extraneous variable associated with the response and explanatory variables.


[Continuous data](#def:ContinuousData)\index{Data!quantitative continuous|textbf}\index{Data!continuous quantitative|textbf}\index{Variables!continuous quantitative|textbf}
: *Continuous* quantitative data has (at least in theory) an infinite number of possible values between any two given values. 


[Control](#def:Control)\index{Control|textbf}
: A *control* is a unit of analysis without the treatment or condition of interest, but as similar as possible in every other way to other units of analysis.


[Convenience sample](#NonRandomSamples)\index{Sampling!convenience|textbf}
: A sample where individuals are selected because they are convenient for the researcher.


[Correlational research question](#def:CorrelationalRQ)\index{Correlational RQ|textbf}
: Correlational RQs explore the relationship between two variables.



[Data](#def:Data)\index{Data|textbf}
: *Data* refers items of information obtained from a study (such as height of seedlings, or the type of medication given).


[Dataset](#def:Data)\index{Dataset|textbf}
: A *dataset* refers to an organised and structured collection of data.


[Descriptive research question](#def:DescriptiveRQ)\index{Research question!descriptive|textbf}
: Descriptive RQs have only a population and an outcome.



[Descriptive study](#def:DescriptiveStudy)\index{Study types!descriptive|textbf}
: *Descriptive studies* answer descriptive research questions.
 
 
[Discrete data](#def:DiscreteData)\index{Data!discrete quantitative|textbf}\index{Variables!discrete quantitative|textbf}
: *Discrete* quantitative data has a countable number of possible values between any two given values of the variable. 
 
 
[Distribution](#def:Distribution)\index{Distribution|textbf}
: The *distribution* of a variable describes what values are present in the data, and how often those values appear.


[Ecological validity](#def:EcologicalValidity)\index{Ecological validity|textbf}
: A study is *ecologically valid* if the study methods, materials and context closely approximate the real situation of interest.


[Event](#SampleSpaceEvents)\index{Event|textbf}
: An *event* is any combination of the elements in the [sample space](#def:SampleSpace).


[Exclusion criteria](#def:InclusionExclusionCriteria)\index{Exclusion criteria!conceptual|textbf}
: *Exclusion criteria* are characteristics that disqualify potential individuals from being included in the study.


[Empirical rule](#def:EmpiricalRule)\index{Empirical rule|textbf}
: For *any* bell-shaped distribution, *approximately* $68$% of observations lie within one standard deviation of the mean, *approximately* $95$% of observations lie within two standard deviations of the mean, and *approximately* $99.7$% of observations lie within three standard deviations of the mean.
Also called the *$68$--$95$--$99.7$ rule*.



[Experimental studies (or Experiments)](#def:Experiment)\index{Study types!experimental|textbf}
: *Experimental studies* (or *experiments*) study relationships *with* an intervention.



[Experimenter effect](#def:ObserverEffect)\index{Observer effect|textbf}
: See *Observer effect*



[Explanatory variable](#def:ExplanatoryVariable)\index{Variables!explanatory|textbf}
: An *explanatory variable* may (partially) explain or be associated with changes in another variable of interest (the response variable).
In an experimental study, it is the variable that can be manipulated by the researchers.


[External validity](#def:ExternalValidity)\index{External validity|textbf}
: *External validity* refers to the ability to generalise the results of the study to the rest of the population, beyond just those in the studied sample.
For a study to be truly externally valid, the sample must be a random sample from the population.



[Extraneous variable](#def:ExtraneousVariable)\index{Variables!extraneous|textbf}
: An *extraneous* variable is any variable associated with the response variable, but is not the explanatory variable.
 
 

[Extrapolation](#def:Extrapolation)\index{Extrapolation|textbf}
: *Extrapolation* refers to making prediction outside the range of the available data.
Extrapolation beyond the data may lead to nonsense predictions.


[Hawthorne effect](#def:HawthorneEffect)\index{Hawthorne effect|textbf}
: The *Hawthorne effect* is the tendency of individuals to change their behaviour if they know (or think) they are being observed (Sect.\ \@ref(HawthorneEffect)).


Hypothesis\index{Hypotheses|textbf}
: A *hypothesis* is a possible answer to a (research) question.
Also see *null hypothesis*,  *alternative hypothesis*.


Hypothesis test\index{Hypothesis testing|textbf}
: A *hypothesis test* is a way to formally answer questions about a population, based on information obtained from a sample.
In this book, we have looked at some specific hypothesis tests (see Sect.\ \@ref(Symbols)); there are hundreds more.


[Inclusion criteria](#def:InclusionExclusionCriteria)\index{Inclusion criteria!conceptual|textbf}
: *Inclusion criteria* are characteristics that individuals must meet explicitly to be included in the study. 
 
 
[Independence](#def:Independence)\index{Independence|textbf}
: Two events are *independent* if the probability of one event doesn't change depending on whether or not other event has happened.


[Individuals](#def:Population)\index{Individuals|textbf}
: The individual units in the population from which the observations of interest could be made; the [*units of analysis*](#def:UnitOfAnalysis).
Also called *cases*, or *subjects* when the individuals are people.


[Internal validity](#def:InternalValidity)\index{Internal validity|textbf}
: *Internally validity* refers to the extent to which a cause-and-effect relationship can be established in a study.
A study with *high* internal validity shows that the changes in the response variable can be attributed to changes in the explanatory variables; other explanations have been ruled out.

 
[Intervention](#def:Intervention)\index{Intervention|textbf}
: An intervention is present when *researchers* can manipulate (or impose) the values of the *explanatory variable* on the individuals to determine the impact on the response variable.



[IQR](#def:IQR)\index{Inter-quartile range (IQR)|textbf}
: The *IQR* is the range in which the middle $50$ of the data lie;
the difference between the third and the first quartiles.


[IQR rule for identifying outliers](#def:IQRRuleForIdentifyingOutliers)\index{Outliers!IQR rule|textbf}
: The IQR rule can identify outliers as either *extreme* (observations $3\times\text{IQR}$ more unusual than $Q_1$ or $Q_3$) or *mild* (observations $1.5\times \text{IQR}$ more unusual than $Q_1$ or $Q_3$, that are not extreme outliers).


[Jittering](#exm:DotsChartsQuant2)\index{Overplotting!jittering|textbf}
: *Jittering* is when a small amount of randomness is added in either the horizontal or vertical direction (or sometimes both) to separate points that would otherwise be overplotted.


[Judgement sample](#NonRandomSamples)\index{Sampling!judgement sample|textbf}
: A sample where individuals are selected, based on the researchers' judgement, depending on whether the researcher thinks they are likely to be agreeable or helpful.
  

[Levels](#def:Levels) of a qualitative variable\index{Levels|textbf}\index{Qualitative data!levels|textbf}
: The *levels* (or the *values*) of a qualitative variable refer to the names of the distinct categories. 



[Lurking variable](#def:LurkingVariable)\index{Variables!lurking|textbf}
: A *lurking variable* is an extraneous variable associated with the response *and* explanatory variables (that is, is a *confounding* variable), but whose values *are not* recorded in the study data.


[Mean](#def:Mean)\index{Mean|textbf}
: The *mean* is one way to measure the 'average' value of quantitative data.
The *arithmetic mean* is the 'balance point' of the data, and the value such that the positive and negative distances from the mean add to zero.


[Median](#def:Median)\index{Median|textbf}
: The *median* is one way to measure the 'average' value of some data.
A *median* is a value such that half the values are larger than the median, and half the values are smaller than the median.


[Mode](#def:Mode)\index{Mode|textbf}
: A *mode* is one way to measure the 'average' value of some qualitative data.
A *mode* is the level (or levels) of a qualitative variable with the most observations.


[Multi-stage sampling](#MultistageSampling)\index{Sampling!multistage|textbf}
: A sample where large groups are selected using a *simple random sample*, then smaller groups within those large groups are selected using a *simple random sample*.
The simple randomly sampling can continue for as many levels as necessary.


[Nominal variable](#def:Nominal)\index{Variables!ordinal|textbf}\index{Data!nominal|textbf}
: A *nominal* qualitative variable is a qualitative variable where the levels *do not* have a natural order. 


[Normal distribution](#DistributionsExample)\index{Normal distribution|textbf}
: A *normal distribution* is symmetrical distribution, with most values in the centre of the distribution.
The normal distribution is described by its *mean* and *standard deviation*.
A picture of a normal distribution is shown in Fig.\ \@ref(fig:DefNormalDistribution).
Normal distributions are also called *bell-shaped* distributions.

```{r DefNormalDistribution, fig.cap="A normal distribution", fig.align="center", fig.width=5, fig.height=2}
par( mar = c(4, 1, 1, 1) + 0.1)
zz <- seq(-4, 4,
          length = 100)
yy <- dnorm(zz)

plot( yy ~ zz,
   col = ifelse( knitr::is_latex_output(), "grey", "blue"),
   axes = FALSE,
   type = "l",
   xlab = expression(italic(z)-score),
   ylab = "",
   lwd = 2)
axis(side = 1)
abline(h = 0,
       col = "grey")
abline(v = 0,
       col = "grey")
```


[Null hypothesis](#def:NullHypothesis)\index{Hypotheses!null|textbf}
: The *null hypothesis* proposes that any difference, change or relationship observed in the sample can be explained by sampling variation (that is, no difference, change or relationship exists the *population*).


[Observational study](#def:ObservationalStudy)\index{Study types!observational|textbf}
: *Observational studies* study relationships *without* an intervention.


[Observer effect](#def:ObserverEffect)\index{Observer effect|textbf}
: The *observer effect* occurs when the researchers (unconsciously) change their behaviour to conform to expectations because they know what values of the explanatory variable apply to the individuals.
This may then cause the *individuals* to change their behaviour or reporting also.


[Odds](#def:Odds)\index{Odds|textbf}
: The *odds* of some event is the proportion (or percentage, or number) of times that an event happens, divided by the proportion (or percentage, or number) of times that the event does *not* happen.


[Odds ratio](#def:OddsRatio)\index{Odds ratio|textbf}
: The *odds ratio* is how many *times* greater the odds of an event are in one group, compared to the odds of the *same* event in a *different* group.
 

[Operational definition](#def:OperationalDefinition)\index{Definitions!operational|textbf}
: An *operational definition* articulates exactly *how* something will be identified, measured, observed or assessed.


[Ordinal variable](#def:Ordinal)\index{Variables!ordinal|textbf}\index{Data!nominal|textbf}
: An *ordinal* qualitative variable is a qualitative variable where the levels *do* have a natural order. 


[Outcome](#def:Outcome)\index{Outcome|textbf}
: The *outcome* in a RQ is the result, output, consequence or effect of interest in a study, numerically summarised for a group.  


[Outliers](#def:Outliers)\index{Outliers|textbf}
: An *outlier* is an observation that is 'unusual' compared to the bulk of the data (either larger or smaller).
Rules for identifying outliers are arbitrary.



[$P$-value](#def:Pvalue)\index{P@$P$-values|textbf}
: A $P$-value is the probability of observing the sample results (or something even more extreme) over repeated sampling, under the assumption that the null hypothesis about the population is true.



[Parameter](#def:Parameter)\index{Parameter|textbf}
: A *parameter* is a number, usually unknown, describing some feature of a population, and estimated by a [statistic](#def:Statistic).



[Paired data](#def:PairedData)\index{Paired data|textbf}
: *Paired data* occurs when the outcome is compared for two different, distinct situations for each individual.



[Percentage](#def:Percentage)\index{Percentages|textbf}
: A *percentage* is a [proportion](#def:Proportion), multiplied by $100$.
In this context, percentages are numbers between $0$% and $100$%.


[Percentiles](#def:Percentiles)\index{Percentiles|textbf}
: The $p$th percentile of the data is a value separating the smallest $p$% of the data from the rest.


[Pilot study](#def:PilotStudy)\index{Pilot study|textbf}
: A *pilot study* is a small test run of the study protocol used to check that the protocol is appropriate and practical, and to identify (and hence fix) possible problems with the research design or protocol.


[Placebo](#def:Placebo)\index{Placebo|textbf}
: A *placebo* is a treatment with no intended effect or active ingredient, but appears to be the real treatment.


[Placebo effect](#def:PlaceboEffect)\index{Placebo effect|textbf}
: The *placebo effect* occurs when individuals report perceived or actual effects despite not receiving an active treatment or condition, in experimental studies (Sect.\ \@ref(PlaceboEffect)).



[Plagiarism](#Referencing)\index{Plagiarism|textbf}
: *Plagiarism* is using other people’s ideas and research to develop new conclusions, or confirm existing conclusion. 
All sources used when writing research should be acknowledged, otherwise you are committing plagiarism. 
Plagiarism can be deliberate or accidental.


[Population](#def:Population)\index{Population|textbf}
: A *population* is a group of individuals (or cases, or subjects if the individuals are people) from which the total set of observations of interest could be made, and to which the results will (hopefully) generalise.


[Precision](#PrecisionAccuracy)\index{Precision|textbf}
: *Precision* refers to how similar the sample estimates from different samples are likely to be to each other (that is, how much variation is likely in the sample estimates).


[Probability](#Probability)\index{Probability|textbf}
: A *probability* is a number between $0$ and $1$ inclusive (or between $0$% and $100$% inclusive) that quantifies the likelihood that a certain *event* will occur.
A probability of zero (or $0$%) means the event is 'impossible' (will *never* occur), and a probability of one (or $100$%) means that the event is *certain* to happen (will always occur).
Most events have a probability between the extremes of $0$% and $100$%.


[Proportion](#def:Proportion)\index{Proportions|textbf}
: A *proportion* is a fraction out of a total.
Proportions are numbers between $0$ and $1$.


[Protocol](#def:Protocol)\index{Protocol|textbf}
: A *protocol* is a predefined procedure detailing the design and implementation of studies, and for data collection. 


[Qualitative data](#def:QualitativeData)\index{Qualitative data|textbf}
: *Qualitative data* is not *mathematically* numerical data: it comprises mutually exclusive, exhaustive  categories or labels (even if those labels are numbers). 
Also called *categorical* data.


[Quantitative data](#def:QuantitativeData)\index{Quantitative data|textbf}
: *Quantitative data* is *mathematically* numerical data: the numbers have numerical meaning, and represent quantities or amounts.
Quantitative data generally arise from counting or measuring.


[Quantitative research](#def:QualitativeResearch)\index{Research!quantitative|textbf}
: *Quantitative research* summarises and analyses data using numerical methods, such as producing averages and percentages. 

      
[Quartiles](#def:Quartiles)\index{Quartiles|textbf}
: *Quartiles* describe the variation and shape of data.
  The first quartile $Q_1$ is a value that separates the smallest $25$% of observations from the largest $75$%; it is like the median of the *smaller* half of the data, halfway between the minimum value and the median.
  The second quartile $Q_2$ is a value that separates the smallest $50$% of observations from the largest $50$% (and is the *median*).
  The third quartile $Q_3$ is a value that separates the smallest $75$% of observations from the largest $25$%; it is like the median of the *larger* half of the data, halfway between the median and the maximum value.


[Quasi-experiment](#def:QuasiExperiment)\index{Study types!experimental!quasi|textbf}
: In a *quasi-experiment*, the researchers (1)\ allocate treatments to groups of individuals (i.e., allocate the values of the explanatory variable to the individuals), but (2)\ do not determine who or what is in those groups.


[Questionnaire](#def:Questionnaire)\index{Questionnaire|textbf}
: A questionnaire is a set of questions for respondents to answer.


[Random](#def:Random)\index{Random|textbf}
: *Random* means 'determined completely by impersonal chance'.


[Random procedure](#def:RandomProcedure)\index{Random procedure|textbf}
: A *random procedure* is a sequence of well-defined steps that (a)\ can be repeated, in theory, indefinitely under essentially identical conditions; (b)\ has well-defined results; and (c)\ has result that are unpredictable from any individual repetition.



[Range](#def:Range)\index{Range|textbf}
: The *range* is the maximum value of a variable minus the minimum value of the variable. 


[Relational research question](#def:RelationalRQ)\index{Relational RQ|textbf}
: Relational RQs have a population, outcome, and a *between*-individuals comparison.


[Relative frequency approach to probability](#def:RelativeFrequencyApproachToProbability)\index{Probability!relative frequency approach|textbf}
: In the *relative frequency approach to probability*, the probability of an event is approximately the number of times the outcomes of interest has appeared in the past, divided by the number of 'attempts' in the past.
This produces an *approximate* probability.


[Repeated-measures research question](#def:RepeatedMeasuresRQ)\index{Repeated-measures RQ|textbf}
: Repeated-measures RQs have a population, outcome and a *within*-individuals comparison.


[Representative samples](#Representative-samples)
: A representative sample is one where the individuals *in* the sample are not likely to be different the individuals *not in* the sample, at least for the variables of interest.




[Research design](#def:StudyDesign)\index{Research design|textbf}
: *Research design* refers to the decisions made by the researchers to maximise *external validity* and *internal validity*.


[Response variable](#def:ResponseVariable)\index{Response variable|textbf}
: A *response variable* records the result, output, consequence or effect of interest from changes in another variable (the explanatory variable).

 
[Sample](#def:Sample)\index{Sample|textbf}
: A *sample* is a subset of individuals from the population from which data are collected.


[Sample space](#def:SampleSpace)\index{Sample space|textbf}
: The *sample space* is a list of all possible and distinct results after administering a random procedure once. 


[Sampling distribution](#def:SamplingDistribution)\index{Sampling distribution|textbf}
: A *sampling distribution* is the distribution of a statistic, showing how its value varies in all possible samples.


[Sampling frame](#def:SamplingFrame)\index{Sampling frame|textbf}
: The *sampling frame* is a list of all the individuals in the population. 



[Sampling mean](#def:SamplingMean)\index{Sampling mean|textbf}
: The *sampling mean* is the mean of the sampling distribution of a statistic.



[Sampling variation](#def:SamplingVariation)\index{Sampling variation|textbf}
:  *Sampling variation* refers to how the sample estimates (statistics) vary from sample to sample, because every possible sample is different.


[Selection bias](#def:SelectionBias)\index{Selection bias|textbf}
: *Selection bias* is the tendency of a sample to over- or under-estimate a population quantity.


[Scale data](#def:QualitativeData)
: See *quantitative data*.


[Simple event](#def:SimpleEvent)\index{Simple event|textbf}
: A *simple event* is a single element of the sample space. 



[Simple random sample](#def:SamplingSRS)\index{Simple random sample|textbf}
: A sample where *every* possible sample of a given size has *same* chance of being selected.



[Stacking](#exm:DotsChartsQuant2)\index{Overplotting!stacking|textbf}
: *Stacking* is when points are plotted above other points with similar values, to separate points that would otherwise be overplotted.


[Standard deviation](#def:StandardDeviation)\index{Standard deviation|textbf}
: The *standard deviation* is, approximately, the mean distance of observations from the mean.


[Standard deviation rule for identifying outliers](#def:StandardDeviationRuleForIdentifyingOutliers)
: For approximately symmetric distributions, any observation more than three standard deviations from the mean can be considered an outlier.


[Standard error](#def:StandardError)\index{Standard error|textbf}
: A *standard error* is the standard deviation of all possible values of the sample estimate (from samples of a certain size). 
Any quantity estimated from a sample has a standard error.


   
[Stratified sampling](#StratifiedSampling)\index{Stratified sample|textbf}
: A sample where the population is split into a small number of large (usually homogeneous) groups called *strata*, then cases are selected using a *simple random sample* from *each* stratum.



[Statistic](#def:Statistic)\index{Statistic|textbf}
: A *statistic* is a number describing some feature of a sample (to estimate the unknown value of the population [parameter](#def:Parameter)).



[Statistical validity](#def:StatisticalValidity)\index{Statistical validity|textbf}
: A result is *statistically valid* if the conditions for the underlying mathematical calculations and assumptions to be approximately correct are met, such as the sampling distribution having an approximate normal distribution.
Every confidence interval and hypothesis test has statistical validity conditions.


[Subjective approach to probability](#def:SubjectiveApproachToProbability)\index{Probability!subjective approach|textbf}
: In the *subjective approach to probability*, various factors are incorporated, perhaps subjectively, to determine the probability of an event.



[Subjects](#def:UnitOfAnalysis)\index{Subjects|textbf}
: The individual units in the population when the units are people; the [*units of analysis*](#def:UnitOfAnalysis).
Also called *individuals* or *cases*; however, those two terms do not refer to people.



[Systematic sampling](#SystematicSampling)\index{Systematic sampling|textbf}
: A sample where the first case is *randomly* selected; then, every $n$th individual is selected.


 
[$t$-score](#TestStatistic)\index{t@$t$-scores|textbf}
: A *$t$-score* measures how many standard deviations a value is from the mean.
A $t$-score is like a $z$-score.
Also see *$z$-score*.


[Treatments](#def:Treatments)\index{Treatments|textbf}
: The *treatments* are the values of the explanatory variable that the researchers impose upon the individuals in the *experimental* study.


[True experiment](#def:TrueExperiment)\index{Study types!experimental!true|textbf}
: In a true experiment, the researchers (1)\ allocate treatments to groups of individuals (i.e., values of the explanatory variable to the individuals), *and* and (2)\ determine who or what is in those groups.
While the steps may not happen *explicit*, they happen *conceptually*.


[Unit of observation](#def:UnitOfObservation)\index{Units of observation|textbf}
: The *unit of observation* is the 'who' or 'what' that are observed, from which measurements are taken and data collected.


[Unit of analysis](#def:UnitOfAnalysis)\index{Units of analysis|textbf}
: The smallest collection of units of observations (and perhaps the units of observations themselves) about which conclusions are made; the smallest *independent* elements of the population for which information is analysed.
In an *experimental study*, the unit of analysis is the smallest collection of units of observations that can be randomly allocated to separate treatments.


[Unstandardizing formula](#def:UnstandardizingFormula)\index{Unstandardizing formula|textbf}
: When the $z$-score is known, the *unstandardising formula* determines the corresponding value of the observation $x$.
 

[Values](#def:Levels) of a qualitative variable
: See *Levels*. 


[Variable](#def:Variable)\index{Variables|textbf}
: A *variable* is a single aspect or characteristic associated with the individuals, whose values can vary from individual to individual.


[Voluntary response (self-selecting) sample](#NonRandomSamples)\index{Voluntary response sample|textbf}
: A sample where individuals participate if they wish to.


[Within-individuals comparison](#def:ComparisonWithin)\index{Comparison!within individuals|textbf}
: See *Comparison (within individuals)*, *Comparison*.


 
[$z$-score](#def:zScore)\index{zscores@$z$-scores|textbf}
: A *$z$-score* measures how many standard deviations a value is from the mean.
In symbols:
$$
   z = \frac{x - \mu}{\sigma},
$$
where $x$ is the value, $\mu$ is the mean of the distribution, and $\sigma$ is the standard deviation of the distribution.
Also see $t$-score.


```{=latex}
\end{small}
```