---
title: "Scientific Research and Methodology"
subtitle: "An introduction to quantitative research and statistics"
author: "Peter K. Dunn"
date: '`r  Sys.time()`'
documentclass: krantz
knit: "bookdown::render_book"
output: pdf_document
description: "An introduction to quantitative research in science, engineering and
  health (including research design, hypothesis testing and confidence intervals in
  common situations)"
#url: 'https\://bookdown.org/pkaldunn/SRM-Textbook/'
bibliography: [packages.bib, ReferenceList.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
#lot: yes
#lof: yes
always_allow_html: yes
links-as-notes: yes
site: bookdown::bookdown_site
github-repo: PeterKDunn/SRM-Textbook
fontsize: 12pt
monofont: "Source Code Pro"
monofontoptions: "Scale=0.7"
geometry: margin=27mm
graphics: yes
nocite: |
  @data:Davidson2000:Amputees, @data:Grabosky2016:Trees, @data:Pons:ResponseTime, @data:Edwards2015:PA
---







```{r echo=FALSE, results='asis', results='hide'}
knit_hooks$set(document = 
    function(x) {sub('\\usepackage[]{color}', 
                     '\\usepackage{xcolor}', 
                      x,
		      fixed = TRUE)})

# Change webexercises to show green when correct
style_widgets(correct = "green")
```

# Preface {-}


<!-- TO RENDER -->
<!-- RUN IN R: -->

<!-- # render_book("index.Rmd", clean = FALSE) -->
<!--    where the  clean = TRUE  keeps the md files (for converting to epub?) -->

<!-- # render_book("index.Rmd", "bookdown::pdf_book") -->
<!-- # render_book("index.Rmd"); render_book("index.Rmd", "bookdown::pdf_book") -->

<!--  TO PUBLISH ON BOOKDOWN:  -->
<!-- library(rsconnect) -->
<!-- bookdown::publish_book() -->
<!-- WAS: publish_book(name="Book"); publish_book() -->


<!-- # It seems that figures default to:  -->
<!-- #     fig.width=7 (inches) and fig.height = 5 (inches) -->
<!-- #  -->


This book introduces quantitative research in the scientific and health disciplines.
The whole research process is introduced, from asking a research question to analysis and reporting of the data.
The main focus, however, is the analysis of data.



## Supporting documents {-}

To support this textbook, the following are also available:

* A [Tutorial Book](https://bookdown.org/pkaldunn/SRM-tutorials/); and
* A [Software Support Book](https://bookdown.org/pkaldunn/SRM-software/) with brief instruction for using jamovi and SPSS to perform *most* of the analyses in this book.

These books are both freely available online.



## The data sets used in this book {-}

Almost every data set used in this book is a real data set.
Many are available electronically so that you can download them and work with them in statistical software.
`r if (knitr::is_latex_output()) {
   'These are provided in the online version of this book (Appendix \\@ref(AppendixDataSets)).'
} else {
   'These are listed in Appendix \\@ref(AppendixDataSets).'
}`


## Statistical software {-}

Most of this book can be read without relying on any specific statistical software.
However, sometimes [jamovi](https://www.jamovi.org/) [@Software:jamovi] and/or SPSS [@Software:SPSS] output are shown.

jamovi is **free** to download and use.
While jamovi looks a little like a spreadsheet (as does SPSS), underneath it uses the **R** statistical package [@Software:Rsoftware].
From the [jamovi homepage (sic)](https://www.jamovi.org/):



<!-- ## How to use this book {-} -->
```{r, child = if (knitr::is_html_output(exclude = "epub")) 'UseHTML-bs4.Rmd'}
```



## Call-outs used on this book {-}

The following call-outs are used in this book:

::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}
These chunks introduce the objectives for the chapters of the book.
:::


:::: {.importantBox .important data-latex="{iconmonstr-warning-8-240.png}"}
These chunks highlight common mistakes or warnings, about a particular concept or about using a formula.
:::


::: {.tipBox .tip data-latex="{iconmonstr-info-6-240.png}"}
These chunks offer helpful information.
:::


::: {.softwareBox .software data-latex="{iconmonstr-laptop-4-240.png}"}
These chunks refer to text relevant to using software (jamovi or SPSS) or a calculator.
:::


::: {.pronounceBox .pronounce data-latex="{iconmonstr-microphone-7-240.png}"}
These chunks indicate how certain symbols and terms are pronounced.
:::


::: {.thinkBox .think data-latex="{iconmonstr-light-bulb-2-240.png}"}
These chunks give you some questions to think about.
:::

`r if (knitr::is_html_output()) '<!--'`
::: {.QRanswerBox .QRanswer data-latex="{iconmonstr-check-mark-14-240.png}"}
These chunks provide answers to the *Quick Revision* questions.
:::
`r if (knitr::is_html_output()) '-->'`




## Who can use this book? {-}

The online textbook is **free** for anyone to use:
There is no charge for students, instructors or institutions.

Although not essential, an email to the author (explaining how the textbook is being used, who is using the textbook, and your thoughts on the textbook) would be appreciated: `pdunn2 <at> usc.edu.au`.


## How this book was made {-}

This book was made using **R** [@Software:Rsoftware], and the **bookdown** package [@Software:Rbookdown], which is based on [Markdown](https://en.wikipedia.org/wiki/Markdown) syntax, using **knitr** [@package:knitr].

Numerous other **R** packages were used too:

* The diagrams were made in base **R**,
  or using the **diagram** package [@Rpackage:diagram].
* The animations in the html version were made using the **animation** package [@package:animation], 
  and the stills for the PDF version captured by the **webshot** package [@package:webshot].
* The **gifski** package was used to create animations in the online version [@package:gifski].
* The **kableExtra** package was used for nicer tables [@package:kableExtra].
* The maps of Australia were generated using the **oz** package [@package:oz] and plotted using the **ggplot2** package [@package:ggplot2].
* The display of some data tables in the online version use the **DT** package [@package:DT].
* Some plots use the **plotrix** package [@package:plotrix].
* The NHANES data is provided by the **NHANES** package [@data:NHANES:Rpackage].
* Some data are from the **GLMsData** package [@computing:GLMsData:Rpackage].
* The **scales** package is used to rescale data [@package:scales].
* The carousels in the online version (for example, Sect. \@ref(UnderstandingScatterplots)) are made using the **slickR** package [@package:slickR].
* The **dygraphs** package [@package:dygraphs] is used to make interactive graphs (for example, Sect. \@ref(TimePlots)) for the online version.
* The **plotly** package [@package:plotly] is used to make some interactive graphs (for example, in Sect. \@ref(Histograms)) for the online version.
* The **dplyr** package is used in some data manipulations [@package:dplyr].
* The **viridis** package is used for some colour specifications [@package:viridis] to make colours easier for those with colour-blindess to distinguish colours, and for better grey scale printing.
* The **webexercises** package was used to create the interactive web exercises [@package:webex].
* The **plotfunctions** package was used to add images to existing plots [@package:plotfunctions].
* The **qrcode** package was used to generate QR-codes for some activities so that those using the hard-copy version can use activities that appear in the online version [@package:qrcode].

All of this software is *free* and open source.

Other resources used include:

* Some activities are created using [LearningApps.org](http://LearningApps.org), which is free to use.
* Icons are from [**iconmonstr**](https://iconmonstr.com/) and are freely available.
* The images of the cards used in Sect. \@ref(NeedForDecisionMaking) are from https://code.google.com/archive/p/vector-playing-cards/, and are in the public domain.
* The text folding (in the html version) was implemented by adapting advice from [StackOverflow](https://stackoverflow.com/questions/53810294/folding-general-text-in-r-bookdown).
* The images of dice used in, for example, Sect. \@ref(ProbClassical), are from https://pixabay.com/vectors/dice-games-game-six-sided-face-26772/, and are free.
* The images used in the online book are free (and used according to their guidelines), as listed in Appendix \@ref(Image-Credits).

Special thanks to **Dr Amanda Shaker** (La Trobe University), who has advised of numerous issues (with the corrections).
 


## Learning Outcomes {-}

::: {.objectivesBox .objectives data-latex="{iconmonstr-target-4-240.png}"}
In this book, you will learn to:

* Develop quantitative research questions and testable hypotheses.
* Design quantitative studies to answer simple quantitative research questions.
* Select and produce appropriate graphical, numerical and statistical analyses.
* Select, apply and interpret the results of the correct statistical technique to analyse data.
* Comprehend, apply and communicate in the language of research and statistics.
* Demonstrate professional integrity in planning, interpreting and reporting the results of quantitative studies.
:::






## How to cite this book {-}

Peter K. Dunn (2022). *Scientific Research and Methodology: An introduction to quantitative research in science and health*.
https://bookdown.org/pkaldunn/SRM-Textbook

The [CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/) licence is applied to this textbook.


```{block2, type='flushright', html.tag='p'}
Peter K. Dunn  
Sippy Downs, Australia
```
