\begin{cols}
\begin{col}{0.63\textwidth}

Using the classical approach to probability (Sect. \ref{ProbClassical}), the probability of the ball landing on an \emph{odd} number 
(an '\emph{odd-spin}') is $p = 18/37 =  0.486$.
This is the \emph{population proportion}.
\medskip

If the wheel is spun (say) 15 times, the \emph{sample} proportion of odd-spins in those 15 spins, denoted $\hat{p}$, will vary.
(Of course, the \emph{sample} proportion $\hat{p}$ of odd-spins can vary after spinning the wheel 30, 50 or 100 times also.)
\medskip

But, \emph{how} does $\hat{p}$ vary from one set of 15 spins to another set of 15 spins? 
Can we describe \emph{how} the value of $\hat{p}$ varies?
\end{col}
\begin{col}{0.07\textwidth}
\null\ \null
\end{col}
\begin{col}{0.30\textwidth}

```{r echo=FALSE, results='hide', cache = FALSE, out.width='100%'}
date.str <- date()
seed <- as.numeric( paste( substr(date.str, 18, 19), # Second
                           substr(date.str, 15, 16),
                           sep = "") # Minutes
                       )
# For Roulette wheel
```
```{r RouletteWheel, echo=FALSE, cache=FALSE, out.width = "100%", fig.height=2.5,fig.width=2.25}
# Wheel  
DD <- 10
dd <- 4

spin <- plotSpinningWheel( angle = deg2rad(360/37)/2, 
                           radius = dd, 
                           wheelSize = c(DD, dd),
                           plotShadow = FALSE,  
                           main = "")
```
\end{col}
\end{cols}

