
<!-- ```{r echo=FALSE, fig.width=2.2, fig.height=0.9, fig.align="center"} -->
<!-- par(mar=c(2.1, 0.05, 0.05, 0.05)) -->
<!-- delta <- 0.01 -->
<!-- z <- seq(-3.5, 3.5, by=delta) -->
<!-- zy <- dnorm(z) -->

<!-- zval <- 1.875 -->

<!-- plot( zy ~ z, -->
<!--       lwd=2, -->
<!--       type="l", -->
<!--       xlab="", -->
<!--       ylab="", -->
<!--       axes=FALSE) -->
<!-- #axis(side=1, at=-3:3) -->
<!-- axis(side=1, at=zval, label=expression(italic(z))) -->
<!-- abline(h=0, lwd=2) -->

<!-- abline(v=zval, lwd=2, col="grey") -->

<!-- xshade <- c(  seq(min(z), zval, by=delta), -->
<!--               seq(zval, min(z), by=-delta) ) -->
<!-- yshade <- c( rep(0, length(xshade)/2), -->
<!--              rev(zy[ z<zval]) ) -->

<!-- polygon( xshade, yshade, col="wheat") -->

<!-- arrows(-2.5, max(zy)/2, -->
<!--        0, max(zy)/2,  -->
<!--        angle=15, -->
<!--        length=0.1) -->
<!-- text(-2.5, max(zy)/2,  -->
<!--      "Area", -->
<!--      pos=3) -->
<!-- ``` -->

<!-- \small The table gives the probability (area) that a $z$-score is **less** than the $z$-score looked up. -->
<!-- For example:  -->
<!-- Look up $z = 1.87$; the area *less than* $z=1.87$ is about 0.9693, -->
<!-- or about 96.9%. -->

<!-- ```{r AppendixTables, echo=FALSE} -->
<!-- z1 <- c( z1n <- seq(-3.5, 0, by=0.1), z1p <- seq(0, 3.5, by=0.1))  # As zero gets two rows: pos and neg -->
<!-- z2 <- seq(0.00, 0.09, by=0.01) -->

<!-- Ztab <- array( dim=c(length(z1), length(z2)) )   -->

<!-- colnames(Ztab) <- sprintf("%.2f", z2 ) -->
<!-- rownames(Ztab) <- sprintf("%.1f", z1) -->
<!-- rownames(Ztab)[length(z1n)] <- "-0.0" -->
<!-- rownames(Ztab)[length(z1n)+1] <- "0.0" -->

<!-- zero.count <- 0 -->
<!-- for (i in 1:length(z1)){ -->

<!--   if ( z1[i] > 0 ){ -->
<!--     Ztab[i, ] <- pnorm( z1[i] + z2 ) -->
<!--   }  -->
<!--   if ( z1[i] == 0 ){ -->
<!--     if (zero.count == 0 ){ # The first zero row: negative 0.00 -->
<!--       zero.count <- zero.count + 1  -->
<!--       Ztab[i, ] <- pnorm( z1[i] - z2 ) -->

<!--     } else {               # The second zero row: positive 0.00 -->
<!--       Ztab[i, ] <- pnorm( z1[i] + z2 ) -->
<!--     } -->
<!--   } -->
<!--   if ( z1[i] < 0 ){ -->
<!--     Ztab[i, ] <- pnorm( z1[i] - z2 ) -->
<!--   } -->
<!-- } -->

<!-- #if( knitr:::is_html_output() ) { -->
<!-- #  kable(Ztab, -->
<!-- #        "html", -->
<!-- #      caption="The probability that the area is **less than** the value of $z$ that is looked up", -->
<!-- #      digits=4, -->
<!-- #      align="r", -->
<!-- #      booktab=TRUE) %>% -->
<!-- #  kable_styling() %>% -->
<!-- #  kableExtra::scroll_box(width = "100%", height = "400px") -->
<!-- #} -->
<!-- if( knitr:::is_latex_output() ) { -->
<!--    kable(Ztab, -->
<!--         format="latex", -->
<!--       caption="The probability that the area is less than the value of $z$ that is looked up", -->
<!--       digits=4, -->
<!--       longtable=TRUE, -->
<!--       align="r", -->
<!--       booktab=TRUE) %>% -->
<!--       kable_styling(font_size=9, latex_options="repeat_header") -->
<!-- } -->
<!-- ``` -->



\pagebreak
\null
\pagebreak

## Normal distribution: negative $z$-values probabilities {#ZTablesNEG}

```{r echo=FALSE, fig.width=2.5, fig.height=1, fig.align="center"}
par(mar=c(2.1, 0.05, 0.05, 0.05))
delta <- 0.01
z <- seq(-3.5, 3.5, by=delta)
zy <- dnorm(z)

zval <- 1.875

plot( zy ~ z,
      lwd=2,
      type="l",
      xlab="",
      ylab="",
      axes=FALSE)
#axis(side=1, at=-3:3)
axis(side=1, at=zval, label=expression(italic(z)))
abline(h=0, lwd=2)

abline(v=zval, lwd=2, col="grey")

xshade <- c(  seq(min(z), zval, by=delta),
              seq(zval, min(z), by=-delta) )
yshade <- c( rep(0, length(xshade)/2),
             rev(zy[ z<zval]) )
            
polygon( xshade, yshade, 
         col = plot.colour)

arrows(-2.5, max(zy)/2,
       0, max(zy)/2, 
       angle=15,
       length=0.1)
text(-2.5, max(zy)/2, 
     "Area",
     pos=3)
```




\begin{small}
The table gives the probability (area) that a $z$-score is \textbf{less} than the $z$-score looked up.
For example: 
Look up $z = -1.87$; the area \textit{less than} $z=-1.87$ is about 0.0307,
or about 3.07\%.
\end{small}

```{r echo=FALSE}
z1 <- c( z1n <- seq(-3.5, 0, by=0.1), z1p <- seq(0, 3.5, by=0.1))  # As zero gets two rows: pos and neg
z2 <- seq(0.00, 0.09, by=0.01)

Ztab <- array( dim=c(length(z1), length(z2)) )  

colnames(Ztab) <- sprintf("%.2f", z2 )
rownames(Ztab) <- paste("$", sprintf("%.1f", z1), "$")
rownames(Ztab)[length(z1n)] <- "$-0.0$"
rownames(Ztab)[length(z1n)+1] <- "$0.0$"

zero.count <- 0
for (i in 1:length(z1)){
  
  if ( z1[i] > 0 ){
    Ztab[i, ] <- pnorm( z1[i] + z2 )
  } 
  if ( z1[i] == 0 ){
    if (zero.count == 0 ){ # The first zero row: negative 0.00
      zero.count <- zero.count + 1 
      Ztab[i, ] <- pnorm( z1[i] - z2 )
      
    } else {               # The second zero row: positive 0.00
      Ztab[i, ] <- pnorm( z1[i] + z2 )
    }
  }
  if ( z1[i] < 0 ){
    Ztab[i, ] <- pnorm( z1[i] - z2 )
  }
}

#if( knitr:::is_html_output() ) {
#  kable(Ztab,
#        "html",
#      caption="The probability that the area is **less than** the value of $z$ that is looked up",
#      digits=4,
#      align="r",
#      booktab=TRUE) %>%
#  kable_styling() %>%
#  kableExtra::scroll_box(width = "100%", height = "400px")
#}
```

```{r AppendixTablesNegativeZ, echo=FALSE}

if( knitr:::is_latex_output() ) {
   kable(Ztab[1: length(z1n), ],
        format="latex",
      caption="The probability that the area is less than the value of $z$ that is looked up",
      digits=4,
      longtable=TRUE,
      escape=FALSE,
      #align="r",
      booktab=TRUE) %>%
      kable_styling(font_size=9)        
}
```





\pagebreak


```{r echo=FALSE, fig.width=2.2, fig.height=1, fig.align="center"}
par(mar=c(2.1, 0.05, 0.05, 0.05))
delta <- 0.01
z <- seq(-3.5, 3.5, by=delta)
zy <- dnorm(z)

zval <- 1.875

plot( zy ~ z,
      lwd = 2,
      type = "l",
      xlab = "",
      ylab = "",
      axes = FALSE)
#axis(side=1, at=-3:3)
axis(side = 1, 
     at = zval, 
     label = expression( italic(z) ) )
abline(h = 0, 
       lwd = 2)

abline(v = zval, 
       lwd = 2, 
       col = "grey")

xshade <- c(  seq(min(z), zval, 
                  by = delta),
              seq(zval, min(z), 
                  by = -delta) )
yshade <- c( rep(0, length(xshade)/2),
             rev(zy[ z<zval]) )
            
polygon( xshade, yshade, 
         col = plot.colour)

arrows(-2.5, max(zy)/2,
       0, max(zy)/2, 
       angle = 15,
       length = 0.1)
text(-2.5, max(zy)/2, 
     "Area",
     pos = 3)
```


## Normal distribution: positive $z$-values probabilities  {#ZTablesPOS}


\begin{small}
The table gives the probability (area) that a $z$-score is \textbf{less} than the $z$-score looked up.
For example: 
Look up $z = 1.87$; the area \textit{less than} $z=1.87$ is about 0.9693,
or about 96.9\%.
\end{small}

```{r AppendixTablesPositiveZ, echo=FALSE}
if( knitr:::is_latex_output() ) {
   kable(Ztab[ (length(z1n)+1): length(z1), ],
        format="latex",
      caption="The probability that the area is less than the value of $z$ that is looked up",
      digits=4,
      longtable=TRUE,
      escape=FALSE,
      align="r",
      booktab=TRUE) %>%
      kable_styling(font_size=9)
}
```




