## When the $z$-score is known, and the area is sought {#ZTablesOnline}

```{r echo=FALSE, fig.width=2.25, fig.height=1.25, fig.align="center"}
par(mar=c(2.1, 0.5, 0.5, 0.5))
delta <- 0.01
z <- seq(-3.5, 3.5, by=delta)
zy <- dnorm(z)

zval <- 1.875

plot( zy ~ z,
      lwd = 2,
      type = "l",
      xlab = "",
      ylab = "",
      axes = FALSE)
#axis(side=1, at=-3:3)
axis(side = 1, 
     at = zval, 
     label = expression( italic(z) ) )
abline(h = 0, 
       lwd = 2)

abline(v = zval, 
       lwd = 2, 
       col = "grey")

xshade <- c(  seq(min(z), zval, by =  delta),
              seq(zval, min(z), by = -delta) )
yshade <- c( rep(0, length(xshade)/2),
             rev(zy[ z<zval]) )
            
polygon( xshade, yshade, 
         col = plot.colour)

arrows(-2.5, max(zy)/2,
       0, max(zy)/2, 
       angle=15,
       length=0.1)
text(-2.5, max(zy)/2, 
     "Area",
     pos = 3)
```

The table gives the probability (area) that a $z$-score is **less** than the $z$-score looked up.
For example: 
Look up $z = -1.87$; the area *less than* $z = -1.87$ is about 0.0307,
or about 3.1%.

To use this table,
enter the $z$-score in the search box *under* the `z.score` column. 
The area will be shown.
The table includes $z$-values between -4 and 4.
(Alternatively, you can search through the table manually.)

```{r echo=FALSE, fig.width=2.5, fig.height=1.5, fig.align="center"}
zall <- round(seq(-4, 4, by=0.01) , 2) # As zero gets two rows: pos and neg
znorm <- round( pnorm(zall), 4)

ZtabAll <- data.frame(
  "z score" = sprintf("%.2f", zall),
  "Area to left" = as.numeric(format( round(znorm, 4), nsmall=4) )
)

#colnames(ZtabAll) <- c("z-score", "Area to the *left*")
#rownames(ZtabAll) <- as.character( round(zall, 2) )

datatable(ZtabAll, 
          filter = 'top', 
          rownames=FALSE,
          options = list(
            pageLength = 10, 
            autoWidth = FALSE
          )
)
```









