## When the area is known, and the $z$-score is sought {#ZTablesOnlineBackwards}


```{r echo=FALSE, fig.width=2.5, fig.height=1.5, fig.align="center"}
par(mar=c(2.1, 0.5, 0.5, 0.5))
delta <- 0.01
z <- seq(-3.5, 3.5, 
         by = delta)
zy <- dnorm(z)

zval <- 1.875

plot( zy ~ z,
      lwd = 2,
      type = "l",
      xlab = "",
      ylab = "",
      axes = FALSE)
#axis(side=1, at=-3:3)
axis(side = 1, 
     at = zval, 
     label=expression(italic(z)))
abline(h = 0, 
       lwd = 2)

abline(v = zval, 
       lwd = 2, 
       col = "grey")

xshade <- c(  seq(min(z), zval, by =  delta),
              seq(zval, min(z), by = -delta) )
yshade <- c( rep(0, length(xshade)/2),
             rev(zy[ z<zval]) )
            
polygon( xshade, 
         yshade, 
         col = plot.colour)

arrows(-2.5, 
       max(zy)/2,
       0, 
       max(zy)/2, 
       angle = 15,
       length = 0.1)
text(-2.5, 
     max(zy)/2, 
     "Area",
     pos = 3)
```

The table gives the $z$-score such that a given probability (area) is to the left of the $z$-score.
For example: 
Look up an area of 10%, and the corresponding $z$-score is $z = -1.282$; 
that is,
the area to the left of $z = 1.282$ is about 10%.

To use this table,
enter that area *to the left* in the search box *under* the `Area.to.left` column. 
The corresponding $z$-score will be shown.
(Alternatively, you can search through the table manually.)

```{r echo=FALSE, fig.width=2.5, fig.height=1.5, fig.align="center"}
AreaAll <- c(0.001, 
             seq(0.01, 0.99, by = 0.001), 
             0.999)
zscores <- qnorm(AreaAll)

ZtabAllBackwards <- data.frame(
  "Area to left" = sprintf("%.3f", AreaAll ),
  "z score" = as.numeric(format( round(zscores, 3), nsmall=3) )
)

#colnames(ZtabAll) <- c("z-score", "Area to the *left*")
#rownames(ZtabAll) <- as.character( round(zall, 2) )

datatable(ZtabAllBackwards, 
          filter = 'top', 
          rownames=FALSE,
          options = list(pageLength = 10,
                         autoWidth = FALSE
          )
)
```
