



\pagebreak
\null
\pagebreak

## Normal distribution: negative $z$-values probabilities {#ZTablesNEG}


```{r echo=FALSE, fig.width=2.5, fig.height=1, fig.align="center"}
par(mar = c(2.1, 
            0.05, 
            0.05, 
            0.05))
delta <- 0.01
z <- seq(-3.5, 3.5, 
         by = delta)
zy <- dnorm(z)

zval <- -1.385 # Add the last digit 5 so shading works. Really, z = -1.38

plot( zy ~ z,
      lwd = 2,
      type = "l",
      xlab = "",
      ylab = "",
      axes = FALSE)
#axis(side=1, at=-3:3)
axis(side = 1, 
     at = zval, 
     label = expression(italic(z)))
abline(h = 0, 
       lwd = 2)

abline(v = zval, 
       lwd = 2, 
       col = "grey")

xshade <- c(  seq(min(z), 
                  zval, 
                  by =  delta),
              seq(zval, 
                  min(z), 
                  by = -delta) )
yshade <- c( rep(0, length(xshade)/2),
             rev(zy[ z < zval]) )
            
polygon( xshade, 
         yshade, 
         col = plot.colour)

arrows(x0 = -2.5, 
       y0 = max(zy)/2,
       x1 = -1.8, 
       y1 = 0.05, 
       angle = 15,
       length = 0.1)
text(-2.5, max(zy)/2, 
     "Area",
     pos = 3)
```




\begin{small}
The table gives the probability (area) that a $z$-score is \textbf{less} than the $z$-score looked up.
For example: Look up $z = -1.38$; the area \emph{less than} $z = -1.38$ is about 0.0838, or about 8.4\%.
\end{small}

```{r echo=FALSE}
z1 <- c( z1n <- seq(-3.5, 0, 
                    by = 0.1), 
         z1p <- seq(0, 3.5, 
                    by = 0.1))  # As zero gets two rows: pos and neg
z2 <- seq(0.00, 0.09, 
          by = 0.01)

Ztab <- array( dim = c(length(z1), 
                       length(z2)) )  

colnames(Ztab) <- sprintf("%.2f", z2 )
rownames(Ztab) <- paste("$", sprintf("%.1f", z1), "$")
rownames(Ztab)[length(z1n)] <- "$-0.0$"
rownames(Ztab)[length(z1n)+1] <- "$0.0$"

zero.count <- 0
for (i in 1:length(z1)){
  
  if ( z1[i] > 0 ){
    Ztab[i, ] <- pnorm( z1[i] + z2 )
  } 
  if ( z1[i] == 0 ){
    if (zero.count == 0 ){ # The first zero row: negative 0.00
      zero.count <- zero.count + 1 
      Ztab[i, ] <- pnorm( z1[i] - z2 )
      
    } else {               # The second zero row: positive 0.00
      Ztab[i, ] <- pnorm( z1[i] + z2 )
    }
  }
  if ( z1[i] < 0 ){
    Ztab[i, ] <- pnorm( z1[i] - z2 )
  }
}

```


```{r AppendixTablesNegativeZ, echo=FALSE}
ZtabNeg <-  formatC( Ztab[1: length(z1n), ], # For small values, that otherwise end up in sci notation
                     format = "f",
                     digits = 4)   

if( knitr::is_latex_output()) {
   kable(ZtabNeg,
        format = "latex",
        caption = "The probability that the area is less than the value of $z$ that is looked up",
        digits = 4,
        longtable = TRUE,
        escape = FALSE,
        align="r",
        booktab = TRUE) %>%
      kable_styling(font_size = 9)        

} else {# knitr::pandoc_to("epub3") ) {
   kable(ZtabNeg,
        format = "html",
        caption = "The probability that the area is less than the value of $z$ that is looked up",
        digits = 4,
        longtable = TRUE,
        escape = FALSE,
        align="r",
        booktab = TRUE)        
}
```

  


\pagebreak


## Normal distribution: positive $z$-values probabilities  {#ZTablesPOS}



```{r echo=FALSE, fig.width=2.2, fig.height=1, fig.align="center"}
par(mar = c(2.1, 
            0.05, 
            0.05, 
            0.05))
delta <- 0.01
z <- seq(-3.5, 3.5, 
         by = delta)
zy <- dnorm(z)

zval <- 1.875 # Add the last digit 5 so shading works. Really, z = 1.87

plot( zy ~ z,
      lwd = 2,
      type = "l",
      xlab = "",
      ylab = "",
      axes = FALSE)
#axis(side=1, at=-3:3)
axis(side = 1, 
     at = zval, 
     label = expression( italic(z) ) )
abline(h = 0, 
       lwd = 2)

abline(v = zval, 
       lwd = 2, 
       col = "grey")

xshade <- c(  seq(min(z), 
                  zval, 
                  by = delta),
              seq(zval, 
                  min(z), 
                  by = -delta) )
yshade <- c( rep(0, length(xshade)/2),
             rev(zy[ z<zval]) )
            
polygon( xshade, yshade, 
         col = plot.colour)

arrows(-2.5, 
       max(zy)/2,
       0, 
       max(zy)/2, 
       angle = 15,
       length = 0.1)
text(-2.5, max(zy)/2, 
     "Area",
     pos = 3)
```


\begin{small}
The table gives the probability (area) that a $z$-score is \textbf{less} than the $z$-score looked up.
For example: Look up $z = 1.87$; the area \textit{less than} $z = 1.87$ is about 0.9693, or about 96.9\%.
\end{small}

```{r AppendixTablesPositiveZLaTeX, echo=FALSE}
ZtabPos <- formatC( Ztab[ (length(z1n) + 1): length(z1), ],
                    format = "f",
                    digits = 4)

if( knitr::is_latex_output() ) {

   kable(ZtabPos,
        format = "latex",
        caption = "The probability that the area is less than the value of $z$ that is looked up",
        digits = 4,
        longtable = TRUE,
        escape = FALSE,
        align = "r",
        booktab = TRUE) %>%
      kable_styling(font_size = 9)
}
if( knitr::pandoc_to("epub3")) {
   kable(Ztab[ (length(z1n) + 1): length(z1), ],
        format = "html",
        caption = "The probability that the area is less than the value of $z$ that is looked up",
        digits = 4,
        longtable = TRUE,
        escape = FALSE,
        align = "r",
        booktab = TRUE) 
}
```




